### 2015-11-23 jss/article.tex
jss update

\documentclass[article]{jss}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% declarations for jss.cls %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% almost as usual
\author{Paul Deveau\\Institut Curie\\PSL Research University\\INSERM U900\\Mines-ParisTech \And 
        Eric Bonnet\\Institut Curie\\PSL Research University\\INSERM U900\\Mines-ParisTech}
\title{Calculating Module Enrichment and Visualizing Data on Large-scale Molecular Maps with the R packages \pkg{ACSNMineR} and \pkg{RNaviCell}}

%% for pretty printing and a nice hypersummary also set:
\Plainauthor{Paul Deveau, Eric Bonnet} %% comma-separated
\Plaintitle{Calculating Module Enrichment and Visualizing Data on Large-scale Molecular Maps with the R packages ACSNMineR and RNaviCel} %% without formatting
\Shorttitle{Module Enrichment and Data Visualization with ACSNMineR and RNaviCell} %% a short title (if necessary)

%% an abstract and keywords
\Abstract{
  The abstract of the article.
}
\Keywords{keywords, comma-separated, not capitalized, \proglang{Java}}
\Plainkeywords{keywords, comma-separated, not capitalized, Java} %% without formatting
%% at least one keyword must be supplied

%% publication information
%% NOTE: Typically, this can be left commented and will be filled out by the technical editor
%% \Volume{50}
%% \Issue{9}
%% \Month{June}
%% \Year{2012}
%% \Submitdate{2012-06-04}
%% \Acceptdate{2012-06-04}

%% The address of (at least) one author should be given
%% in the following format:
\Address{
  Eric Bonnet\\
  Computational Systems Biology of Cancer\\
  Institut Curie\\
  26, rue d'Ulm\\
  75248 Paris, France\\
  E-mail: \email{eric.bonnet@curie.fr}\\
}
%% It is also possible to add a telephone and fax number
%% before the e-mail in the following format:
%% Telephone: +43/512/507-7103
%% Fax: +43/512/507-2851

%% for those who use Sweave please include the following line (with % symbols):
%% need no \usepackage{Sweave.sty}

%% end of declarations %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

%% include your article here, just as usual
%% Note that you should use the \pkg{}, \proglang{} and \code{} commands.

\section[Introduction]{Introduction}
Biological pathways and networks comprise sets of interactions or functional
relationships, occurring at the molecular level in living cells

\bibliographystyle{jss}
\bibliography{biblio}

\end{document}

### 2015-11-23 jss/article.tex
jss update

@@
 relationships, occurring at the molecular level in living cells
+\citep{adriaens2008public}.  A large body of knowledge on cellular biochemistry
+is organized in publicly available repositories such as the KEGG database
+\citep{kanehisa2011kegg}, Reactome \citep{croft2014reactome} and MINT
+\citep{zanzoni2002mint}. All these biological databases facilitate a large
+spectrum of analyses, improving our understanding of cellular systems. For
+instance, it is a very common practice to cross the output of high-throughput
+experiments, such as mRNA or protein expression levels, with curated biological
+pathways in order to visualize changes, analyze their impact on a network and
+formulate new hypotheses about biological processes. Many biologists and
+computational biologists establish list of genes of interest (e.g. a list of
+genes that are differentially expressed between two conditions, such as normal
+vs disease) and then try to see if known biological pathways are enriched with
+this list of genes.
+
+We have recently released the Atlas of Cancer Signalling Network (ACSN), a
+web-based database which
+describes signaling and regulatory molecular processes that occur in
+a healthy mammalian cell but that are frequently deregulated during
+cancerogenesis. ACSN is
 

### 2015-11-23 R/ACSN_Enrichment.R
[HyperGeom]
Added comments in enrichment()
Added warning for hypergeom and low counts

@@ -38,6 +38,9 @@ enrichment<-function(Genes=NULL,
                      alternative = "greater"){
   
   ### Checking maps
+  if(statistical_test=="hypergeom"){
+    warning("p-values can be unreliable with hypergeometric test if the number of genes in module is low (typically 0)")
+  }
   if(is.data.frame(maps) | is.matrix(maps)){
     maps<-list(maps)
   }
@@ -190,6 +193,9 @@ enrichment<-function(Genes=NULL,
       modules<-map[,1:2]
       
       ### Calculate p-value for map as a whole if map is not ACSN_master
+      ###########################
+      ### SHOULD WE COMPUTE P-VAL FOR WHOLE MAP?
+      ###########################
       if(universe_was_ACSN & map_names[tracker]!="ACSN_master"){
         compute_module_p.value<-TRUE
       }
@@ -199,71 +205,95 @@ enrichment<-function(Genes=NULL,
       else{
         compute_module_p.value<-FALSE
       }
+      
+      ########################
+      # BEGIN MAP TESTING
+      ########################
       if( compute_module_p.value ){ ###
         mapgenes<-unique(as.character(unlist(map[,-c(1:2)])))
         
         mapsize<-length(mapgenes)
         genes_in_mapgenes<-(Genes %in% mapgenes)      
-        num<-sum( genes_in_mapgenes)
+        Ngenes_in_mapgenes<-sum(genes_in_mapgenes)
         
-        if(alternative != "both"){        
-          
-          if(statistical_test == "fisher"){
-            p.values<-c(p.val.calc(num,
-                                   mapsize-num,
+        ########################
+        # BEGIN FISHER TESTING
+        ########################
+        if(statistical_test == "fisher"){
+          if(alternative != "both"){        
+            p.values<-c(p.val.calc(Ngenes_in_mapgenes,
+                                   mapsize-Ngenes_in_mapgenes,
                                    length(Genes),
                                    size - length(Genes),
-                                   statistical_test,
+                                   "fisher",
                                    alternative
             ),
             alternative)
-
+            
           }
-          else{
-            p.values<-c(p.val.calc( num,
-                                    mapsize,
-                                    size - mapsize,
-                                    length(Genes),
-                                    statistical_test,
-                                    alternative
-            ),
-            alternative
-            ) 
-          }
-        }
-        else{ ### if both computations
-          if(statistical_test == "fisher"){
-            p.values<-cbind(c(p.val.calc(num,
[…]

### 2015-11-24 R/ACSN_Enrichment.R
[Updated hypergeom]
- coherent output for enrichment in "both" and greater cases
- warning for hypergeometric when computes depletion

@@ -37,10 +37,10 @@ enrichment<-function(Genes=NULL,
                      threshold = 0.05,
                      alternative = "greater"){
   
-  ### Checking maps
-  if(statistical_test=="hypergeom"){
-    warning("p-values can be unreliable with hypergeometric test if the number of genes in module is low (typically 0)")
+  if(statistical_test=="hypergeom" & (alternative != "greater")){
+    warning("Depletion with hypergeometric test can be unreliable. Prefer fisher test.")
   }
+  ### Checking maps
   if(is.data.frame(maps) | is.matrix(maps)){
     maps<-list(maps)
   }
@@ -168,11 +168,14 @@ enrichment<-function(Genes=NULL,
     stop("No genes in universe.")
     result<-data.frame()
   }
-  
+  ##################################
   #### Begin calculation of p-values
+  ##################################
+  
   
   ### get from list how many are in each sub-compartment
   result<-data.frame()
+  
   ### what would be the expected?
   tracker<-0
   map_names<-names(maps)
@@ -239,7 +242,7 @@ enrichment<-function(Genes=NULL,
                                          "fisher",
                                          "greater"
             ),
-            "greater"),
+            "enrichment"),
             c(p.val.calc(Ngenes_in_mapgenes,
                          mapsize-Ngenes_in_mapgenes,
                          Genes_size,
@@ -247,7 +250,7 @@ enrichment<-function(Genes=NULL,
                          "fisher",
                          "less"
             ),
-            "less")
+            "depletion")
             )
           }
         }
@@ -267,7 +270,10 @@ enrichment<-function(Genes=NULL,
               Ngcalc<-0
             }
           }
-          if(alternative != "both"){        
+          if(alternative == "greater" & Ngenes_in_mapgenes==0){ ### You cannot have enrichment if you have 0 genes in module
+            p.values<-c(1,alternative) 
+          }
+          else if(alternative != "both"){ # alternative is (less) or (greater & Ngenes>0)        
             
             p.values<-c(p.val.calc( Ngcalc,
                                     length(Genes),
@@ -279,27 +285,43 @@ enrichment<-function(Genes=NULL,
             alternative
             ) 
           }
-          
           else{ ### Testing both hypotheses
-            
-            p.values<-rbind(c(p.val.calc( Ngcalc,
-                                          length(Genes),
-                                          size - length(Genes),
-                                          mapsize,
-                                          "hypergeom",
-                                          'greater'
-            ),
-            'greater')
[…]

### 2015-11-24 jss/article.tex
jss update

@@
 We have recently released the Atlas of Cancer Signalling Network (ACSN), a
-web-based database which
-describes signaling and regulatory molecular processes that occur in
-a healthy mammalian cell but that are frequently deregulated during
-cancerogenesis. ACSN is
+web-based database which describes signaling and regulatory molecular processes
+that occur in a healthy mammalian cell but that are frequently deregulated
+during cancerogenesis \citep{kuperstein2013navicell}.  The ACSN atlas aims to
+be a comprehensive description of cancer-related mechanisms retrieved from the
+most recent literature. The web interface for ACSN is using the NaviCell
+technology, a software framework dedicated to web-based visualization and
+navigation for biological pathway maps \citep{kuperstein2013navicell}. This
+environment is providing an easy navigation of maps through the use of the
+Google Maps JavaScript library, a community interface with a web blog system,
+and a comprehensive module for visualization and analysis of high-throughput
+data \citep{bonnet2015navicell}.
 
-\bibliographystyle{jss}
+In this article, we describe two software packages related to ACSN analysis and
+data visualization for the popular R statistical enrironment \citep{mainRref,
+vance2009data}. The package \pkg{ACSNMineR} is designed for the calculation of
+gene enrichment and depletion in ACSN maps, while \pkg{RNaviCell} is dedicated
+to data visualization on ACSN maps. Both packages are available on the
+Comprehensive R Archive Network
+(\url{https://cran.r-project.org/web/packages/ACSNMineR/} and
+\url{https://cran.r-project.org/web/packages/RNaviCell/}), and on the GitHub
+repository (\url{https://github.com/sysbio-curie/ACSNMineR} and
+\url{https://github.com/sysbio-curie/RNaviCell}).
+
+%\bibliographystyle{jss}
 \bibliography{biblio}

### 2015-11-25 jss/article.tex
jss update

@@
 repository (\url{https://github.com/sysbio-curie/ACSNMineR} and
-\url{https://github.com/sysbio-curie/RNaviCell}).
+\url{https://github.com/sysbio-curie/RNaviCell}). For the remainder of this
+article, we describe the organization of each package and illustrate their
+capacities with several concrete examples demonstrating their capabilities.
+
+\section[Package organization]{Packages organization}
+
+\subsection{ACSNMineR}
+Currently, ACSN maps covers signaling pathways involved in DNA repair, cell cycle,
+cell survival, cell death, epithelial-to-mesenchymal transition (EMT) and cell
+motility. The information included in ACSN is organized in three hierarchical
+levels: a global map, which is composed of five individual maps, themselves
+composed of many functional modules covering individual pathways. Each map in
+ACSN covers hundreds of molecular players, biochemical reactions and causal
+relationships between the molecular players and cellular phenotypes. ACSN
+represents a large-scale biochemical reaction network of 4,826 reactions
+involving 2,371 proteins, and is continuously updated and expanded. 
+We have included the three hierarchical levels in the \pkg{ACSNMineR} package,
+in order to be able to calculate enrichments at all three levels.
+
+\subsection{RNaviCell}
 

### 2015-11-30 jss/article.tex
jss update
[skip ci] jss update

@@
 \subsection{ACSNMineR}
-Currently, ACSN maps covers signaling pathways involved in DNA repair, cell cycle,
-cell survival, cell death, epithelial-to-mesenchymal transition (EMT) and cell
-motility. The information included in ACSN is organized in three hierarchical
-levels: a global map, which is composed of five individual maps, themselves
-composed of many functional modules covering individual pathways. Each map in
-ACSN covers hundreds of molecular players, biochemical reactions and causal
+Currently, ACSN maps covers signaling pathways involved in DNA repair, cell
+cycle, cell survival, cell death, epithelial-to-mesenchymal transition (EMT) and
+cell motility. Each of these large-scale molecular map is decomposed in a number
+of functional modules. The maps themselves are merged into a global ACSN map.
+Thus the information included in ACSN is organized in three hierarchical levels:
+a global map, five individual maps, and a total of 55 functional modules. Each
+ACSN map covers hundreds of molecular players, biochemical reactions and causal
 relationships between the molecular players and cellular phenotypes.  ACSN
 represents a large-scale biochemical reaction network of 4,826 reactions
-involving 2,371 proteins, and is continuously updated and expanded. 
-We have included the three hierarchical levels in the \pkg{ACSNMineR} package,
-in order to be able to calculate enrichments at all three levels.
+involving 2,371 proteins, and is continuously updated and expanded.  We have
+included the three hierarchical levels in the \pkg{ACSNMineR} package, in order
+to be able to calculate enrichments at all three levels. The calculations are
+made by counting the number of occurences of gene symbols (HUGO gene names) from
+a given list of genes of interest in all ACSN maps and modules. Table
+\ref{tab:table1} is detailling the number of gene symbols contained in all the ACSN
+maps.
+
+\begin{table}[h!]
+ \centering
+  \caption{ACSN maps included in the \pkg{ACSNMineR} package. Map: map name, Total: total
+  number of gene symbols (HUGO), Nb mod.: number of modules, Min: mimimum
+  number of gene symbols in the modules, Max: maximum number of gene symbols in
+  the modules, Mean: average number of gene sybols per module. N.B.: one gene
+  symbol may be present in several modules of the map.}
+  \label{tab:table1}
+  \begin{tabular}{l|c|c|c|c|c}
+    \hline
+    Map & Total & Nb mod. & Min & Max & Mean\\
+    \hline
+  ACSN global & 2137 & 55 & 2 &625& 85\\
+  Survival  &1065&46  &1 &434 &54\\
+  Apoptosis & 668&44 & 1& 382& 33\\
+  EMT \& Cell motility &620 &41  &1 &615 &43\\
+  DNA repair &337&48  &1 &69  &19\\
+  Cell cycle &119&45  &1 &27  &8\\
+  \hline
+
+  \end{tabular}
+
+\end{table}
+
+The statistical significance of the counts in the modules is assessed by using
+either the Fisher exact test \citep{fisher1922interpretation,
+fisher1934statistical} or the hypergeometric test, which are equivalent for this
+purpose \cite{rivals2007enrichment}.
+
+The current ACSN maps are included in the \pkg{ACSNMineR} package, as a list of character matrices.
+
+\begin{verbatim}
+> length(ACSN_maps)
+[1] 6
+> names(ACSN_maps)
+[1] "Apoptosis"    "CellCycle"    "DNA_repair"   "EMT_motility" "ACSN_master" 
+[6] "Survival"    
+\end{verbatim}
+
+For each matrix, rows represent a module, with the name of the module in the
+first column, followed by a description of the module (optional), and then
+followed by all the gene symbols of the module. The maps will be updated
+according to every ACSN major release.
+
+The main function of the \pkg{ACSNMineR} package is the \code{enrichment}
+function, which is calculating over-representation or depletion of genes in the
+ACSN maps and modules.
 

### 2015-12-01 jss/article.tex
[skip ci] jss update

@@
 function, which is calculating over-representation or depletion of genes in the
-ACSN maps and modules.
+ACSN maps and modules. We have included a small list of 12 Cell Cycle related
+genes in the package, named \code{genes_test} that can be used to test the main
+enrichment function and to get familiar with its different options.
+
+\begin{verbatim}
+> genes_test
+ [1] "ATM"     "ATR"     "CHEK2"   "CREBBP"  "TFDP1"   "E2F1"    "EP300"  
+ [8] "HDAC1"   "KAT2B"   "GTF2H1"  "GTF2H2"  "GTF2H2B"
+\end{verbatim}
+
+\begin{verbatim}
+> library(ACSNMineR)
+> results <- enrichment(genes_test)
+> dim(results)
+[1] 36  9
+> results[1,]
+            module module_size nb_genes_in_module      genes_in_module
+V12 Apoptosis:E2F1           5                  4 ATM E2F1 EP300 KAT2B
+    universe_size nb_genes_in_universe      p.value p.value.corrected    test
+V12          2133                   12 1.043478e-08       2.81739e-07 greater
+\end{verbatim}
 

### 2015-12-03 jss/article.tex
[skip ci] jss update

@@
 
+The example shown below is the simplest command that can be done to test a gene
+list for over-representation on the six included ACSN maps. With the list of 12
+genes mentionned above and a default p-value cutoff of $0.05$, we have a set of
+36 maps or modules that are significantly enriched. The results are structured
+as a data frame with nine columns displaying the module name, the module size,
+the number of genes from the list in the module, the names of the genes that are
+present in the module, the size of the reference universe, the number of genes
+from the list that are present in the universe, the raw p-value, the p-value
+corrected for multiple testing and the type of test performed. The module field
+in the results data frame indicate the map name and the module name separated by
+a column character. If a complete map is significantly enriched or depleted,
+then only the map name is shown, without any module or column character. For
+instance, the first line of the results object below concern the E2F1 module of
+the Apoptosis map.
+
 \begin{verbatim}
@@
 
+%    enrichment(Genes = NULL, maps = ACSNMineR::ACSN_maps,
+%       correction_multitest = "BH", statistical_test = "fisher",
+%       min_module_size = 5, universe = "map_defined", threshold = 0.05,
+%       alternative = "greater")
+
+The \code{enrichment} function can take up to eight arguments: the gene list (as
+a character vector), the list of maps that will be used to calculate enrichment
+or depletion, the type of statistical test (either the Fisher exact test or the
+hypergeometric test), the module minimal size for which the calculations will be
+done, the universe, the p-value threshold and the alternative ("greater" for
+calculating over-representation, "less" for depletion and "both" for both
+tests).
+
+Only the gene list is mandatory to call the \code{enrichment} function, all the
+other arguments have default values.
+
 \subsection{RNaviCell}

### 2015-12-04 jss/article.tex
[skip ci] jss update
Added enrichment default values

@@
 
+[explain default values]\\
+The \code{maps} argument can either be a dataframe imported from a gmt file with the \code{format_from_gmt} function or a list of dataframes generated by the same procedure.
+By default, the function uses the ACSN maps previously described.
+
+The correction for multiple testing by default is set to false discovery rate (fdr), which is equivalent to Benjamini \& Hochberg correction, but can be changed to any of the usual corrections such as Bonferroni, Holm, Hochberg, Holm, or Benjamini \& Yekutieli \cite{Benjamini2003FDR}, or even disabled .
+
+The minimal module size represents the smallest size value of a module that will be used to compute enrichment or depletion.
+This is meant to remove results of low significance for module of small size.
+
+The universe in which the computation is made by default is defined by the maps.
+Which means that all genes that were given as input and that are not present on the maps will be discarded.
+To keep all genes, the user can set the universe to \code{HUGO},referring to the database.
+
+The threshold is the maximal value of the corrected p-value (unless user chose not to correct for multiple testing) that will be displayed in the result table.
+
+[give examples with maps, universe, correction multtest]
+
+[explain multiple samples]
+
+[explain graphs generation]
+
 \subsection{RNaviCell}

### 2015-12-04 jss/article.tex
[skip ci] jss update

--- 2015-12-03 jss/article.tex
@@
 
+[explain default values]
+
+[give examples with maps, universe, correction multtest]
+
+[explain multiple samples]
+
+[explain graphs generation]
+
 \subsection{RNaviCell}
+
+The NaviCell Web Service provides a server mode, which allows automating
+visualization tasks and retrieving data from molecular maps via RESTful
+(standard http/https) calls. Bindings to different programming languages are
+provided in order to facilitate the development of data visualization workflows and
+third-party applications \citep{bonnet2015navicell}. RNaviCell is the R binding
+to the NaviCell Web Service. It is implemented as a standard R package, using
+the R object-oriented framework known as Reference Classes \citep{hwR5}. Most
+of the work done by the user using graphical point-and-click operations on the
+NaviCell web interface are encoded as functions in the library encapsulating
+http calls to the server with appropriate parameters and data. Calls to the
+NaviCell server are performed using the library RCurl \citep{rcurl2015}.
+
+\section[Examples]{Examples}
 

### 2015-12-04 jss/article.tex
[skip ci] jss update

--- 2015-12-04 jss/article.tex
@@
 
+The NaviCell Web Service provides a server mode, which allows automating
+visualization tasks and retrieving data from molecular maps via RESTful
+(standard http/https) calls. Bindings to different programming languages are
+provided in order to facilitate the development of data visualization workflows and
+third-party applications \citep{bonnet2015navicell}. RNaviCell is the R binding
+to the NaviCell Web Service. It is implemented as a standard R package, using
+the R object-oriented framework known as Reference Classes \citep{hwR5}. Most
+of the work done by the user using graphical point-and-click operations on the
+NaviCell web interface are encoded as functions in the library encapsulating
+http calls to the server with appropriate parameters and data. Calls to the
+NaviCell server are performed using the library RCurl \citep{rcurl2015}, while
+data encoding/decoding in JSON format is performed with the RJSONIO library
+\citep{rjsonio2014}.
+
+\section[Examples]{Examples}
+
 %\bibliographystyle{jss}

### 2015-12-07 jss/article.tex
Added multisample and representation
[skip ci]

@@
 
-[give examples with maps, universe, correction multtest]
+[give examples with maps, universe, correction multitest]
 
-[explain multiple samples]
+%multisample_enrichment<-function(Genes_by_sample=NULL,
+%                                 maps = ACSNMineR::ACSN_maps, 
+%                                 correction_multitest = "BH",
+%                                 statistical_test = "fisher",
+%                                 min_module_size = 5,
+%                                 universe = "map_defined",
+%                                 threshold = 0.05,
+%                                 cohort_threshold = TRUE,
+%                                 alternative = "greater"){
+It may be of interest to compare enrichment of pathwways in different cohorts or experiments.
+For example, enrichment of highly expressed pathways can reveal differences between two cancer types or two cell lines.
+To ease such comparisons, \pkg{ACSNMineR} provides a \code{multisample_enrichment} function.
+It relies on the \code{enrichment} function but takes a list of character vector genes.
+The name of each element of the list will be assumed to be the name of the sample for further analysis.
+Most of the arguments given to \code{multisample_enrichment} are the same as the ones passed to \code{enrichment}.
+However, the \code{cohort_threshold} is designed to filter out modules which would not pass the significance threshold in all samples.
 
-[explain graphs generation]
+%represent_enrichment<-function(enrichment, plot = "heatmap" , scale = "log", 
+%                               low = "steelblue" , high ="white",
+%                              nrow = 1,sample_name = "Sample",
+%                               na.value = "grey"){
+Finally, to facilitate visualization of results, \pkg{ACSNMineR} integrates a representation function based on \pkg{ggplot2} syntax \cite{ggplot2}.
+It allows representation of results from \code{enrichment} or \code{multisample_enrichment} with a limited number of parameters.
+For examples, two types of display are available: heat-map tiles or bars.
+For multiple samples using a barplot representation the number of rows used can be provided, all plots will be on the same row otherwise.
+For the heatmap, the color of the non-significant modules, and boundaries of the gradient for significant values can also be tuned.
 

### 2015-12-07 jss/article.tex
Changed cite to citep

@@
 
-The correction for multiple testing by default is set to false discovery rate (fdr), which is equivalent to Benjamini \& Hochberg correction, but can be changed to any of the usual corrections such as Bonferroni, Holm, Hochberg, Holm, or Benjamini \& Yekutieli \cite{Benjamini2003FDR}, or even disabled .
+The correction for multiple testing by default is set to false discovery rate (fdr), which is equivalent to Benjamini \& Hochberg correction, but can be changed to any of the usual corrections such as Bonferroni, Holm, Hochberg, Holm, or Benjamini \& Yekutieli \citep{Benjamini2003FDR}, or even disabled .
 
@@
 %                               na.value = "grey"){
-Finally, to facilitate visualization of results, \pkg{ACSNMineR} integrates a representation function based on \pkg{ggplot2} syntax \cite{ggplot2}.
+Finally, to facilitate visualization of results, \pkg{ACSNMineR} integrates a representation function based on \pkg{ggplot2} syntax \citep{ggplot2}.
 It allows representation of results from \code{enrichment} or \code{multisample_enrichment} with a limited number of parameters.

### 2015-12-07 jss/article.tex
[skip ci] jss update

--- 2015-12-04 jss/article.tex
@@
 fisher1934statistical} or the hypergeometric test, which are equivalent for this
-purpose \cite{rivals2007enrichment}.
+purpose \citep{rivals2007enrichment}.
 
@@
 
-The correction for multiple testing by default is set to false discovery rate (fdr), which is equivalent to Benjamini \& Hochberg correction, but can be changed to any of the usual corrections such as Bonferroni, Holm, Hochberg, Holm, or Benjamini \& Yekutieli \cite{Benjamini2003FDR}, or even disabled .
+The correction for multiple testing by default is set to false discovery rate
+(fdr), which is equivalent to Benjamini \& Hochberg correction, but can be
+changed to any of the usual corrections such as Bonferroni, Holm, Hochberg,
+Holm, or Benjamini \& Yekutieli \citep{Benjamini2003FDR}, or even disabled .
 
@@
 
+Once the RNaviCell library is installed and loaded, the first step is to create
+a NaviCell object and launch the browser session. This will automatically create
+a unique session ID with the NaviCell server. Once the session is established,
+various functions can be called to send data to the web session, set graphical
+options, visualize data on a map or get data from the map. There are 125
+functions available in the current version of RNaviCell. All of them are
+described with their different options in the RNaviCell documentation, and we
+provide a tutorial on the GitHub repository wiki
+(\url{https://github.com/sysbio-curie/RNaviCell/wiki/Tutorial}).
+
 \section[Examples]{Examples}

### 2015-12-08 jss/article.tex
Added example of uncorrected test

@@
 %                                 alternative = "greater"){
-It may be of interest to compare enrichment of pathwways in different cohorts or experiments.
+It may be of interest to compare enrichment of pathways in different cohorts or experiments.
 For example, enrichment of highly expressed pathways can reveal differences between two cancer types or two cell lines.
@@
 For the heatmap, the color of the non-significant modules, and boundaries of the gradient for significant values can also be tuned.
+
+The package also includes a use-case that we detail below.
+Using, \code{genes_test} it is possible to explore the different possibilities of exploration of the maps.
+We previously computed the p-value of \code{genes_test} with default parameters.
+The number of modules which have a p-value below 0.05 was 36, that can be compared to the 49 obtained without correction:
+\code{enrichment(genes_test,correction_multitest = FALSE)}
+\begin{table}[h!]
+  \centering
+  \caption{Overview of results from enrichment analysis without correction.
+Module : name of the module.
+Module size: size of the module.
+Genes in module: genes from input which are found in the module.
+p-value: uncorrected p-value.
+Test : null hypothesis used, greater is synonym of enrichment.}
+  \label{tab:table2}
+
+  \begin{tabular}{l|c|c|c|c}
+	\hline
+Module & Module size & Genes in module & p-value & Test\\
+	\hline
+Apoptosis:E2F1 & 5 &  ATM E2F1 EP300 KAT2B  & 1.043478e-08 & greater \\
+Apoptosis:G2 M CHECKPOINT & 35  & ATM KAT2B & 2.040546e-02 & greater \\
+Apoptosis:WNT CANONICAL & 53 & E2F1 EP300 KAT2B  & 4.984433e-03 & greater \\
+	\end{tabular}
+\end{table}
 

### 2015-12-08 jss/article.tex
Updated affiliation Paul

@@
 %% almost as usual
-\author{Paul Deveau\\Institut Curie\\PSL Research University\\INSERM U900\\Mines-ParisTech \And 
+\author{Paul Deveau\\Institut Curie\\PSL Research University\\Univ. Paris-Sud\\INSERM U830\\INSERM U900\\Mines-ParisTech \And 
         Eric Bonnet\\Institut Curie\\PSL Research University\\INSERM U900\\Mines-ParisTech}

### 2015-12-08 jss/article.tex
[skip ci] jss update

@@
 Module : name of the module.
-Module size: size of the module.
+Mod.
+size: size of the module.
 Genes in module: genes from input which are found in the module.
@@
 	\hline
-Module & Module size & Genes in module & p-value & Test\\
+Module & Mod. size & Genes in module & p-value & Test\\
 	\hline
-Apoptosis:E2F1 & 5 &  ATM E2F1 EP300 KAT2B  & 1.043478e-08 & greater \\
-Apoptosis:G2 M CHECKPOINT & 35  & ATM KAT2B & 2.040546e-02 & greater \\
-Apoptosis:WNT CANONICAL & 53 & E2F1 EP300 KAT2B  & 4.984433e-03 & greater \\
+Apoptosis:E2F1 & 5 &  ATM E2F1 EP300 & 1.04e-08 & greater \\
+               &   &  KAT2B                 &          &         \\
+Apoptosis:G2 M CHECKPOINT & 35  & ATM KAT2B & 2.04e-02 & greater \\
+Apoptosis:WNT CANONICAL & 53 & E2F1 EP300 KAT2B  & 4.98e-03 & greater \\
 	\end{tabular}

### 2015-12-08 jss/article.tex
[skip ci] jss update

@@
 For example, enrichment of highly expressed pathways can reveal differences between two cancer types or two cell lines.
-To ease such comparisons, \pkg{ACSNMineR} provides a \code{multisample_enrichment} function.
+To facilitate such comparisons, \pkg{ACSNMineR} provides a \code{multisample_enrichment} function.
 It relies on the \code{enrichment} function but takes a list of character vector genes.
@@
 %                               na.value = "grey"){
-Finally, to facilitate visualization of results, \pkg{ACSNMineR} integrates a representation function based on \pkg{ggplot2} syntax \citep{ggplot2}.
-It allows representation of results from \code{enrichment} or \code{multisample_enrichment} with a limited number of parameters.
-For examples, two types of display are available: heat-map tiles or bars.
-For multiple samples using a barplot representation the number of rows used can be provided, all plots will be on the same row otherwise.
-For the heatmap, the color of the non-significant modules, and boundaries of the gradient for significant values can also be tuned.
 
-The package also includes a use-case that we detail below.
-Using, \code{genes_test} it is possible to explore the different possibilities of exploration of the maps.
-We previously computed the p-value of \code{genes_test} with default parameters.
-The number of modules which have a p-value below 0.05 was 36, that can be compared to the 49 obtained without correction:
-\code{enrichment(genes_test,correction_multitest = FALSE)}
+Finally, to facilitate visualization of results, \pkg{ACSNMineR} integrates a
+representation function based on \pkg{ggplot2} syntax \citep{ggplot2}. It
+allows representation of results from \code{enrichment} or
+\code{multisample_enrichment} with a limited number of parameters. For
+examples, two types of display are available: heat-map tiles or bars. For
+multiple samples using a barplot representation the number of rows used can be
+provided, all plots will be on the same row otherwise. For the heatmap, the
+color of the non-significant modules, and boundaries of the gradient for
+significant values can also be tuned.
+
+The package also includes a use-case that we detail below. Using,
+\code{genes_test} it is possible to explore the different possibilities of
+exploration of the maps. We previously computed the p-value of
+\code{genes_test} with default parameters.  The number of modules which have a
+p-value below 0.05 was 36, that can be compared to the 49 obtained without
+correction: \code{enrichment(genes_test,correction_multitest = FALSE)}
+
 \begin{table}[h!]
@@
   \caption{Overview of results from enrichment analysis without correction.
-Module : name of the module.
-Mod.
-size: size of the module.
-Genes in module: genes from input which are found in the module.
-p-value: uncorrected p-value.
+  Module : name of the module. Mod. size: size of the module. Genes in module:
+  genes from input which are found in the module. p-value: uncorrected p-value.
   Test : null hypothesis used, greater is synonym of enrichment.}
@@
 Apoptosis:WNT CANONICAL & 53 & E2F1 EP300 KAT2B  & 4.98e-03 & greater \\
+	\hline
 	\end{tabular}
@@
 
+\begin{verbatim}
+> library(RNaviCell)
+> navicell <- NaviCell()
+> navicell$launchBrowser()
+
+\end{verbatim}
+
 \section[Examples]{Examples}

### 2015-12-15 jss/article.tex
[skip ci] jss update

@@
 
+In the simple example detailed below, we create a NaviCell session, then load
+an expression data set from a local (tab-delimited) file. The data represent
+gene expression measured in a prostate cancer cell line resistant to hormonal
+treatment (agressive), and is taken from the Cell Line Encyclopedia project
+\citep{barretina2012cancer}. We visualize the data values on the Cell Cycle map
+(the default map), using heat maps. With this visualization mode, gene
+expression values are represented as a color gradient (green to red) in
+squares positioned next to the entities where the gene has been mapped.
+
 \begin{verbatim}
-> library(RNaviCell)
-> navicell <- NaviCell()
-> navicell$launchBrowser()
+# load RNaviCell library
+
+library(RNaviCell)
+
+# create a NaviCell object and launch a server session
+# this will automatically open a browser on the client
+
+navicell <- NaviCell()
+navicell$launchBrowser()
+
+data_matrix <- navicell$readDatatable('DU145_data.txt')
+navicell$importDatatable("mRNA expression data", "DU145", data_matrix)
+
+navicell$heatmapEditorSelectSample('0','data')
+navicell$heatmapEditorSelectDatatable('0','DU145')
+navicell$heatmapEditorApply()
 
@@
 
+\begin{figure}[!ht]
+  \caption{Gene expression values from a prostate cancer cell line visualized
+on the cell cycle map as heat map plots.}
+  \centering
+    \includegraphics[width=0.8\textwidth]{figures/heatmap.pdf}
+\end{figure}
+
+%navicell$launchBrowser()
 \section[Examples]{Examples}

### 2015-12-16 jss/article.tex
[skip ci] jss update

@@
 Only the gene list is mandatory to call the \code{enrichment} function, all the
-other arguments have default values.
-
-[explain default values]\\
-The \code{maps} argument can either be a dataframe imported from a gmt file with the \code{format_from_gmt} function or a list of dataframes generated by the same procedure.
-By default, the function uses the ACSN maps previously described.
-
-The correction for multiple testing by default is set to false discovery rate (fdr), which is equivalent to Benjamini \& Hochberg correction, but can be changed to any of the usual corrections such as Bonferroni, Holm, Hochberg, Holm, or Benjamini \& Yekutieli \citep{Benjamini2003FDR}, or even disabled .
-
-The minimal module size represents the smallest size value of a module that will be used to compute enrichment or depletion.
-This is meant to remove results of low significance for module of small size.
-
-The universe in which the computation is made by default is defined by the maps.
-Which means that all genes that were given as input and that are not present on the maps will be discarded.
-To keep all genes, the user can set the universe to \code{HUGO},referring to the database.
-
-The threshold is the maximal value of the corrected p-value (unless user chose not to correct for multiple testing) that will be displayed in the result table.
-
-[give examples with maps, universe, correction multitest]
+other arguments have default values.  The \code{maps} argument can either be a
+dataframe imported from a gmt file with the \code{format_from_gmt} function or
+a list of dataframes generated by the same procedure. By default, the function
+uses the ACSN maps previously described.  The correction for multiple testing
+is set by default to false discovery rate (FDR), which is equivalent to
+Benjamini \& Hochberg correction, but can be changed to any of the usual
+correction methods (Bonferroni, Holm, Hochberg, Holm, or Benjamini \& Yekutieli
+\citep{Benjamini2003FDR}), or even disabled .  The minimal module size
+represents the smallest size value of a module that will be used to compute
+enrichment or depletion. This is meant to remove results of low significance
+for module of small size.  The universe in which the computation is made by
+default is defined by all the gene symbols conatined in the maps. All the genes
+that were given as input and that are not present on the maps will be
+discarded. To keep all genes, the user can change the universe to \code{HUGO},
+and in that case, the complete list of HUGO gene symbols will be used as the
+reference ($>$ 39,000 genes). The threshold corresponds to the maximal value of the
+corrected p-value (unless the user chose not to correct for multiple testing) that
+will be displayed in the result table.
 
@@
 %                                 alternative = "greater"){
-It may be of interest to compare enrichment of pathways in different cohorts or experiments.
-For example, enrichment of highly expressed pathways can reveal differences between two cancer types or two cell lines.
-To facilitate such comparisons, \pkg{ACSNMineR} provides a \code{multisample_enrichment} function.
-It relies on the \code{enrichment} function but takes a list of character vector genes.
-The name of each element of the list will be assumed to be the name of the sample for further analysis.
-Most of the arguments given to \code{multisample_enrichment} are the same as the ones passed to \code{enrichment}.
-However, the \code{cohort_threshold} is designed to filter out modules which would not pass the significance threshold in all samples.
+
+It may be of interest to compare enrichment of pathways in different cohorts or
+experiments. For example, enrichment of highly expressed pathways can reveal
+differences between two cancer types or two cell lines.  To facilitate such
+comparisons, \pkg{ACSNMineR} provides a \code{multisample_enrichment} function.
+It relies on the \code{enrichment} function but takes a list of character
+vector genes. The name of each element of the list will be assumed to be the
+name of the sample for further analysis.  Most of the arguments given to
+\code{multisample_enrichment} are the same as the ones passed to
+\code{enrichment}. However, the \code{cohort_threshold} is designed to filter
+out modules which would not pass the significance threshold in all samples.
 
@@
 expression values are represented as a color gradient (green to red) in
-squares positioned next to the entities where the gene has been mapped.
+squares positioned next to the entities where the gene has been mapped (Figure~\ref{fig:du145}).
+
+\begin{figure}[!ht]
+  \caption{Gene expression values from a prostate cancer cell line visualized
+on the cell cycle map as heat map plots. The figure is a screenshot of the
+NaviCell map browser, with the map set at the top (the less detailed) zoom
+level. The essential phases of the cell cycle are indicated on the map
+(G1/S/G2/M).} 
+  \centering
+  \includegraphics[width=0.8\textwidth]{figures/heatmap.pdf}
+  \label{fig:du145}
+\end{figure}
 
 \begin{verbatim}
+# a short RNaviCell script example
+
 # load RNaviCell library
@@
 
+# import a gene expression matrix and 
+# send the data to the NaviCell server
+# NB: the data_matrix object is a regular R matrix
+
 data_matrix <- navicell$readDatatable('DU145_data.txt')
 navicell$importDatatable("mRNA expression data", "DU145", data_matrix)
+
+# set data set and sample for heat map representation
 
@@
 \end{verbatim}
-
-\begin{figure}[!ht]
-  \caption{Gene expression values from a prostate cancer cell line visualized
-on the cell cycle map as heat map plots.}
-  \centering
-    \includegraphics[width=0.8\textwidth]{figures/heatmap.pdf}
-\end{figure}
 

### 2015-12-17 jss/article.tex
Added examples
[skip ci]

--- 2015-12-08 jss/article.tex
@@
 For example, enrichment of highly expressed pathways can reveal differences between two cancer types or two cell lines.
-To facilitate such comparisons, \pkg{ACSNMineR} provides a \code{multisample_enrichment} function.
+To ease such comparisons, \pkg{ACSNMineR} provides a \code{multisample_enrichment} function.
 It relies on the \code{enrichment} function but takes a list of character vector genes.
@@
 %                               na.value = "grey"){
+Finally, to facilitate visualization of results, \pkg{ACSNMineR} integrates a representation function based on \pkg{ggplot2} syntax \citep{ggplot2}.
+It allows representation of results from \code{enrichment} or \code{multisample_enrichment} with a limited number of parameters.
+For examples, two types of display are available: heat-map tiles or bars.
+For multiple samples using a barplot representation the number of rows used can be provided, all plots will be on the same row otherwise.
+For the heatmap, the color of the non-significant modules, and boundaries of the gradient for significant values can also be tuned.
 
-Finally, to facilitate visualization of results, \pkg{ACSNMineR} integrates a
-representation function based on \pkg{ggplot2} syntax \citep{ggplot2}. It
-allows representation of results from \code{enrichment} or
-\code{multisample_enrichment} with a limited number of parameters. For
-examples, two types of display are available: heat-map tiles or bars. For
-multiple samples using a barplot representation the number of rows used can be
-provided, all plots will be on the same row otherwise. For the heatmap, the
-color of the non-significant modules, and boundaries of the gradient for
-significant values can also be tuned.
-
-The package also includes a use-case that we detail below. Using,
-\code{genes_test} it is possible to explore the different possibilities of
-exploration of the maps. We previously computed the p-value of
-\code{genes_test} with default parameters.  The number of modules which have a
-p-value below 0.05 was 36, that can be compared to the 49 obtained without
-correction: \code{enrichment(genes_test,correction_multitest = FALSE)}
-
+The package also includes a use-case that we detail below.
+Using, \code{genes_test} it is possible to explore the different possibilities of exploration of the maps.
+We previously computed the p-value of \code{genes_test} with default parameters.
+The number of modules which have a p-value below 0.05 was 36, that can be compared to the 49 obtained without correction:
+\code{enrichment(genes_test,correction_multitest = FALSE)}
 \begin{table}[h!]
@@
   \caption{Overview of results from enrichment analysis without correction.
-  Module : name of the module. Mod. size: size of the module. Genes in module:
-  genes from input which are found in the module. p-value: uncorrected p-value.
+Module : name of the module.
+Module size: size of the module.
+Genes in module: genes from input which are found in the module.
+p-value: uncorrected p-value.
 Test : null hypothesis used, greater is synonym of enrichment.}
@@
 	\hline
-Module & Mod. size & Genes in module & p-value & Test\\
+Module & Module size & Genes in module & p-value & Test\\
 	\hline
-Apoptosis:E2F1 & 5 &  ATM E2F1 EP300 & 1.04e-08 & greater \\
-               &   &  KAT2B                 &          &         \\
-Apoptosis:G2 M CHECKPOINT & 35  & ATM KAT2B & 2.04e-02 & greater \\
-Apoptosis:WNT CANONICAL & 53 & E2F1 EP300 KAT2B  & 4.98e-03 & greater \\
+Apoptosis:E2F1 & 5 &  ATM E2F1 EP300 KAT2B  & 1.043478e-08 & greater \\
+Apoptosis:G2 M CHECKPOINT & 35  & ATM KAT2B & 2.040546e-02 & greater \\
+Apoptosis:WNT CANONICAL & 53 & E2F1 EP300 KAT2B  & 4.984433e-03 & greater \\
 	\hline
@@
 \end{table}
+
+We can now plot the first ten rows of the results obtained for corrected and uncorrected fisher test (figure 1).
+
+\begin{figure}
+	\centering
+	\caption{Representation of the enriched modules (first ten rows), either with either Bonferroni correction or no correction.
+Grey tiles means that the data is not available for this module in this sample.
+P-values of low signifance are in white, whereas p-values of high significance are represent in blue.
+}
+	\includegraphics{comparison_corrected_unc.pdf}
+\end{figure}
 
@@
 
-\begin{verbatim}
-> library(RNaviCell)
-> navicell <- NaviCell()
-> navicell$launchBrowser()
-
-\end{verbatim}
-
 \section[Examples]{Examples}

### 2015-12-17 jss/article.tex
Moved figure
[skip ci]

--- 2015-12-16 jss/article.tex
@@
 	\hline
+
 	\end{tabular}
 \end{table}
+
+We can now plot the first ten rows of the results obtained for corrected and uncorrected fisher test (figure 1).
+
+\begin{figure}
+	\centering
+	\caption{Representation of the enriched modules (first ten rows), either with either Bonferroni correction or no correction.
+Grey tiles means that the data is not available for this module in this sample.
+P-values of low signifance are in white, whereas p-values of high significance are represent in blue.
+}
+	\includegraphics{figures/comparison_corrected_unc.pdf}
+\end{figure}
 

### 2015-12-17 jss/article.tex
Added barplot
[skip ci]

@@
 Grey tiles means that the data is not available for this module in this sample.
-P-values of low signifance are in white, whereas p-values of high significance are represent in blue.
+P-values of low significance are in white, whereas p-values of high significance are represent in blue.
+Samples are in abscissa and modules are on the Y axis  }
+	\includegraphics{figures/comparison_corrected_unc.pdf}
+\end{figure}
+
+\begin{figure}
+	\caption{Representation of the enriched modules (first ten rows), either with either Bonferroni correction (left) or no correction (right).
+The modules are on the X axis and the p-values are on the Y axis.
 }
-	\includegraphics{figures/comparison_corrected_unc.pdf}
+	\includegraphics{figures/comparison_corrected_unc_bars.pdf}
 \end{figure}

### 2015-12-17 jss/article.tex
Added ref of figures
[skip ci]
[skip ci] jss update

@@
 
-The package also includes a use-case that we detail below. Using,
-\code{genes_test} it is possible to explore the different possibilities of
-exploration of the maps. We previously computed the p-value of
-\code{genes_test} with default parameters.  The number of modules which have a
-p-value below 0.05 was 36, that can be compared to the 49 obtained without
-correction: \code{enrichment(genes_test,correction_multitest = FALSE)}
+We previously computed the p-value of the \code{genes_test} list with default
+parameters. The number of modules which have a p-value below 0.05 was 36, that
+can be compared to the 49 obtained without correction with the simple command
+shown below (some of the results are displayed in table \ref{tab:table2}).
+
+\begin{verbatim}
+enrichment(genes_test,correction_multitest = FALSE)
+\end{verbatim}
 
@@
 
-We can now plot the first ten rows of the results obtained for corrected and uncorrected fisher test (figure 1).
+We can now plot the first ten rows of the results obtained for corrected and uncorrected fisher test with heatmap format (Figure~\ref{fig:heatm}) or barplot (Figure~\ref{fig:barp}).
 
@@
 	\includegraphics{figures/comparison_corrected_unc.pdf}
+    \label{fig:heatm}
+
 \end{figure}
@@
 	\includegraphics{figures/comparison_corrected_unc_bars.pdf}
+	\label{fig:barp}
+
 \end{figure}

### 2015-12-17 jss/article.tex
[skip ci] jss update

@@
 
-We can now plot the first ten rows of the results obtained for corrected and uncorrected fisher test with heatmap format (Figure~\ref{fig:heatm}) or barplot (Figure~\ref{fig:barp}).
+We can now plot the first ten rows of the results obtained for corrected and
+uncorrected fisher test with heatmap format (Figure~\ref{fig:heatm}) or barplot
+(Figure~\ref{fig:barp}) with the following commands:
+
+\begin{verbatim}
+# heatmap
+
+represent_enrichment(enrichment = list(SampleA = enrichment_test[1:10,], 
+  SampleB = enrichment_test[3:10,]), plot = "heatmap", scale = "log", 
+  low = "steelblue" , high ="white", na.value = "grey")
+
+# barplot
+
+represent_enrichment(enrichment = list(SampleA = enrichment_test[1:10,],
+ SampleB = enrichment_test[3:10,]), plot = "bar", scale = "log", nrow = 1)
+\end{verbatim}
 

### 2016-01-07 jss/article.tex
[skip ci] started to write breast cancer example.

@@
 
+\subsection{Analysis of breast cancer expression data}
+In a study published in 2008, Schmidt and colleagues analyzed gene expression
+patterns of 200 breast cancer patients not treated by systemic therapy after
+surgery using discovery approach to reveal additional prognostic motifs
+\cite{schmidt2008humoral}. Estrogen receptor (ER) expression and proliferative
+activity of breast carcinomas are well-known and described prognostic markers.
+Patients with ER-positive carcinomas have a better prognosis than those with
+ER-negative carcinomas, and rapidly proliferating carcinomas have an adverse
+prognosis. Knowledge about the molecular mechanisms involved in the
+processes of estrogen-dependent tumor growth and proliferative activity has led
+to the successful development of therapeutic concepts, such as  antiendocrine and
+cytotoxic chemotherapy.
+
+The dataset corresponding to this study is available as a
+Bioconductor package. The code shown below is creating a list of differentially
+expressed genes
+
+\begin{verbatim}
+
+library(breastCancerMAINZ)
+library(Biobase)
+library(limma)
+library(ACSNMineR)
+library(hgu133a.db)
+library(RNaviCell)
+
+data(mainz)
+eset <- exprs(mainz)
+pdat <- pData(mainz)
+design <- model.matrix(~factor(pdat$er == '1'))
+lmFit(eset, design) -> fit
+eBayes(fit) -> ebayes
+toptable(ebayes, coef=2,n=25000)->tt
+#which(abs(tt$logFC) >=1 & tt$adj < 0.05) -> sel
+which(tt$adj < 0.05) -> sel
+rownames(tt[sel,]) -> probe_list
+mget(probe_list, env=hgu133aSYMBOL) -> slist
+slist <- as.character(slist)
+
+enrichment(slist) -> res
+
+#apply(eset[probe_list,pdat$er == 1],1,mean)->er_plus_mean
+#apply(eset[probe_list,pdat$er == 0],1,mean)->er_minus_mean
+#cbind(er_plus_mean, er_minus_mean)-> data_er
+#rownames(data_er) <- slist
+
+#n$map_url <- "http://acsn.curie.fr/navicell/maps/apoptosis/master/index.php"
+#n$importDatatable("mRNA expression data", "exp", data_er)
+
+apply(eset[probe_list,pdat$er == 0],1,mean) -> er_minus_mean
+names(er_minus_mean) <- slist
+er_minus_mean <- as.matrix(er_minus_mean)
+colnames(er_minus_mean) <- c('exp')
+
+n <- NaviCell()
+n$proxy_url <- "https://acsn.curie.fr/cgi-bin/nv_proxy.php"
+n$map_url <- "https://acsn.curie.fr/navicell/maps/acsn/master/index.php"
+
+n$launchBrowser()
+n$importDatatable("mRNA expression data", "GBM_exp", er_minus_mean)
+n$heatmapEditorSelectSample('0','exp')
+n$heatmapEditorSelectDatatable('0','GBM_exp')
+n$heatmapEditorApply()
+
+# analysis mtsig signature file
+
+#mt <- format_from_gmt('/Users/eric/Downloads/c2.cp.v5.0.symbols.gmt')
+#unique(slist)->l
+#> length(l)
+#[1] 5179
+#> head(l)
+#[1] "ESR1"   "CA12"   "GATA3"  "TBC1D9" "FOXA1"  "IL6ST" 
+#> enrichment(l, maps=mt)
+
+\end{verbatim}
+
 %\bibliographystyle{jss}

### 2016-01-08 jss/article.tex
[skip ci] breast cancer example update

--- 2015-12-17 jss/article.tex
@@
 
+\subsection{Analysis of breast cancer expression data}
+In a study published in 2008, Schmidt and colleagues analyzed gene expression
+patterns of 200 breast cancer patients not treated by systemic therapy after
+surgery using discovery approach to reveal additional prognostic motifs
+\citep{schmidt2008humoral}. Estrogen receptor (ER) expression and proliferative
+activity of breast carcinomas are well-known and described prognostic markers.
+Patients with ER-positive carcinomas have a better prognosis than those with
+ER-negative carcinomas, and rapidly proliferating carcinomas have an adverse
+prognosis. Knowledge about the molecular mechanisms involved in the
+processes of estrogen-dependent tumor growth and proliferative activity has led
+to the successful development of therapeutic concepts, such as  antiendocrine and
+cytotoxic chemotherapy.
+
+The dataset corresponding to this study is available as a Bioconductor package.
+The code shown below is creating a list of differentially expressed genes
+
+\begin{verbatim}
+# load all necessary packages
+library(breastCancerMAINZ)
+library(Biobase)
+library(limma)
+library(ACSNMineR)
+library(hgu133a.db)
+library(RNaviCell)
+
+# load data and extract expression and phenotype data 
+data(mainz)
+eset <- exprs(mainz)
+pdat <- pData(mainz)
+
+# Create list of genes differentially expressed between ER positive and 
+# ER negative samples using moderated t-test statistics 
+design <- model.matrix(~factor(pdat$er == '1'))
+lmFit(eset, design) -> fit
+eBayes(fit) -> ebayes
+toptable(ebayes, coef=2,n=25000) -> tt
+which(tt$adj < 0.05) -> selection
+rownames(tt[selection,]) -> probe_list
+mget(probe_list, env=hgu133aSYMBOL) -> symbol_list
+symbol_list <- as.character(symbol_list)
+
+# calculate enrichment in ACSN maps
+
+enrichment(symbol_list) -> results
+
+\end{verbatim}
+
+%> as.character(res$module)
+%[1] "Apoptosis:HEDGEHOG"          "Apoptosis:AKT_MTOR"
+%[3] "DNA_repair"                  "DNA_repair:G1_CC_PHASE"
+%[5] "ACSN_master:AKT_MTOR"        "ACSN_master:G1_CC_PHASE"
+%[7] "Survival:S_PHASE_CHECKPOINT"
+
+\begin{table}[h!]
+  \centering
+  \caption{ACSN maps enrichment for genes differentially expressed between ER
+positive and ER negative samples in breast cancer.  Module : name of the
+module. Mod. size: size of the module. Nb genes: number of genes from input
+which are found in the module. pval: corrected p-value.}
+\label{tab:table_mainz}
+
+\begin{tabular}{l|c|c|c}
+\hline
+Module & Mod. size & Nb genes & pval \\
+\hline
+Apoptosis:HEDGEHOG &  &  & 1.04e-08 \\
+Apoptosis:AKT\_MTOR &  &  & 1.04e-08 \\
+DNA\_repair &  &  & 1.04e-08 \\
+DNA\_repair:G1\_CC\_PHASE &  &  & 1.04e-08 \\
+ACSN\_master:AKT\_MTOR &  &  & 1.04e-08 \\
+ACSN\_master:G1\_CC\_PHASE &  &  & 1.04e-08 \\
+Survival:S\_PHASE\_CHECKPOINT &  &  & 1.04e-08 \\
+\hline
+
+\end{tabular}
+\end{table}
+
+\begin{verbatim}
+# analysis mtsig signature file
+
+mtsig <- format_from_gmt('c2.cp.v5.0.symbols.gmt')
+enrichment(symbol_list, maps = mtsig)
+
+\end{verbatim}
+
+\begin{verbatim}
+apply(eset[probe_list,pdat$er == 0],1,mean) -> er_minus_mean
+names(er_minus_mean) <- symbol_list
+er_minus_mean <- as.matrix(er_minus_mean)
+colnames(er_minus_mean) <- c('exp')
+
+navicell <- NaviCell()
+navicell$proxy_url <- "https://acsn.curie.fr/cgi-bin/nv_proxy.php"
+navicell$map_url <- "https://acsn.curie.fr/navicell/maps/acsn/master/index.php"
+
+navicell$launchBrowser()
+navicell$importDatatable("mRNA expression data", "GBM_exp", er_minus_mean)
+navicell$heatmapEditorSelectSample('0','exp')
+navicell$heatmapEditorSelectDatatable('0','GBM_exp')
+navicell$heatmapEditorApply()
+\end{verbatim}
+
 %\bibliographystyle{jss}

### 2016-01-09 jss/article.tex
[skip ci] update examples.

@@
 
+dim(results)
+[1] 7 9
 \end{verbatim}
-
-%> as.character(res$module)
-%[1] "Apoptosis:HEDGEHOG"          "Apoptosis:AKT_MTOR"
-%[3] "DNA_repair"                  "DNA_repair:G1_CC_PHASE"
-%[5] "ACSN_master:AKT_MTOR"        "ACSN_master:G1_CC_PHASE"
-%[7] "Survival:S_PHASE_CHECKPOINT"
 
@@
 module. Mod. size: size of the module. Nb genes: number of genes from input
-which are found in the module. pval: corrected p-value.}
+which are found in the module. pval: raw p-value. Cor. pval: corrected p-value.}
 \label{tab:table_mainz}
 
-\begin{tabular}{l|c|c|c}
+\begin{tabular}{l|c|c|c|c}
 \hline
-Module & Mod. size & Nb genes & pval \\
+Module & Mod. size & Nb genes & pval & Cor. pval\\
 \hline
-Apoptosis:HEDGEHOG &  &  & 1.04e-08 \\
-Apoptosis:AKT\_MTOR &  &  & 1.04e-08 \\
-DNA\_repair &  &  & 1.04e-08 \\
-DNA\_repair:G1\_CC\_PHASE &  &  & 1.04e-08 \\
-ACSN\_master:AKT\_MTOR &  &  & 1.04e-08 \\
-ACSN\_master:G1\_CC\_PHASE &  &  & 1.04e-08 \\
-Survival:S\_PHASE\_CHECKPOINT &  &  & 1.04e-08 \\
+Apoptosis:HEDGEHOG & 31 & 21 & 0.0016 & 0.042 \\
+Apoptosis:AKT\_MTOR &  79 & 47 & 0.00037 & 0.038 \\
+DNA\_repair & 335 & 163 & 0.0012 & 0.038 \\
+DNA\_repair:G1\_CC\_PHASE &  25 & 18 & 0.0011 & 0.038 \\
+ACSN\_master:AKT\_MTOR & 92 & 53 & 0.0005 & 0.038 \\
+ACSN\_master:G1\_CC\_PHASE & 25 & 18 & 0.0011 & 0.038 \\
+Survival:S\_PHASE\_CHECKPOINT & 8 & 8 & 0.0006 & 0.038 \\
 \hline

### 2016-01-09 jss/article.tex
[skip ci] update examples.

@@
 
-%navicell$launchBrowser()
 \section[Examples]{Examples}
@@
 The code shown below is creating a list of differentially expressed genes
+between ER positive and ER negative samples, and calculates the enrichment in
+ACSN maps from this list of genes. As seen in Table \ref{tab:table_mainz}, there
+are four maps (Apoptosis, DNA repair, ACSN master and Survival) and seven
+modules enriched. Most of the modules are in fact linked to different phases of
+the cell cycle.
 
@@
 \hline
-
 \end{tabular}
 \end{table}
+
+The Molecular Signatures Database (MSigDB) is one of the most widely used
+repository of well-annotated  gene sets representing the universe of biological
+processes \citep{liberzon2011molecular}. We downloaded the canonical pathways
+set, counting more than 1,300 gene sets representing canonical pathways compiled
+by domain experts. The dataset is encoded with the 'gmt' format, and can be
+imported within ACSNMineR with the \code{format\_from\_gmt} function. We
+calculate the enrichment for the breast cancer differentially expressed gene
+list, simply specifying the MSigDB data we just imported as the \code{maps}
+option. Table \ref{tab:table_msigdb} is displaying the pathways having a
+corrected p-value $< 0.05$. The prefix is indicating the database source, so we
+see that we have pathways from the KEGG, Reactome and PID databases. Consistent
+with our previous results, most of the enriched pathways are related to the cell
+cycle regulation.
 
@@
 \end{verbatim}
+
+\begin{table}[h!]
+  \centering
+  \caption{MSigDB canonical pathway database enrichment for genes differentially expressed between ER
+positive and ER negative samples in breast cancer.  Module : name of the
+module. Mod. size: size of the module. Nb genes: number of genes from input
+which are found in the module. Cor. pval: corrected p-value.}
+\label{tab:table_msigdb}
+
+\begin{tabular}{l|c|c|c}
+\hline
+Pathway & Mod. size & Nb genes & Cor. pval\\
+\hline
+KEGG\_CELL\_CYCLE & 128 & 76 & 1.98e-07\\
+REACTOME\_CELL\_CYCLE\_MITOTIC & 325 & 158 & 2.40e-07\\
+REACTOME\_DNA\_REPLICATION & 192 & 98 & 1.33e-05\\
+PID\_FOXM1\_PATHWAY & 40 & 29 & 5.05e-05\\
+REACTOME\_MITOTIC\_M\_M\_G1\_PHASES & 172 & 87 & 7.38e-05\\
+REACTOME\_CELL\_CYCLE & 421 & 181 & 2.97e-04\\
+PID\_AURORA\_B\_PATHWAY & 39 & 27 & 3.38e-04\\
+REACTOME\_MITOTIC\_G1\_G1\_S\_PHASES & 137 & 70 & 3.83e-04\\ 
+PID\_SYNDECAN\_1\_PATHWAY & 46 & 30 & 4.40e-04\\
+REACTOME\_S\_PHASE & 109 & 58 & 4.40e-04\\
+\hline
+\end{tabular}
+\end{table}
 

### 2016-01-10 jss/article.tex
[skip ci] update examples.

@@
 \begin{verbatim}
-# analysis mtsig signature file
+# Import MSigDB canonical pathways and calculate enrichment on this database
 
@@
 
+At last, we visualize the mean expression values for ER negative samples for all
+genes differentially expressed on the ACSN master map using \code{ACSNMineR}
+commands to create heatmaps.
+
 \begin{verbatim}
+# Select ER negative samples and calculate mean expression values
+
 apply(eset[probe_list,pdat$er == 0],1,mean) -> er_minus_mean
@@
 colnames(er_minus_mean) <- c('exp')
+
+# create a NaviCell session, import the expression matrix on the map and create
+# heatmaps to represent the data points.
 
@@
 
+The Figure \ref{fig:mainz} is displaying the map for gene that having a
+corrected p-value $< 0.05$ and a log fold-change $>1$. we can see that the genes
+are concentrated in the regions of the map corresponding to the cell cycle, cell
+motility, apoptosis and survival.
+
+\begin{figure}[!ht]
+  \caption{Mean expression values for ER negative differentially expressed
+  genes in breast cancer visualized as heatmaps on the ACSN master map.} 
+  \centering
+  \includegraphics[width=0.8\textwidth]{figures/mainz_acsn.pdf}
+  \label{fig:mainz}
+\end{figure}
+
+\subsection{Analysis of glioblastoma mutation frequencies}
+
+\begin{table}[h!]
+  \centering
+  \caption{MSigDB canonical pathway database enrichment for genes differentially expressed between ER
+positive and ER negative samples in breast cancer.  Module : name of the
+module. Mod. size: size of the module. Nb genes: number of genes from input
+which are found in the module. Cor. pval: corrected p-value.}
+\label{tab:table_gbm}
+
+\begin{tabular}{l|c|c|c}
+\hline
+module & Mod. size & Nb genes & Cor. pval\\ 
+\hline
+CellCycle:G2\_M\_CHECKPOINT & 14 & 8 & 3.92e-02\\
+CellCycle:S\_PHASE\_CHECKPOINT & 6 & 5 & 3.92e-02\\
+DNA\_repair:S\_PHASE\_CHECKPOINT & 45 & 19 & 1.58e-02\\
+EMT\_motility & 621 & 176 & 1.06e-03\\
+EMT\_motility:CELL\_MATRIX\_ADHESIONS & 72 & 44 & 2.55e-11\\
+EMT\_motility:ECM & 137 & 64 & 2.34e-09\\
+EMT\_motility:EMT\_REGULATORS & 615 & 173 & 1.49e-03\\
+EMT\_motility:CYTOSKELETON\_POLARITY & 154 & 47 & 3.92e-02\\
+EMT\_motility:DESMOSOMES & 29 & 15 & 4.62e-03\\
+ACSN\_master:CELL\_MATRIX\_ADHESIONS & 72 & 44 & 2.55e-11\\
+ACSN\_master:DESMOSOMES & 29 & 15 & 4.62e-03\\
+ACSN\_master:S\_PHASE\_CHECKPOINT & 46 & 20 & 8.23e-03\\
+ACSN\_master:ECM & 138 & 64 & 2.56e-09\\
+ACSN\_master:EMT\_REGULATORS & 625 & 177 & 1.06e-03\\
+ACSN\_master:CYTOSKELETON\_POLARITY & 154 & 47 & 3.92e-02\\
+Survival:ECM & 23 & 11 & 3.92e-02\\
+\hline
+\end{tabular}
+\end{table}
+
+\begin{figure}[!ht]
+  \caption{Mean expression values for ER negative differentially expressed
+  genes in breast cancer visualized as heatmaps on the ACSN master map.} 
+  \centering
+  \includegraphics[width=0.8\textwidth]{figures/gbm.pdf}
+  \label{fig:mainz}
+\end{figure}
+
 %\bibliographystyle{jss}

### 2016-01-10 jss/article.tex
[skip ci] update examples.

@@
 
+Recent years have witnessed a dramatic increase in new technologies for
+interrogating the activity levels of various cellular components on a
+genome-wide scale, including genomic, epigenomic, transcriptomic, and proteomic
+information \citep{hawkins2010next}. Integrating these
+heterogeneous datasets provides more biological insights than performing
+separate analyses.
+
+For instance, international consortia such as The Cancer Genome Atlas (TCGA)
+have launched large-scale
+initiatives to characterize multiple types of cancer at different levels
+(genomic, transcriptomic, epigenomic, etc.) on several hundreds of samples.
+These integrative studies have already led to the identification of novel cancer
+genes \citep{mclendon2008comprehensive}.
+
+Glioblastoma multiforme (GBM) is the most malignant form of gliomas, and despite
+intense investigation of this disease in the past decades, most patients with
+GBM die within approximately 15 months of diagnosis \citep{maher2001malignant}.
+
 \begin{table}[h!]

### 2016-01-10 jss/article.tex
[skip ci] update gbm.

@@
 
+\begin{verbatim}
+library(RNaviCell)
+
+n <- NaviCell()
+
+n$proxy_url <- "https://acsn.curie.fr/cgi-bin/nv_proxy.php"
+n$map_url <- "https://acsn.curie.fr/navicell/maps/acsn/master/index.php"
+
+n$launchBrowser()
+
+mat <- n$readDatatable('gbm.txt')
+n$importDatatable("Mutation data", "GBM", mat)
+
+n$drawingConfigSelectGlyph(1, TRUE)
+
+n$glyphEditorOpen(1)
+n$glyphEditorSelectSample(1, "categ")
+n$glyphEditorSelectShapeDatatable(1, "GBM")
+n$glyphEditorSelectColorDatatable(1, "GBM")
+n$glyphEditorSelectSizeDatatable(1, "GBM")
+n$glyphEditorApply(1)
+
+n$unorderedConfigSetDiscreteShape("GBM", "sample", 0, 1)
+n$unorderedConfigSetDiscreteShape("GBM", "sample", 1, 5)
+n$unorderedConfigApply("GBM", "shape")
+
+n$unorderedConfigSetDiscreteColor("GBM", "sample", 0, "398BC3")
+n$unorderedConfigSetDiscreteColor("GBM", "sample", 1, "CC5746")
+n$unorderedConfigApply("GBM", "color")
+
+n$unorderedConfigSetDiscreteSize("GBM", "sample", 0, 4)
+n$unorderedConfigSetDiscreteSize("GBM", "sample", 1, 14)
+n$unorderedConfigApply("GBM", "size")
+
+\end{verbatim}
+
 \begin{figure}[!ht]

### 2016-01-11 jss/article.tex
[skip ci] gbm example update

--- 2016-01-10 jss/article.tex
@@
 
-\section[Package organization]{Packages organization}
+\section[Packages organization]{Packages organization}
 
@@
 
+Recent years have witnessed a dramatic increase in new technologies for
+interrogating the activity levels of various cellular components on a
+genome-wide scale, including genomic, epigenomic, transcriptomic, and proteomic
+information \citep{hawkins2010next}. Integrating these heterogeneous datasets
+provides more biological insights than performing separate analyses. For
+instance, international consortia such as The Cancer Genome Atlas (TCGA) have
+launched large-scale initiatives to characterize multiple types of cancer at
+different levels on hundreds of samples.  These integrative studies have
+already led to the identification of novel cancer genes
+\citep{mclendon2008comprehensive}.
+
+Malignant gliomas, the most common subtype of primary brain tumors, are
+aggressive, highly invasive, and neurologically destructive tumors considered to
+be among the deadliest of human cancers. In its most aggressive manifestation,
+glioblastoma (GBM), median survival ranges from 9 to 12 months, despite maximum
+treatment efforts \citep{maher2001malignant}. In this study we have analyzed
+whole-genome mutation data generated by the TCGA project on hundreds of
+patients. More specifically, we parsed the MAF (Mutation Annotation Format) GBM
+files produced by different sequencing centers to count and calculate gene
+mutation frequencies. We kept the mutations having a status likely to disturb
+the target protein's function (i.e, Frame\_Shift\_Del, Frame\_Shift\_Ins,
+In\_Frame\_Del, In\_Frame\_Ins, Missense\_Mutation, Nonsense\_Mutation). In
+total, we collected mutations for xxx genes in a total of xxx samples. In order
+to retain the most frequently mutated genes, we calculated frequencies across
+all mutated samples, and kept genes having a frequency greater than 1\%. We
+further labelled genes having a frequency greater than 1\% and less than 5\% as "1"
+and genes highly mutated (frequency higher than 5\%) as "2".
+
+We loaded the data as a matrix in R and calculated the enrichment in ACSN maps
+with the ACSNMineR function \code{enrichment}. The results are displayed in
+table \ref{tab:table_gbm}. There are 16 modules significantly enriched in the
+Cell Cycle, DNA repair, EMT motility and Survival. Cell matrix adhesions and ECM
+(extra cellular matrix) are the modules with highest significance (corrected
+p-values of 2.55e-11 and 2.34e-9 respectively). Visualization of the list of
+frequently mutated genes is shown on figure \ref{fig:gbm}. This figure was
+generated with the ACSNMineR commands detailed below.
+
 \begin{table}[h!]
@@
 
+\begin{verbatim}
+library(RNaviCell)
+
+# Create a NaviCell object, point it to the ACSN master map and launch
+# a session.
+
+navicell <- NaviCell()
+navicell$proxy_url <- "https://acsn.curie.fr/cgi-bin/nv_proxy.php"
+navicell$map_url <- "https://acsn.curie.fr/navicell/maps/acsn/master/index.php"
+navicell$launchBrowser()
+
+# Read the GBM data file and import it into the session.
+
+mat <- navicell$readDatatable('gbm.txt')
+navicell$importDatatable("Mutation data", "GBM", mat)
+
+# set datatable and sample names for the glyph editor
+
+navicell$drawingConfigSelectGlyph(1, TRUE)
+navicell$glyphEditorSelectSample(1, "categ")
+navicell$glyphEditorSelectShapeDatatable(1, "GBM")
+navicell$glyphEditorSelectColorDatatable(1, "GBM")
+navicell$glyphEditorSelectSizeDatatable(1, "GBM")
+navicell$glyphEditorApply(1)
+
+# set color, shape and size parameters for glyphs
+
+navicell$unorderedConfigSetDiscreteShape("GBM", "sample", 0, 1)
+navicell$unorderedConfigSetDiscreteShape("GBM", "sample", 1, 5)
+navicell$unorderedConfigApply("GBM", "shape")
+
+navicell$unorderedConfigSetDiscreteColor("GBM", "sample", 0, "398BC3")
+navicell$unorderedConfigSetDiscreteColor("GBM", "sample", 1, "CC5746")
+navicell$unorderedConfigApply("GBM", "color")
+
+navicell$unorderedConfigSetDiscreteSize("GBM", "sample", 0, 4)
+navicell$unorderedConfigSetDiscreteSize("GBM", "sample", 1, 14)
+
+navicell$unorderedConfigApply("GBM", "size")
+\end{verbatim}
+
 \begin{figure}[!ht]
@@
   \includegraphics[width=0.8\textwidth]{figures/gbm.pdf}
-  \label{fig:mainz}
+  \label{fig:gbm}
 \end{figure}

### 2016-01-11 jss/article.tex
[skip ci] gbm example update

@@
 Cell Cycle, DNA repair, EMT motility and Survival. Cell matrix adhesions and ECM
-(extra cellular matrix) are the modules with highest significance (corrected
-p-values of 2.55e-11 and 2.34e-9 respectively). Visualization of the list of
-frequently mutated genes is shown on figure \ref{fig:gbm}. This figure was
-generated with the ACSNMineR commands detailed below.
+(extra cellular matrix), part of the EMT motility map, are the modules with
+highest significance (corrected p-values of 2.55e-11 and 2.34e-9 respectively).
 
@@
 \end{table}
+
+Visualization of the list of frequently mutated genes is shown on figure
+\ref{fig:gbm}. This figure was generated with the ACSNMineR commands detailed
+below. Results of the enrichment test correlate well with the visualization on
+the map, with a high density of low and high frequency mutated genes in the EMT
+motility , survival, cell cycle and DNA repair regions (maps) of the global ACSN
+map.
 

### 2016-01-11 jss/article.tex
[skip ci] update examples

@@
   \centering
-  \caption{MSigDB canonical pathway database enrichment for genes differentially expressed between ER
-positive and ER negative samples in breast cancer.  Module : name of the
-module. Mod. size: size of the module. Nb genes: number of genes from input
-which are found in the module. Cor. pval: corrected p-value.}
+  \caption{ACSN maps enrichment for frequently mutated glioblastoma genes.
+  Module : name of the module. Mod. size: size of the module. Nb genes: number
+  of genes from input which are found in the module. Cor. pval: corrected
+  p-value.} 
   \label{tab:table_gbm}
@@
 
-Visualization of the list of frequently mutated genes is shown on figure
+Visualization of the list of glioblastoma mutated genes is shown on figure
 \ref{fig:gbm}. This figure was generated with the ACSNMineR commands detailed
@@
 \begin{figure}[!ht]
-  \caption{Mean expression values for ER negative differentially expressed
-  genes in breast cancer visualized as heatmaps on the ACSN master map.} 
+  \caption{Glioblastoma gene mutation frequency categories represented as
+  glyphs on the ACSN global cancer map. High frequency mutated genes are
+  pictured as large red circles, while low frequency mutated genes are depicted
+  as small blue squares.
+  } 
   \centering

### 2016-01-12 jss/article.tex
[skip ci] abstract and acknowledgments

--- 2016-01-11 jss/article.tex
@@
         Eric Bonnet\\Institut Curie\\PSL Research University\\INSERM U900\\Mines-ParisTech}
-\title{Calculating Module Enrichment and Visualizing Data on Large-scale Molecular Maps with the R packages \pkg{ACSNMineR} and \pkg{RNaviCell}}
+\title{Calculating Biological Module Enrichment or Depletion and Visualizing Data on Large-scale Molecular Maps with the R packages \pkg{ACSNMineR} and \pkg{RNaviCell}}
 
@@
 \Plainauthor{Paul Deveau, Eric Bonnet} %% comma-separated
-\Plaintitle{Calculating Module Enrichment and Visualizing Data on Large-scale Molecular Maps with the R packages ACSNMineR and RNaviCel} %% without formatting
+\Plaintitle{Calculating Biological Module Enrichment or Depletion and Visualizing Data on Large-scale Molecular Maps with the R packages ACSNMineR and RNaviCell} %% without formatting
 \Shorttitle{Module Enrichment and Data Visualization with ACSNMineR and RNaviCell} %% a short title (if necessary)
@@
 \Abstract{
-  The abstract of the article.
+Biological pathways or modules represent sets of interactions or functional
+relationships occurring at the molecular level in living cells. A large body of
+knowledge on pathways is organized in public databases such as the KEGG,
+Reactome, MINT databases, but also in more specialized repositories such as the
+recently released Atlas of Cancer Signaling Networks (ACSN). All these open and
+popular biological databases facilitate a large spectrum of analyses, thereby
+improving our understanding of cellular systems. Here we describe the
+\pkg{ACSNMineR}, a novel R package for calculation of enrichment or depletion
+of list of genes of interest in biological pathways. The package has built-in
+support for ACSN molecular pathways, but can use any molecular pathway database
+encoded as a GMT file, such as any of the gene signatures available in the
+Molecular Signatures Database (MSigDB). We also present the R package
+\pkg{RNaviCell}, that can be used in conjunction with \pkg{ACSNMineR} to
+visualize different data types on web-based, interactive ACSN maps. We
+illustrate the functionalities of the two packages with real-world biological
+data taken from large-scale cancer datasets.
+
 }
-\Keywords{keywords, comma-separated, not capitalized, \proglang{Java}}
-\Plainkeywords{keywords, comma-separated, not capitalized, Java} %% without formatting
+\Keywords{biological module, enrichment, depletion, cancer, maps, molecular networks, molecular pathways, \proglang{R}}
+\Plainkeywords{biological module, enrichment, depletion, cancer, maps, molecular networks, molecular pathways, R} %% without formatting
 %% at least one keyword must be supplied
@@
 
-We have recently released the Atlas of Cancer Signalling Network (ACSN), a
+We have recently released the Atlas of Cancer Signalling Networks (ACSN), a
 web-based database which describes signaling and regulatory molecular processes
@@
 
-\section[Examples]{Examples}
+\section[Case studies]{Case studies}
 
@@
 
+\section[Acknowledgments]{Acknowledgements}
+
+The authors would like to thank Pierre Gestraud for his comments on early
+versions of the \pkg{ACSNMineR} package and Eric Viara for guidance and
+assistance on the development of the \pkg{RNaviCell} package.
+
 %\bibliographystyle{jss}

### 2016-01-13 jss/article.tex
[skip ci] summary

@@
 
+\section[Discussion]{Discussion}
+In this work, we presented the R package \pkg{ACSNMineR}, a novel package for
+the calculation of p-values for enrichment or depletion of genes in biological
+pathways (also called modules here). The main function of the package,
+\code{enrichment}, offers many options to set the statistical test used (either
+the Fisher exact test or the hypergeometric test), the modules (or maps) used,
+the universe of reference for the calculations, the miminal size for a module to
+be taken into account, the choice for over or under-representation. Additional
+functions allow users to create graphical representations for p-values
+(\code{represent\_enrichment}) or to perform calculations for multiple samples
+or groups (\code{multisample\_enrichment}). Large-scale molecular maps of the
+Atlas of Cancer Signaling Networks (ACSN) are included in the package (six
+maps), but any other molecular pathway (such as those defined in the REACTOME
+database) can be imported and used for calculations with the command
+\code{format\_from\_gmt}.
+
+We also describe in this work the \pkg{RNaviCell} package, a R package highly
+complementary to \pkg{ACSNMineR}. This package is dedicated to create data
+visualization on ACSN maps, and is communicating with the NaviCell web server to
+create interactive graphical representations. Users might use this tools to
+represent genes of interest that have been shown to be related to the maps by
+calculating enrichment with the  \pkg{ACSNMineR}.  Creating maps with the
+graphical user interface of the ACSN website can be a tedious task if the user
+has multiple samples or gene lists, and wants to compare their representations
+on ACSN maps. The \pkg{RNaviCell} package can be used to automate the process of
+creating the graphical representations automatically.
+
+Of course, \pkg{ACSNMineR} is not the only R package for enrichment calculations. For
+instance, \pkg{GOstats} \citep{falcon2007using} is probably one of the first
+packages that was created to calculate enirchment for Gene Ontology categories.
+\pkg{GOstats} can also be used to calculate enrichment for other biological
+pathways categories, such as KEGG pathways (by using an instance of the class
+\code{KEGGHyperGParams}) or PFAM protein families (using
+\code{PFAMHyperGParams}). However, its usage might not be as straightforward as
+\pkg{ACSNMineR}, and it does not seems possible to test user-defined biological
+pathways. Furthermore, other authors have pointed out that the KEGG database
+used by this package has not been updated since 2012. \pkg{clusterProfiler} is a
+recent R package released for the
+
 \section[Acknowledgments]{Acknowledgements}

### 2016-01-14 jss/article.tex
[skip ci] small corrections

--- 2016-01-12 jss/article.tex
@@
 }
-\Keywords{biological module, enrichment, depletion, cancer, maps, molecular networks, molecular pathways, \proglang{R}}
-\Plainkeywords{biological module, enrichment, depletion, cancer, maps, molecular networks, molecular pathways, R} %% without formatting
+\Keywords{biological module, biological pathway, enrichment, depletion, cancer, maps, molecular networks, molecular pathways, \proglang{R}}
+\Plainkeywords{biological module, biological pathway, enrichment, depletion, cancer, maps, molecular networks, molecular pathways, R} %% without formatting
 %% at least one keyword must be supplied
@@
 experiments, such as mRNA or protein expression levels, with curated biological
-pathways in order to visualize changes, analyze their impact on a network and
+pathways in order to visualize the changes, analyze their impact on a network and
 formulate new hypotheses about biological processes. Many biologists and
@@
 that occur in a healthy mammalian cell but that are frequently deregulated
-during cancerogenesis \citep{kuperstein2013navicell}.  The ACSN atlas aims to
+during cancerogenesis \citep{kuperstein2015atlas}.  The ACSN atlas aims to
 be a comprehensive description of cancer-related mechanisms retrieved from the
@@
 vance2009data}. The package \pkg{ACSNMineR} is designed for the calculation of
-gene enrichment and depletion in ACSN maps, while \pkg{RNaviCell} is dedicated
-to data visualization on ACSN maps. Both packages are available on the
-Comprehensive R Archive Network
-(\url{https://cran.r-project.org/web/packages/ACSNMineR/} and
+gene enrichment and depletion in ACSN maps (or any user-defined gene set via
+the import function), while \pkg{RNaviCell} is dedicated to data visualization
+on ACSN maps. Both packages are available on the Comprehensive R Archive
+Network (\url{https://cran.r-project.org/web/packages/ACSNMineR/} and
 \url{https://cran.r-project.org/web/packages/RNaviCell/}), and on the GitHub
@@
 allows representation of results from \code{enrichment} or
-\code{multisample_enrichment} with a limited number of parameters. For
-examples, two types of display are available: heat-map tiles or bars. For
-multiple samples using a barplot representation the number of rows used can be
-provided, all plots will be on the same row otherwise. For the heatmap, the
-color of the non-significant modules, and boundaries of the gradient for
-significant values can also be tuned.
+\code{multisample_enrichment} with a limited number of parameters. Two types of
+display are available: heat-map tiles or bars. For multiple samples using a
+barplot representation, the number of rows used can be provided, otherwise all
+plots will be on the same row. For the heatmap, the color of the
+non-significant modules, and boundaries of the gradient for significant values
+can also be tuned.
 
@@
   \centering
-  \caption{Overview of results from enrichment analysis without correction.
+  \caption{Overview of the results from enrichment analysis without correction.
   Module : name of the module. Mod. size: size of the module. Genes in module:
@@
 level. The essential phases of the cell cycle are indicated on the map
-(G1/S/G2/M).} 
+(G1/S/G2/M). Note that on the web browser the map is interactive and the user
+can zoom in and out, change the graphical parameters, import additional data
+and perform functional analysis.  } 
 \centering
@@
 At last, we visualize the mean expression values for ER negative samples for all
-genes differentially expressed on the ACSN master map using \code{ACSNMineR}
+genes differentially expressed on the ACSN master (global) map using \code{ACSNMineR}
 commands to create heatmaps.
@@
 
+\section[Summary and perspectives]{Summary and perspectives}
+In this work, we presented the R package \pkg{ACSNMineR}, a novel package for
+the calculation of p-values for enrichment or depletion of genes in biological
+pathways. The package includes the six large-scale molecular maps and 55
+functional modules of the Atlas of Cancer Signaling Networks (ACSN) . Enrichment can be
+calculated for those maps and modules with several options to play with, but can
+also be calculated for other databases of molecular pathways, that can be
+imported from GMT formated files.
+
+We also describe in this work the \pkg{RNaviCell} package, a R package highly
+complementary to \pkg{ACSNMineR}. This package is dedicated to create web-based
+and interactive data visualization on ACSN maps. Users can use this tools to
+represent genes of interest that have been shown to be related to the maps by
+calculating enrichment with the  \pkg{ACSNMineR}.  Creating maps with the
+graphical user interface of the ACSN website can be a tedious task if the user
+has multiple samples or gene lists, and wants to compare their representations
+on ACSN maps. The \pkg{RNaviCell} package can be used to automate the process of
+creating the graphical representations automatically.
+
+We have shown how the packages \pkg{ACSNMineR} and \pkg{RNaviCell} can be
+combined to analyze expression data from breast cancer samples, and also to
+analyze the frequency of mutated genes in glioblastoma cancer samples.
+
+Of course, \pkg{ACSNMineR} is not the only R package for enrichment
+calculations. For instance, \pkg{GOstats} \citep{falcon2007using} is probably
+one of the first packages that was created to calculate enirchment for Gene
+Ontology categories.  \pkg{GOstats} can also be used to calculate enrichment for
+other biological pathways categories, such as KEGG pathways (by using an
+instance of the class \code{KEGGHyperGParams}) or PFAM protein families (using
+\code{PFAMHyperGParams}). However, its usage might not be as straightforward as
+\pkg{ACSNMineR}, and it does not seems possible to test user-defined biological
+pathways. Furthermore, other authors have pointed out that the KEGG database
+used by this package has not been updated since 2012. \pkg{clusterProfiler} is a
+recent R package released for enrichment analysis of Gene Ontology and KEGG with
+either hypergeometric test or Gene Set Enrichment Analysis (GSEA)
+\citep{yu2012clusterprofiler}. Via other packages, support for analysis of
+Disease Ontology and Reactome Pathways is possible. Interestingly, this package
+also offers the possibility to import user-defined gene set, through
+tab-delimited pairwise definition files.
+
+In order to improve \pkg{ACSNMineR}, we may in the near future try to improve
+the speed of calculation, which might be a problem if a very large number of
+samples or experiments have to be analyzed rapidly. For instance, we could use
+the \code{foreach} and \code{\%dopar\%} operator to parallelize the most
+computationally demanding operations. It could also be useful to implement more
+sensitive methods of gene set enrichment measures, such as the Gene Set
+Enrichment Analysis (GSEA) method.
+
 \section[Acknowledgments]{Acknowledgements}

### 2016-01-15 jss/article.tex
[skip ci] small corrections

@@
 Malignant gliomas, the most common subtype of primary brain tumors, are
-aggressive, highly invasive, and neurologically destructive tumors considered to
-be among the deadliest of human cancers. In its most aggressive manifestation,
-glioblastoma (GBM), median survival ranges from 9 to 12 months, despite maximum
-treatment efforts \citep{maher2001malignant}. In this study we have analyzed
-whole-genome mutation data generated by the TCGA project on hundreds of
-patients. More specifically, we parsed the MAF (Mutation Annotation Format) GBM
-files produced by different sequencing centers to count and calculate gene
-mutation frequencies. We kept the mutations having a status likely to disturb
-the target protein's function (i.e, Frame\_Shift\_Del, Frame\_Shift\_Ins,
-In\_Frame\_Del, In\_Frame\_Ins, Missense\_Mutation, Nonsense\_Mutation). In
-total, we collected mutations for xxx genes in a total of xxx samples. In order
-to retain the most frequently mutated genes, we calculated frequencies across
-all mutated samples, and kept genes having a frequency greater than 1\%. We
-further labelled genes having a frequency greater than 1\% and less than 5\% as "1"
-and genes highly mutated (frequency higher than 5\%) as "2".
+aggressive, highly invasive, and neurologically destructive tumors considered
+to be among the deadliest of human cancers. In its most aggressive
+manifestation, glioblastoma (GBM), median survival ranges from 9 to 12 months,
+despite maximum treatment efforts \citep{maher2001malignant}. In this study we
+have analyzed whole-genome mutation data generated by the TCGA project on
+hundreds of patients. More specifically, we parsed the MAF (Mutation Annotation
+Format) GBM files produced by different sequencing centers to count and
+calculate gene mutation frequencies. We kept the mutations having a status
+likely to disturb the target protein's function (i.e, Frame\_Shift\_Del,
+Nonstop\_Mutation, In\_Frame\_Del, In\_Frame\_Ins, Missense\_Mutation,
+Nonsense\_Mutation, Splice\_Site, Translation\_Start\_Site). In total, we
+collected mutations for more than 13,000 genes in a total of 379 mutated
+samples. In order to retain the most frequently mutated genes, we calculated
+frequencies across all mutated samples, and kept genes having a frequency
+greater than 1\% (3,293 genes). We further labelled genes having a frequency
+greater than 1\% and less than 5\% as "1" and genes highly mutated (frequency
+higher than 5\%) as "2".
 

### 2016-01-18 jss/article.tex
[skip ci] small corrections

@@
 for module of small size.  The universe in which the computation is made by
-default is defined by all the gene symbols conatined in the maps. All the genes
+default is defined by all the gene symbols contained in the maps. All the genes
 that were given as input and that are not present on the maps will be
@@
 calculations. For instance, \pkg{GOstats} \citep{falcon2007using} is probably
-one of the first packages that was created to calculate enirchment for Gene
+one of the first packages that was created to calculate enrichment for Gene
 Ontology categories.  \pkg{GOstats} can also be used to calculate enrichment for

### 2016-01-22 jss/article.tex
[skip ci] update ACK

@@
 %% almost as usual
-\author{Paul Deveau\\Institut Curie\\PSL Research University\\Univ. Paris-Sud\\INSERM U830\\INSERM U900\\Mines-ParisTech \And 
-        Eric Bonnet\\Institut Curie\\PSL Research University\\INSERM U900\\Mines-ParisTech}
+\author{Paul Deveau\\Institut Curie\\PSL University\\Univ. Paris-Sud\\INSERM U830\\INSERM U900\\Mines-ParisTech \And 
+        Emmanuel Barillot\\Institut Curie\\PSL University\\INSERM U900\\Mines-ParisTech \And 
+        Andrei Zinovyev\\Institut Curie\\PSL University\\INSERM U900\\Mines-ParisTech \And
+        Eric Bonnet\\Institut Curie\\PSL University\\INSERM U900\\Mines-ParisTech}
 \title{Calculating Biological Module Enrichment or Depletion and Visualizing Data on Large-scale Molecular Maps with the R packages \pkg{ACSNMineR} and \pkg{RNaviCell}}
@@
 %% for pretty printing and a nice hypersummary also set:
-\Plainauthor{Paul Deveau, Eric Bonnet} %% comma-separated
+\Plainauthor{Paul Deveau, Emmanuel Barillot, Andrei Zinovyev, Eric Bonnet} %% comma-separated
 \Plaintitle{Calculating Biological Module Enrichment or Depletion and Visualizing Data on Large-scale Molecular Maps with the R packages ACSNMineR and RNaviCell} %% without formatting
@@
 
-The authors would like to thank Pierre Gestraud for his comments on early
-versions of the \pkg{ACSNMineR} package and Eric Viara for guidance and
-assistance on the development of the \pkg{RNaviCell} package.
+This work was supported by a grant ``Projet Incitatif et Collaboratif
+Computational Systems Biology Approach for Cancer" from Institut Curie. The
+authors would like to thank Pierre Gestraud for his comments on early versions
+of the \pkg{ACSNMineR} package and Eric Viara for guidance and assistance on
+the development of the \pkg{RNaviCell} package.
 

### 2016-01-25 jss/scripts/DU145_heatmap.R
[skip ci] added scripts to reproduce results

@@ -1,0 +1,25 @@
+# a short RNaviCell script example
+
+# load RNaviCell library
+
+library(RNaviCell)
+
+# create a NaviCell object and launch a server session
+# this will automatically open a browser on the client
+
+navicell <- NaviCell()
+navicell$launchBrowser()
+
+# import a gene expression matrix and
+# send the data to the NaviCell server
+# NB: the data_matrix object is a regular R matrix
+
+data_matrix <- navicell$readDatatable('DU145_data.txt')
+navicell$importDatatable("mRNA expression data", "DU145", data_matrix)
+
+# set data set and sample for heat map representation
+
+navicell$heatmapEditorSelectSample('0','data')
+navicell$heatmapEditorSelectDatatable('0','DU145')
+navicell$heatmapEditorApply()
+

### 2016-01-25 jss/scripts/GBM_mutations_glyphs.R

@@ -1,0 +1,39 @@
+library(RNaviCell)
+
+# Create a NaviCell object, point it to the ACSN master map and launch
+# a session.
+
+navicell <- NaviCell()
+navicell$proxy_url <- "https://acsn.curie.fr/cgi-bin/nv_proxy.php"
+navicell$map_url <- "https://acsn.curie.fr/navicell/maps/acsn/master/index.php"
+navicell$launchBrowser()
+
+# Read the GBM data file and import it into the session.
+
+mat <- navicell$readDatatable('gbm.txt')
+navicell$importDatatable("Mutation data", "GBM", mat)
+
+# set datatable and sample names for the glyph editor
+
+navicell$drawingConfigSelectGlyph(1, TRUE)
+navicell$glyphEditorSelectSample(1, "categ")
+navicell$glyphEditorSelectShapeDatatable(1, "GBM")
+navicell$glyphEditorSelectColorDatatable(1, "GBM")
+navicell$glyphEditorSelectSizeDatatable(1, "GBM")
+navicell$glyphEditorApply(1)
+
+# set color, shape and size parameters for glyphs
+
+navicell$unorderedConfigSetDiscreteShape("GBM", "sample", 0, 1)
+navicell$unorderedConfigSetDiscreteShape("GBM", "sample", 1, 5)
+navicell$unorderedConfigApply("GBM", "shape")
+
+navicell$unorderedConfigSetDiscreteColor("GBM", "sample", 0, "398BC3")
+navicell$unorderedConfigSetDiscreteColor("GBM", "sample", 1, "CC5746")
+navicell$unorderedConfigApply("GBM", "color")
+
+navicell$unorderedConfigSetDiscreteSize("GBM", "sample", 0, 4)
+navicell$unorderedConfigSetDiscreteSize("GBM", "sample", 1, 14)
+
+navicell$unorderedConfigApply("GBM", "size")
+

### 2016-01-25 jss/scripts/gbm_enrichment.R

@@ -1,0 +1,15 @@
+# calculate ACSN maps enrichment for frequently mutated GBM genes
+
+# load library
+library(ACSNMineR)
+
+# read gbm data
+read.table('gbm.txt', header=T) -> gbm
+
+# create a list from GBM genes and calculate enrichment statistics
+gbm_list <- as.character(gbm$gbm)
+results <- enrichment(gbm_list)
+
+# write results to a text file
+write.table(results[,c(1,2,3,8)], file='gbm_enrichment.txt', sep='\t', quote=F, row.names=F)
+

### 2016-01-27 R/ACSN_Enrichment.R
[Plots]
- Changed log scaling to log10 scaling
- Added "reverselog" that scales with -log10
- updated figures
- updated paper

@@ -639,18 +639,27 @@ represent_enrichment<-function(enrichment, plot = "heatmap" , scale = "log",
                                              y = "module", 
                                              fill = "p.values"))+ggplot2::xlab("")+ ggplot2::ylab("Modules") + ggplot2::geom_tile()
       if(scale == "log"){
-        q<- q + ggplot2::scale_fill_gradient("p-values",low = low , high = high, na.value = na.value, trans = "log")
+        q<- q + ggplot2::scale_fill_gradient("p-values",low = low , high = high, na.value = na.value, trans = "log10")
       }
+      else if(scale == "reverselog"){
+        q<- q + ggplot2::scale_fill_gradient("p-values",low = high , high = low, na.value = na.value, trans = "reverselog")
+      }
+      
       else{
         q<-q+ ggplot2::scale_fill_gradient("p-values",low = low , high = high, na.value = na.value)
       }
       
     }
     else{
-      q<-ggplot2::qplot(x = enrichment$module, y = enrichment$p.values,
+      q<-ggplot2::ggplot(data= enrichment,
+                          aes_string(x= "module",
+                                    y= "p.values"),
                         xlab = "Modules", ylab = "p-values")
       if(scale == "log"){
-        q<- q + ggplot2::scale_y_continuous(trans = "log")
+        q<- q + ggplot2::scale_y_continuous(trans = "log10")
+      }
+      else if(scale == "reverselog"){
+        q<- q + ggplot2::scale_y_continuous(trans = "reverselog")
       }
     }    
     q<-q+ggplot2::theme_minimal()+ggplot2::theme(axis.text.x = element_text(angle = 90, hjust = 0), axis.ticks = ggplot2::element_blank())
@@ -752,7 +761,10 @@ represent_enrichment<-function(enrichment, plot = "heatmap" , scale = "log",
                                     y = "module", 
                                     fill = "p.values"))+ggplot2::xlab("")+ ggplot2::ylab("Modules") + ggplot2::geom_tile()
       if(scale == "log"){
-        q<- q + ggplot2::scale_fill_gradient("p-values",low = low , high = high, na.value = na.value, trans = "log")
+        q<- q + ggplot2::scale_fill_gradient("p-values",low = low , high = high, na.value = na.value, trans = "log10")
+      }
+      else if(scale == "reverselog"){
[…]

### 2016-01-28 jss/article.tex
[Paper]
Included Emmanuel's corrections

@@
 Thus the information included in ACSN is organized in three hierarchical levels:
-a global map, five individual maps, and a total of 55 functional modules. Each
+a global map, five individual maps, and a total of 55 functional modules, as of today. Each
 ACSN map covers hundreds of molecular players, biochemical reactions and causal
@@
 	\centering
-	\caption{Representation of the enriched modules (first ten rows), either with either Bonferroni correction or no correction.
+	\caption{Representation of the enriched modules (first ten rows), with either Bonferroni correction or no correction.
 Grey tiles means that the data is not available for this module in this sample.
@@
 \begin{figure}
-	\caption{Representation of the enriched modules (first ten rows), either with either Bonferroni correction (left) or no correction (right).
+	\caption{Representation of the enriched modules (first ten rows), with either Bonferroni correction (left) or no correction (right).
 The modules are on the X axis and the p-values are on the Y axis.

### 2016-01-28 jss/article.tex
Added script to reproduce ACSNMineR paper results
- edit in Pdf plot lines (coherent with script)
- script to:
* show data structure
* compute enrichment
* generate plots

@@
 
-represent_enrichment(enrichment = list(SampleA = enrichment_test[1:10,], 
-  SampleB = enrichment_test[3:10,]), plot = "heatmap", scale = "log", 
+represent_enrichment(enrichment = list(Corrected = results[1:10,], 
+Uncorrected = results_uncorrected[1:10,]),
+                                plot = "heatmap", scale = "log", 
                                 low = "steelblue" , high ="white", na.value = "grey")
@@
 
-represent_enrichment(enrichment = list(SampleA = enrichment_test[1:10,],
- SampleB = enrichment_test[3:10,]), plot = "bar", scale = "log", nrow = 1)
+ represent_enrichment(enrichment = list(Corrected = results[1:10,], 
+                             Uncorrected = results_uncorrected[1:10,]),
+                               plot = "heatmap", scale = "reverselog", 
+                               nrow = 1)
 \end{verbatim}

### 2016-01-28 jss/scripts/ACSNMineR.R

@@ -1,0 +1,53 @@
+### Script to reproduce results and figures from ACSNMineR paper
+
+library(ACSNMineR)
+
+### print ACSN map structure:
+print(length(ACSNMineR::ACSN_maps))
+print(names(ACSNMineR::ACSN_maps))
+
+for(map in names(ACSNMineR::ACSN_maps)){
+  print(map)
+  MAP<-ACSNMineR::ACSN_maps[[map]]
+  print(paste("Number of genes:",length(unique(as.character(MAP[,-(1:2)])))))
+  print(paste("Number of modules:",nrow(MAP)))
+  print(paste("Smallest module size:",min(MAP[,2])))
+  print(paste("Biggest module size:",max(MAP[,2])))
+  print(paste("Mean module size:", round(MAP[,2])))
+  
+}
+
+### Print genes test
+
+print(genes_test)
+
+### compute corrected enrichment
+results<-enrichment(genes_test)
+
+print(dim(results))
+
+print(results[1,])
+
+### compute uncorrected enrichment
+
+results_uncorrected<-enrichment(genes_test,correction_multitest = FALSE)
+print(head(results_uncorrected,3))
+
+
+### Plot heatmap
+heatmap <- represent_enrichment(enrichment = list(Corrected = results[1:10,], 
+                                       Uncorrected = results_uncorrected[1:10,]),
+                                plot = "heatmap", scale = "log", 
+                                low = "steelblue" , high ="white", na.value = "grey")
+
+print(heatmap)
+
+### Plot bar
+baplot <- represent_enrichment(enrichment = list(Corrected = results[1:10,], 
+                                                 Uncorrected = results_uncorrected[1:10,]),
+                               plot = "heatmap", scale = "reverselog", 
+                               nrow = 1)
+print(barplot)
+
+
+### End of ACSNMineR demo

### 2016-01-29 R/ACSN_Enrichment.R
[Update] Default testing
- enrichment can now have as input the name of acsn maps as input
- default is set to master (testing only on master map)

@@ -10,7 +10,7 @@
 #' 
 #' @param Genes Character vector of genes that should be tested for enrichment
 #' @param maps list of maps generated by format_from_gmt. Names of element of list will be used to track modules.
-#' Default: tests on all acsn maps.
+#' Default: tests on the master map.
 #' @param correction_multitest either FALSE, "bonferroni", "holm", "hochberg", "hommel", "BH", "fdr" (identical to BH), or "BY"
 #' @param statistical_test one of "fisher", "hypergeom"
 #' @param min_module_size will remove from the analysis all modules which are (strictly) smaller than threshold
@@ -29,7 +29,7 @@
 #' @importFrom stats fisher.test p.adjust phyper
 #' @importFrom utils read.csv
 enrichment<-function(Genes=NULL,
-                     maps = ACSNMineR::ACSN_maps, 
+                     maps = "Master", 
                      correction_multitest = "BH",
                      statistical_test = "fisher",
                      min_module_size = 5,
@@ -45,7 +45,28 @@ enrichment<-function(Genes=NULL,
     maps<-list(maps)
   }
   else if(!is.list(maps)){
-    stop("maps should be a dataframe or a list of dataframes. Exiting")
+    if(is.character(maps)){
+      map_names<-names(ACSNMineR::ACSN_maps)
+      w<-numeric()
+      for(map_index in maps){
+        w<-c(w,which(grepl(pattern = map_index,x = map_names,ignore.case = TRUE)))
+      }
+      w<- unique(w)
+      if(length(w)>1){
+        maps<-ACSNMineR::ACSN_maps[[w]]
+      }
+      else if(length(w)){
+        maps<-list(ACSNMineR::ACSN_maps[[w]])
+        names(maps)<-map_names[w]
+      }
+      else{
+        stop("The name of the maps you entered are unknown")
+        
+      }
+    }
+    else{
+        stop("maps should be a dataframe or a list of dataframes. Exiting")
+    }
   }
   
   if(universe[1] == "ACSN"){
@@ -834,19 +855,23 @@ format_from_gmt<-function(path = ""){
     max_length<-max(sapply(X=Lines,FUN = function(z){
       z2<-gsub("\t","",z)
       return(nchar(z)-nchar(z2))
-    }))  
-    gmt<-read.csv(path,header = FALSE, 
+    }))+1  ### max gets number of intervals, so number of items is max +1
[…]

### 2016-01-29 R/ACSN_Enrichment.R
Updating doc

@@ -882,13 +882,12 @@ format_from_gmt<-function(path = ""){
 #' 
 #' Outputs the "-log" of a scale
 #' @param base : base for the log, defaut is e
-
 reverselog_trans<-function(base = 10){
   trans<-function(x) -log(x,base)
   inv<-function(x) base^(-x)
   
-  trans_new(paste0("reverslog-",format(base)),trans,inv,
-              log_breaks(base = base),
+  scales::trans_new(paste0("reverslog-",format(base)),trans,inv,
+              scales::log_breaks(base = base),
               domain = c(.Machine$double.xmin,Inf)
             )
   

### 2016-01-29 R/ACSN_Enrichment.R
Updating maps
- maps have been update to latest version of ACSN (and default of gene
6709 corrected to LTA)

@@ -856,7 +856,7 @@ format_from_gmt<-function(path = ""){
       z2<-gsub("\t","",z)
       return(nchar(z)-nchar(z2))
     }))+1  ### max gets number of intervals, so number of items is max +1
-    gmt<-read.table(path,header = FALSE, 
+    gmt<-read.csv(path,header = FALSE, 
                   sep = "\t",fill = TRUE,
                   col.names = paste("V",1:max_length,sep="")
     )

### 2016-01-29 R/ACSN_Enrichment.R
[Updating vignettes]
Version as of today for maps

@@ -881,7 +881,8 @@ format_from_gmt<-function(path = ""){
 #' Scale for barplots and heatmaps
 #' 
 #' Outputs the "-log" of a scale
-#' @param base : base for the log, defaut is e
+#' @param base : base for the log, defaut is 10
+#' @importFrom scales trans_new log_breaks
 reverselog_trans<-function(base = 10){
   trans<-function(x) -log(x,base)
   inv<-function(x) base^(-x)

### 2016-01-29 R/ACSN_Enrichment.R
[vignette issue] Reverselog scale not working

@@ -883,6 +883,7 @@ format_from_gmt<-function(path = ""){
 #' Outputs the "-log" of a scale
 #' @param base : base for the log, defaut is 10
 #' @importFrom scales trans_new log_breaks
+#' @export
 reverselog_trans<-function(base = 10){
   trans<-function(x) -log(x,base)
   inv<-function(x) base^(-x)

### 2016-02-04 jss/article.tex
[skip ci] update table

@@
         Emmanuel Barillot\\Institut Curie\\PSL University\\INSERM U900\\Mines-ParisTech \And 
+        Valentina Boeva\\Institut Curie\\PSL University\\INSERM U900\\Mines-ParisTech \AND 
         Andrei Zinovyev\\Institut Curie\\PSL University\\INSERM U900\\Mines-ParisTech \And
@@
 %% for pretty printing and a nice hypersummary also set:
-\Plainauthor{Paul Deveau, Emmanuel Barillot, Andrei Zinovyev, Eric Bonnet} %% comma-separated
+\Plainauthor{Paul Deveau, Emmanuel Barillot, Valentina Boeva, Andrei Zinovyev, Eric Bonnet} %% comma-separated
 \Plaintitle{Calculating Biological Module Enrichment or Depletion and Visualizing Data on Large-scale Molecular Maps with the R packages ACSNMineR and RNaviCell} %% without formatting
@@
 Thus the information included in ACSN is organized in three hierarchical levels:
-a global map, five individual maps, and a total of 55 functional modules, as of today. Each
+a global map, five individual maps, and several functional modules. Each
 ACSN map covers hundreds of molecular players, biochemical reactions and causal
@@
     \hline
-  ACSN global & 2137 & 55 & 2 &625& 85\\
-  Survival  &1065&46  &1 &434 &54\\
-  Apoptosis & 668&44 & 1& 382& 33\\
-  EMT \& Cell motility &620 &41  &1 &615 &43\\
-  DNA repair &337&48  &1 &69  &19\\
-  Cell cycle &119&45  &1 &27  &8\\
+  ACSN global & 2243 & 63 & 2 &629& 83\\
+  Survival  &1053&5  &208 &431 &328\\
+  Apoptosis & 667&7 & 19& 382& 136\\
+  EMT \& Cell motility &634 &9  &18 &629 &136\\
+  DNA repair &345&21  &3 &171  &45\\
+  Cell cycle &250&25  &2 &130  &19\\
   \hline

### 2016-02-08 R/ACSN_Enrichment.R
[Updates]
- Map name Surival corrected to Survival
- Reverselog transformation now works correctly

@@ -49,11 +49,15 @@ enrichment<-function(Genes=NULL,
       map_names<-names(ACSNMineR::ACSN_maps)
       w<-numeric()
       for(map_index in maps){
-        w<-c(w,which(grepl(pattern = map_index,x = map_names,ignore.case = TRUE)))
+        w<-c(w,which(grepl(pattern = map_index, x = map_names, ignore.case = TRUE)))
       }
       w<- unique(w)
+	#print(w)
       if(length(w)>1){
-        maps<-ACSNMineR::ACSN_maps[[w]]
+        maps<-ACSNMineR::ACSN_maps[w]
+	  mapnames<-names(ACSNMineR::ACSN_maps)[w]
+	
+	  names(maps)<-mapnames
       }
       else if(length(w)){
         maps<-list(ACSNMineR::ACSN_maps[[w]])
@@ -663,7 +667,7 @@ represent_enrichment<-function(enrichment, plot = "heatmap" , scale = "log",
         q<- q + ggplot2::scale_fill_gradient("p-values",low = low , high = high, na.value = na.value, trans = "log10")
       }
       else if(scale == "reverselog"){
-        q<- q + ggplot2::scale_fill_gradient("p-values",low = high , high = low, na.value = na.value, trans = "reverselog")
+        q<- q + ggplot2::scale_fill_gradient("p-values",low = high , high = low, na.value = na.value, trans = reverselog_trans())
       }
       
       else{
@@ -680,7 +684,7 @@ represent_enrichment<-function(enrichment, plot = "heatmap" , scale = "log",
         q<- q + ggplot2::scale_y_continuous(trans = "log10")
       }
       else if(scale == "reverselog"){
-        q<- q + ggplot2::scale_y_continuous(trans = "reverselog")
+        q<- q + ggplot2::scale_y_continuous(trans = reverselog_trans())
       }
     }    
     q<-q+ggplot2::theme_minimal()+ggplot2::theme(axis.text.x = element_text(angle = 90, hjust = 0), axis.ticks = ggplot2::element_blank())
@@ -785,8 +789,10 @@ represent_enrichment<-function(enrichment, plot = "heatmap" , scale = "log",
         q<- q + ggplot2::scale_fill_gradient("p-values",low = low , high = high, na.value = na.value, trans = "log10")
       }
       else if(scale == "reverselog"){
-        q<- q + ggplot2::scale_fill_gradient("p-values",low = high , high = low, na.value = na.value, trans = "reverselog")
-      }
[…]

### 2016-02-08 R/ACSN_Enrichment.R
[Correction] Heatmap with single sample

@@ -667,7 +667,9 @@ represent_enrichment<-function(enrichment, plot = "heatmap" , scale = "log",
         q<- q + ggplot2::scale_fill_gradient("p-values",low = low , high = high, na.value = na.value, trans = "log10")
       }
       else if(scale == "reverselog"){
-        q<- q + ggplot2::scale_fill_gradient("p-values",low = high , high = low, na.value = na.value, trans = reverselog_trans())
+        q<- q + ggplot2::scale_fill_gradient("p-values",low = high , 
+                                             high = low, na.value = na.value, 
+                                             trans = reverselog_trans())
       }
       
       else{
@@ -686,9 +688,9 @@ represent_enrichment<-function(enrichment, plot = "heatmap" , scale = "log",
       else if(scale == "reverselog"){
         q<- q + ggplot2::scale_y_continuous(trans = reverselog_trans())
       }
+      q<-q+ggplot2::theme_minimal()+ggplot2::theme(axis.text.x = element_text(angle = 90, hjust = 0), axis.ticks = ggplot2::element_blank())
+      q<-q+ggplot2::geom_bar(stat = "identity")
     }    
-    q<-q+ggplot2::theme_minimal()+ggplot2::theme(axis.text.x = element_text(angle = 90, hjust = 0), axis.ticks = ggplot2::element_blank())
-    q<-q+ggplot2::geom_bar(stat = "identity")
   }
   else if(is.list(enrichment)){
     sample_names<-names(enrichment)

### 2016-02-08 jss/article.tex
[Updated article]
- Updated figures
- updated default testing on all maps (but not master)
- updated enrichment results in article

@@
              module module_size nb_genes_in_module    genes_in_module 
-V12 Apoptosis:E2F1           5                  4 ATM E2F1 EP300 KAT2B
+V17 APOPTOSIS_ENTRY          10                  4 ATM ATR CHEK2 E2F1          
     universe_size nb_genes_in_universe      p.value	p.value.corrected    test
-V12          2133                   12 1.043478e-08       2.81739e-07 greater
+V17      2241                   12 3.491372e-07	6.400849e-06 greater
 \end{verbatim}
@@
 	\hline
-Apoptosis:E2F1 & 5 &  ATM E2F1 EP300 & 1.04e-08 & greater \\
-               &   &  KAT2B                 &          &         \\
-Apoptosis:G2 M CHECKPOINT & 35  & ATM KAT2B & 2.04e-02 & greater \\
-Apoptosis:WNT CANONICAL & 53 & E2F1 EP300 KAT2B  & 4.98e-03 & greater \\
+CellCycle & 242 & ATM ATR CHEK2   & 5.36e-07 & greater \\
+          &   &  CREBBP KAT2B TFDP1   &          &         \\
+          &   &  E2F1 EP300 HDAC1    &          &         \\
+          &	  & GTF2H1 GTF2H2 GTF2H2B &	& \\
+CellCycle: APOPTOSIS ENTRY & 10 & ATM ATR CHEK2 E2F1 & 3.49e-07 & greater \\
+CellCycle: CYCLINB			& 7  & ATM 				  & 3.98e-02 & greater \\
+
 	\hline
@@
 
-represent_enrichment(enrichment = list(Corrected = results[1:10,], 
-Uncorrected = results_uncorrected[1:10,]),
-                                plot = "heatmap", scale = "log", 
+represent_enrichment(enrichment = list(Corrected = results[1:6,], 
+Uncorrected = results_uncorrected[1:6,]),
+                                plot = "heatmap", scale = "reverselog", 
                                 low = "steelblue" , high ="white", na.value = "grey")
@@
 
- represent_enrichment(enrichment = list(Corrected = results[1:10,], 
-                             Uncorrected = results_uncorrected[1:10,]),
+ represent_enrichment(enrichment = list(Corrected = results[1:6,], 
+                             Uncorrected = results_uncorrected[1:6,]),
                                plot = "heatmap", scale = "reverselog", 

### 2016-02-08 R/ACSN_Enrichment.R

@@ -29,7 +29,11 @@
 #' @importFrom stats fisher.test p.adjust phyper
 #' @importFrom utils read.csv
 enrichment<-function(Genes=NULL,
-                     maps = "Master", 
+                     maps = c("Apoptosis",
+                              "CellCycle",
+                              "DNA_repair",
+                              "EMT_motility",
+                              "Survival" ), 
                      correction_multitest = "BH",
                      statistical_test = "fisher",
                      min_module_size = 5,
@@ -546,7 +550,11 @@ enrichment<-function(Genes=NULL,
 #' @export
 
 multisample_enrichment<-function(Genes_by_sample=NULL,
-                                 maps = ACSNMineR::ACSN_maps, 
+                                 maps = c("Apoptosis",
+                                          "CellCycle",
+                                          "DNA_repair",
+                                          "EMT_motility",
+                                          "Survival" ), 
                                  correction_multitest = "BH",
                                  statistical_test = "fisher",
                                  min_module_size = 5,

### 2016-02-08 jss/scripts/ACSNMineR.R

@@ -9,12 +9,14 @@ print(names(ACSNMineR::ACSN_maps))
 for(map in names(ACSNMineR::ACSN_maps)){
   print(map)
   MAP<-ACSNMineR::ACSN_maps[[map]]
-  print(paste("Number of genes:",length(unique(as.character(MAP[,-(1:2)])))))
-  print(paste("Number of modules:",nrow(MAP)))
-  print(paste("Smallest module size:",min(MAP[,2])))
-  print(paste("Biggest module size:",max(MAP[,2])))
-  print(paste("Mean module size:", round(MAP[,2])))
   
+  print(paste("Number of genes:",sum(unique(as.character(MAP[,-(1:2)]))!="")))
+  MODS<-as.numeric(as.character(MAP[,2]))
+  print(paste("Number of modules:",length(MODS)))
+  print(paste("Smallest module size:",min(MODS)))
+  print(paste("Biggest module size:",max(MODS)))
+  print(paste("Mean module size:", round(mean(MODS))))
+  print("**************")
 }
 
 ### Print genes test
@@ -35,16 +37,16 @@ print(head(results_uncorrected,3))
 
 
 ### Plot heatmap
-heatmap <- represent_enrichment(enrichment = list(Corrected = results[1:10,], 
-                                       Uncorrected = results_uncorrected[1:10,]),
-                                plot = "heatmap", scale = "log", 
+heatmap <- represent_enrichment(enrichment = list(Corrected = results[1:6,], 
+                                       Uncorrected = results_uncorrected[1:6,]),
+                                plot = "heatmap", scale = "reverselog", 
                                 low = "steelblue" , high ="white", na.value = "grey")
 
 print(heatmap)
 
 ### Plot bar
-baplot <- represent_enrichment(enrichment = list(Corrected = results[1:10,], 
-                                                 Uncorrected = results_uncorrected[1:10,]),
+baplot <- represent_enrichment(enrichment = list(Corrected = results[1:6,], 
+                                                 Uncorrected = results_uncorrected[1:6,]),
                                plot = "heatmap", scale = "reverselog", 
                                nrow = 1)
 print(barplot)

### 2016-02-09 jss/article.tex
[skip ci] update table

@@
         Emmanuel Barillot\\Institut Curie\\PSL University\\INSERM U900\\Mines-ParisTech \And 
-        Valentina Boeva\\Institut Curie\\PSL University\\INSERM U900\\Mines-ParisTech \AND 
+        Valentina Boeva\\Institut Curie\\PSL University\\INSERM U900\\Mines-ParisTech\\Institut Cochin\\INSERM U1016\\CNRS UMR 8104\\Paris Descartes University UMR-S1016 \AND 
         Andrei Zinovyev\\Institut Curie\\PSL University\\INSERM U900\\Mines-ParisTech \And
@@
 then only the map name is shown, without any module or column character. For
-instance, the first line of the results object below concern the E2F1 module of
-the Apoptosis map.
+instance, the third line of the results object below concern the E2F1 module of
+the CellCycle map.
 
@@
 > dim(results)
-[1] 36  9
-> results[1,]
-             module module_size nb_genes_in_module    genes_in_module 
-V17 APOPTOSIS_ENTRY          10                  4 ATM ATR CHEK2 E2F1          
+[1] 8  9
+> results[3,]
+            module module_size nb_genes_in_module
+V161 CellCycle:E2F1          19                 12
+                                                             genes_in_module
+V161 ATM ATR CHEK2 CREBBP TFDP1 E2F1 EP300 HDAC1 KAT2B GTF2H1 GTF2H2 GTF2H2B
      universe_size nb_genes_in_universe      p.value p.value.corrected    test
-V17      2241                   12 3.491372e-07	6.400849e-06 greater
+V161          2237                   12 3.735018e-21      2.353061e-19 greater
 \end{verbatim}
@@
 between ER positive and ER negative samples, and calculates the enrichment in
-ACSN maps from this list of genes. As seen in Table \ref{tab:table_mainz}, there
-are four maps (Apoptosis, DNA repair, ACSN master and Survival) and seven
-modules enriched. Most of the modules are in fact linked to different phases of
-the cell cycle.
+ACSN maps from this list of genes. As seen in Table \ref{tab:table_mainz},
+there is one map (Cell Cycle) and six modules (belonging to the Cell Cycle, DNA
+repair and EMT motility maps) enriched.
 
@@
 \hline
-Apoptosis:HEDGEHOG & 31 & 21 & 0.0016 & 0.042 \\
-Apoptosis:AKT\_MTOR &  79 & 47 & 0.00037 & 0.038 \\
-DNA\_repair & 335 & 163 & 0.0012 & 0.038 \\
-DNA\_repair:G1\_CC\_PHASE &  25 & 18 & 0.0011 & 0.038 \\
-ACSN\_master:AKT\_MTOR & 92 & 53 & 0.0005 & 0.038 \\
-ACSN\_master:G1\_CC\_PHASE & 25 & 18 & 0.0011 & 0.038 \\
-Survival:S\_PHASE\_CHECKPOINT & 8 & 8 & 0.0006 & 0.038 \\
+CellCycle & 242 & 20 & 1.70e-03 & 0.021 \\
+CellCycle:E2F1\_TARGETS & 130 & 14 & 6.26e-04 & 0.009 \\
+CellCycle:E2F2\_TARGETS & 35 & 8 & 4.70e-05 & 0.009 \\
+CellCycle:E2F3\_TARGETS & 51 & 11 & 3.82e-06 & 0.0002 \\
+CellCycle:E2F4\_TARGETS & 100 & 15 & 9.99e-06 & 0.0003 \\
+DNA\_repair:SPINDLE\_CHECKPOINT & 28 & 5 & 3.82e-03 & 0.04 \\
+EMT\_motility:ECM & 147 & 13 & 5.09e-03 & 0.045 \\
 \hline

### 2016-02-10 jss/article.tex
[skip ci] update

@@
 with the ACSNMineR function \code{enrichment}. The results are displayed in
-table \ref{tab:table_gbm}. There are 16 modules significantly enriched in the
-Cell Cycle, DNA repair, EMT motility and Survival. Cell matrix adhesions and ECM
+table \ref{tab:table_gbm}. There are 6 modules significantly enriched in the
+DNA repair and EMT motility maps. Cell matrix adhesions and ECM
 (extra cellular matrix), part of the EMT motility map, are the modules with
-highest significance (corrected p-values of 2.55e-11 and 2.34e-9 respectively).
+highest significance. The EMT motility map is significantly enriched at the
+global map level (second line in the table).
 
@@
 \hline
-CellCycle:G2\_M\_CHECKPOINT & 14 & 8 & 3.92e-02\\
-CellCycle:S\_PHASE\_CHECKPOINT & 6 & 5 & 3.92e-02\\
-DNA\_repair:S\_PHASE\_CHECKPOINT & 45 & 19 & 1.58e-02\\
-EMT\_motility & 621 & 176 & 1.06e-03\\
-EMT\_motility:CELL\_MATRIX\_ADHESIONS & 72 & 44 & 2.55e-11\\
-EMT\_motility:ECM & 137 & 64 & 2.34e-09\\
-EMT\_motility:EMT\_REGULATORS & 615 & 173 & 1.49e-03\\
-EMT\_motility:CYTOSKELETON\_POLARITY & 154 & 47 & 3.92e-02\\
-EMT\_motility:DESMOSOMES & 29 & 15 & 4.62e-03\\
-ACSN\_master:CELL\_MATRIX\_ADHESIONS & 72 & 44 & 2.55e-11\\
-ACSN\_master:DESMOSOMES & 29 & 15 & 4.62e-03\\
-ACSN\_master:S\_PHASE\_CHECKPOINT & 46 & 20 & 8.23e-03\\
-ACSN\_master:ECM & 138 & 64 & 2.56e-09\\
-ACSN\_master:EMT\_REGULATORS & 625 & 177 & 1.06e-03\\
-ACSN\_master:CYTOSKELETON\_POLARITY & 154 & 47 & 3.92e-02\\
-Survival:ECM & 23 & 11 & 3.92e-02\\
+DNA\_repair:S\_PHASE\_CHECKPOINT &   45 &      19 &     0.008\\
+EMT\_motility  &  635  &   181  &   0.0002\\
+EMT\_motility:CELL\_MATRIX\_ADHESIONS   &   73  &    45  &    3.73e-12\\
+EMT\_motility:CYTOSKELETON\_POLARITY   &   154 &    47  &    0.022\\
+EMT\_motility:DESMOSOMES & 29  &    15  &    0.002\\
+EMT\_motility:ECM    &    147  &   69   &   9.77e-11\\
+EMT\_motility:EMT\_REGULATORS  &   629   &  178  &   0.0002\\
 \hline
@@
 the map, with a high density of low and high frequency mutated genes in the EMT
-motility , survival, cell cycle and DNA repair regions (maps) of the global ACSN
-map.
+motility and DNA repair regions (maps) of the global ACSN map. Although they
+are not statistically significant, quite high densities can also be seen in
+other regions of the map.
 

### 2016-02-11 jss/article.tex
[skip ci] update

@@
 Reactome, MINT databases, but also in more specialized repositories such as the
-recently released Atlas of Cancer Signaling Networks (ACSN). All these open and
+recently released Atlas of Cancer Signaling Network (ACSN). All these open and
 popular biological databases facilitate a large spectrum of analyses, thereby
@@
 support for ACSN molecular pathways, but can use any molecular pathway database
-encoded as a GMT file, such as any of the gene signatures available in the
+encoded as a GMT file, such as sets of gene signatures available in the
 Molecular Signatures Database (MSigDB). We also present the R package
@@
 genes that are differentially expressed between two conditions, such as normal
-vs disease) and then try to see if known biological pathways are enriched with
-this list of genes.
+vs disease) and then evaluate if known biological pathways have significant
+overlap with this list of genes.
 
-We have recently released the Atlas of Cancer Signalling Networks (ACSN), a
+We have recently released the Atlas of Cancer Signalling Network (ACSN), a
 web-based database which describes signaling and regulatory molecular processes
@@
 represents a large-scale biochemical reaction network of 4,826 reactions
-involving 2,371 proteins, and is continuously updated and expanded.  We have
+involving 2,371 proteins (as of today), and is continuously updated and expanded.  We have
 included the three hierarchical levels in the \pkg{ACSNMineR} package, in order
@@
  \centering
-  \caption{ACSN maps included in the \pkg{ACSNMineR} package. Map: map name, Total: total
-  number of gene symbols (HUGO), Nb mod.: number of modules, Min: mimimum
-  number of gene symbols in the modules, Max: maximum number of gene symbols in
-  the modules, Mean: average number of gene sybols per module. N.B.: one gene
-  symbol may be present in several modules of the map.}
-  \label{tab:table1}
+  \caption{ACSN maps included in the \pkg{ACSNMineR} package. Map: map name,
+  Total: total number of gene symbols (HUGO) used to construct the map, Nb
+  mod.: number of modules, Min: mimimum number of gene symbols in the modules,
+  Max: maximum number of gene symbols in the modules, Mean: average number of
+  gene sybols per module. N.B.: one gene symbol may be present in several
+  modules of the map.} \label{tab:table1}
   \begin{tabular}{l|c|c|c|c|c}
@@
 uses the ACSN maps previously described.  The correction for multiple testing
-is set by default to false discovery rate (FDR), which is equivalent to
-Benjamini \& Hochberg correction, but can be changed to any of the usual
-correction methods (Bonferroni, Holm, Hochberg, Holm, or Benjamini \& Yekutieli
-\citep{Benjamini2003FDR}), or even disabled .  The minimal module size
-represents the smallest size value of a module that will be used to compute
-enrichment or depletion. This is meant to remove results of low significance
-for module of small size.  The universe in which the computation is made by
-default is defined by all the gene symbols contained in the maps. All the genes
-that were given as input and that are not present on the maps will be
-discarded. To keep all genes, the user can change the universe to \code{HUGO},
-and in that case, the complete list of HUGO gene symbols will be used as the
-reference ($>$ 39,000 genes). The threshold corresponds to the maximal value of the
-corrected p-value (unless the user chose not to correct for multiple testing) that
-will be displayed in the result table.
+is set by default to use the method of Benjamini \& Hochberg, but can be
+changed to any of the usual correction methods (Bonferroni, Holm, Hochberg,
+Holm, or Benjamini \& Yekutieli \citep{Benjamini2003FDR}), or even disabled .
+The minimal module size represents the smallest size value of a module that
+will be used to compute enrichment or depletion. This is meant to remove
+results of low significance for module of small size.  The universe in which
+the computation is made by default is defined by all the gene symbols contained
+in the maps. All the genes that were given as input and that are not present on
+the maps will be discarded. To keep all genes, the user can change the universe
+to \code{HUGO}, and in that case, the complete list of HUGO gene symbols will
+be used as the reference ($>$ 39,000 genes). The threshold corresponds to the
+maximal value of the corrected p-value (unless the user chose not to correct
+for multiple testing) that will be displayed in the result table.
 
@@
 	\centering
-	\caption{Representation of the enriched modules (first ten rows), with either Bonferroni correction or no correction.
-Grey tiles means that the data is not available for this module in this sample.
-P-values of low significance are in white, whereas p-values of high significance are represent in blue.
-Samples are in abscissa and modules are on the Y axis  }
+    \caption{Representation of the enriched modules (first ten rows), with
+    either Bonferroni correction or no correction. Grey tiles means that the
+    data is not available for this module in this sample. P-values of low
+    significance are in white, whereas p-values of high significance are
+    represented in blue.}
     \includegraphics{figures/comparison_corrected_unc.pdf}
@@
 pathways. The package includes the six large-scale molecular maps and 55
-functional modules of the Atlas of Cancer Signaling Networks (ACSN) . Enrichment can be
+functional modules of the Atlas of Cancer Signaling Network (ACSN) . Enrichment can be
 calculated for those maps and modules with several options to play with, but can
@@
 
-We also describe in this work the \pkg{RNaviCell} package, a R package highly
-complementary to \pkg{ACSNMineR}. This package is dedicated to create web-based
+We also describe in this work the \pkg{RNaviCell} package, a R package convenient to use with \pkg{ACSNMineR}. This package is dedicated to create web-based
 and interactive data visualization on ACSN maps. Users can use this tools to
@@
 \code{PFAMHyperGParams}). However, its usage might not be as straightforward as
-\pkg{ACSNMineR}, and it does not seems possible to test user-defined biological
+\pkg{ACSNMineR}, and it does not seem possible to test user-defined biological
 pathways. Furthermore, other authors have pointed out that the KEGG database

### 2016-02-15 jss/article.tex
[skip ci] update

@@
 relationships, occurring at the molecular level in living cells
-\citep{adriaens2008public}.  A large body of knowledge on cellular biochemistry
-is organized in publicly available repositories such as the KEGG database
-\citep{kanehisa2011kegg}, Reactome \citep{croft2014reactome} and MINT
-\citep{zanzoni2002mint}. All these biological databases facilitate a large
-spectrum of analyses, improving our understanding of cellular systems. For
-instance, it is a very common practice to cross the output of high-throughput
-experiments, such as mRNA or protein expression levels, with curated biological
-pathways in order to visualize the changes, analyze their impact on a network and
-formulate new hypotheses about biological processes. Many biologists and
-computational biologists establish list of genes of interest (e.g. a list of
-genes that are differentially expressed between two conditions, such as normal
-vs disease) and then evaluate if known biological pathways have significant
-overlap with this list of genes.
+\citep{adriaens2008public, barillot2012computational}.  A large body of
+knowledge on cellular biochemistry is organized in publicly available
+repositories such as the KEGG database \citep{kanehisa2011kegg}, Reactome
+\citep{croft2014reactome} and MINT \citep{zanzoni2002mint}. All these
+biological databases facilitate a large spectrum of analyses, improving our
+understanding of cellular systems. For instance, it is a very common practice
+to cross the output of high-throughput experiments, such as mRNA or protein
+expression levels, with curated biological pathways in order to visualize the
+changes, analyze their impact on a network and formulate new hypotheses about
+biological processes. Many biologists and computational biologists establish
+list of genes of interest (e.g. a list of genes that are differentially
+expressed between two conditions, such as normal vs disease) and then evaluate
+if known biological pathways have significant overlap with this list of genes.
 
-We have recently released the Atlas of Cancer Signalling Network (ACSN), a
+We have recently released the Atlas of Cancer Signaling Network (ACSN), a
 web-based database which describes signaling and regulatory molecular processes
@@
 \subsection{ACSNMineR}
-Currently, ACSN maps covers signaling pathways involved in DNA repair, cell
+Currently, ACSN maps cover signaling pathways involved in DNA repair, cell
 cycle, cell survival, cell death, epithelial-to-mesenchymal transition (EMT) and
@@
 > names(ACSN_maps)
-[1] "Apoptosis"    "CellCycle"    "DNA_repair"   "EMT_motility" "ACSN_master" 
+[1] "Apoptosis"    "CellCycle"    "DNA_repair"   "EMT_motility" "Master" 
 [6] "Survival"    

### 2016-03-09 jss/ACSNMineR.tex
Changed JSS article to RJournal
- updated title (removed R reference)
- updated abstract (150 words)
- Edited style and format, figures, etc.
[skip CI]

--- 2016-02-15 jss/article.tex
@@
-\documentclass[article]{jss}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% declarations for jss.cls %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%% almost as usual
-\author{Paul Deveau\\Institut Curie\\PSL University\\Univ. Paris-Sud\\INSERM U830\\INSERM U900\\Mines-ParisTech \And 
-        Emmanuel Barillot\\Institut Curie\\PSL University\\INSERM U900\\Mines-ParisTech \And 
-        Valentina Boeva\\Institut Curie\\PSL University\\INSERM U900\\Mines-ParisTech\\Institut Cochin\\INSERM U1016\\CNRS UMR 8104\\Paris Descartes University UMR-S1016 \AND 
-        Andrei Zinovyev\\Institut Curie\\PSL University\\INSERM U900\\Mines-ParisTech \And
-        Eric Bonnet\\Institut Curie\\PSL University\\INSERM U900\\Mines-ParisTech}
-\title{Calculating Biological Module Enrichment or Depletion and Visualizing Data on Large-scale Molecular Maps with the R packages \pkg{ACSNMineR} and \pkg{RNaviCell}}
+%% authors
+%\author{Paul Deveau\\Institut Curie\\PSL University\\Univ. Paris-Sud\\INSERM U830\\INSERM U900\\Mines-ParisTech \And 
+%        Emmanuel Barillot\\Institut Curie\\PSL University\\INSERM U900\\Mines-ParisTech \And 
+%        Valentina Boeva\\Institut Curie\\PSL University\\INSERM U900\\Mines-ParisTech\\Institut Cochin\\INSERM U1016\\CNRS UMR 8104\\Paris Descartes University UMR-S1016 \AND 
+%        Andrei Zinovyev\\Institut Curie\\PSL University\\INSERM U900\\Mines-ParisTech \And
+%        Eric Bonnet\\Institut Curie\\PSL University\\INSERM U900\\Mines-ParisTech}
+%\title{Calculating Biological Module Enrichment or Depletion and Visualizing Data on Large-scale Molecular Maps with the \R packages \pkg{ACSNMineR} and \pkg{RNaviCell}}
 
 %% for pretty printing and a nice hypersummary also set:
-\Plainauthor{Paul Deveau, Emmanuel Barillot, Valentina Boeva, Andrei Zinovyev, Eric Bonnet} %% comma-separated
-\Plaintitle{Calculating Biological Module Enrichment or Depletion and Visualizing Data on Large-scale Molecular Maps with the R packages ACSNMineR and RNaviCell} %% without formatting
-\Shorttitle{Module Enrichment and Data Visualization with ACSNMineR and RNaviCell} %% a short title (if necessary)
+\author{by Paul Deveau, Emmanuel Barillot, Valentina Boeva, Andrei Zinovyev and Eric Bonnet} %% comma-separated
+\title{Calculating Biological Module Enrichment or Depletion and Visualizing Data on Large-scale Molecular Maps with ACSNMineR and RNaviCell packages} %% without formatting
+%\Shorttitle{Module Enrichment and Data Visualization with ACSNMineR and RNaviCell} %% a short title (if necessary)
+\maketitle
+%% publication information
 
 %% an abstract and keywords
-\Abstract{
+\abstract{
 Biological pathways or modules represent sets of interactions or functional 
-relationships occurring at the molecular level in living cells. A large body of
-knowledge on pathways is organized in public databases such as the KEGG,
-Reactome, MINT databases, but also in more specialized repositories such as the
-recently released Atlas of Cancer Signaling Network (ACSN). All these open and
-popular biological databases facilitate a large spectrum of analyses, thereby
-improving our understanding of cellular systems. Here we describe the
-\pkg{ACSNMineR}, a novel R package for calculation of enrichment or depletion
-of list of genes of interest in biological pathways. The package has built-in
-support for ACSN molecular pathways, but can use any molecular pathway database
-encoded as a GMT file, such as sets of gene signatures available in the
-Molecular Signatures Database (MSigDB). We also present the R package
-\pkg{RNaviCell}, that can be used in conjunction with \pkg{ACSNMineR} to
+relationships occurring at the molecular level in living cells. A large body 
+of knowledge on pathways is organized in public databases such as the KEGG,
+Reactome, or in more specialized repositories, the Atlas of Cancer Signaling 
+Network (ACSN) being an example. All these open and popular biological databases
+facilitate analyses, improving our understanding of cellular systems. Here we describe
+\CRANpkg{ACSNMineR} for calculation of enrichment or depletion
+of gene-lists of interest in biological pathways. It has integrated ACSN molecular
+pathways, but can use any molecular pathway encoded as a GMT file, for instance sets 
+of genes available in the Molecular Signatures Database (MSigDB). We also present 
+\CRANpkg{RNaviCell},that can be used in conjunction with \pkg{ACSNMineR} to
 visualize different data types on web-based, interactive ACSN maps. We
-illustrate the functionalities of the two packages with real-world biological
+illustrate the functionalities of the two packages with biological
 data taken from large-scale cancer datasets.
@@
 }
-\Keywords{biological module, biological pathway, enrichment, depletion, cancer, maps, molecular networks, molecular pathways, \proglang{R}}
-\Plainkeywords{biological module, biological pathway, enrichment, depletion, cancer, maps, molecular networks, molecular pathways, R} %% without formatting
-%% at least one keyword must be supplied
-
-%% publication information
-%% NOTE: Typically, this can be left commented and will be filled out by the technical editor
-%% \Volume{50}
-%% \Issue{9}
-%% \Month{June}
-%% \Year{2012}
-%% \Submitdate{2012-06-04}
-%% \Acceptdate{2012-06-04}
-
-%% The address of (at least) one author should be given
-%% in the following format:
-\Address{
-  Eric Bonnet\\
-  Computational Systems Biology of Cancer\\
-  Institut Curie\\
-  26, rue d'Ulm\\
-  75248 Paris, France\\
-  E-mail: \email{eric.bonnet@curie.fr}\\
-}
-%% It is also possible to add a telephone and fax number
-%% before the e-mail in the following format:
-%% Telephone: +43/512/507-7103
-%% Fax: +43/512/507-2851
-
-%% for those who use Sweave please include the following line (with % symbols):
-%% need no \usepackage{Sweave.sty}
-
-%% end of declarations %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{document}
+%\Keywords{biological module, biological pathway, enrichment, depletion, cancer, maps, molecular networks, molecular pathways, \proglang{R}}
+%\Plainkeywords{biological module, biological pathway, enrichment, depletion, cancer, maps, molecular networks, molecular pathways, R} %% without formatting
 
@@
 
-In this article, we describe two software packages related to ACSN analysis and
-data visualization for the popular R statistical enrironment \citep{mainRref,
-vance2009data}. The package \pkg{ACSNMineR} is designed for the calculation of
+In this article, we describe two packages related to ACSN analysis and
+data visualization. The package \CRANpkg{ACSNMineR} is designed for the calculation of
 gene enrichment and depletion in ACSN maps (or any user-defined gene set via
-the import function), while \pkg{RNaviCell} is dedicated to data visualization
+the import function), while \CRANpkg{RNaviCell} is dedicated to data visualization
 on ACSN maps. Both packages are available on the Comprehensive R Archive
@@
 cycle, cell survival, cell death, epithelial-to-mesenchymal transition (EMT) and
-cell motility. Each of these large-scale molecular map is decomposed in a number
+cell motility. Each of these large-scale molecular maps is decomposed in a number
 of functional modules. The maps themselves are merged into a global ACSN map.
@@
   modules of the map.} \label{tab:table1}
-  \begin{tabular}{l|c|c|c|c|c}
-    \hline
+  \begin{tabular}{l c c c c c}
+    \toprule
     Map & Total & Nb mod. & Min & Max & Mean\\
-    \hline
+    \midrule
   ACSN global & 2243 & 63 & 2 &629& 83\\
@@
   Cell cycle &250&25  &2 &130  &19\\
-  \hline
+  \bottomrule
 
@@
 
-\begin{verbatim}
+\begin{example}
 > length(ACSN_maps)
@@
 [6] "Survival"    
-\end{verbatim}
+\end{example}
 
@@
 ACSN maps and modules. We have included a small list of 12 Cell Cycle related
-genes in the package, named \code{genes_test} that can be used to test the main
+genes in the package, named \code{genes\_test} that can be used to test the main
 enrichment function and to get familiar with its different options.
 
-\begin{verbatim}
+\begin{example}
 > genes_test
@@
  [8] "HDAC1"   "KAT2B"   "GTF2H1"  "GTF2H2"  "GTF2H2B"
-\end{verbatim}
+\end{example}
 
@@
 
-\begin{verbatim}
+\begin{example}
 > library(ACSNMineR)
@@
 V161          2237                   12 3.735018e-21      2.353061e-19 greater
-\end{verbatim}
+\end{example}
 
@@
 other arguments have default values.  The \code{maps} argument can either be a
-dataframe imported from a gmt file with the \code{format_from_gmt} function or
+dataframe imported from a gmt file with the \code{format\_from\_gmt} function or
 a list of dataframes generated by the same procedure. By default, the function
@@
 differences between two cancer types or two cell lines.  To facilitate such
-comparisons, \pkg{ACSNMineR} provides a \code{multisample_enrichment} function.
+comparisons, \pkg{ACSNMineR} provides a \code{multisample\_enrichment} function.
 It relies on the \code{enrichment} function but takes a list of character
@@
 name of the sample for further analysis.  Most of the arguments given to
-\code{multisample_enrichment} are the same as the ones passed to
-\code{enrichment}. However, the \code{cohort_threshold} is designed to filter
+\code{multisample\_enrichment} are the same as the ones passed to
+\code{enrichment}. However, the \code{cohort\_threshold} is designed to filter
 out modules which would not pass the significance threshold in all samples.
@@
 Finally, to facilitate visualization of results, \pkg{ACSNMineR} integrates a
-representation function based on \pkg{ggplot2} syntax \citep{ggplot2}. It
+representation function based on \CRANpkg{ggplot2} syntax \citep{ggplot2}. It
 allows representation of results from \code{enrichment} or
-\code{multisample_enrichment} with a limited number of parameters. Two types of
+\code{multisample\_enrichment} with a limited number of parameters. Two types of
 display are available: heat-map tiles or bars. For multiple samples using a
@@
 
-We previously computed the p-value of the \code{genes_test} list with default
+We previously computed the p-value of the \code{genes\_test} list with default
 parameters. The number of modules which have a p-value below 0.05 was 36, that
@@
 
-\begin{verbatim}
+\begin{example}
 enrichment(genes_test,correction_multitest = FALSE)
-\end{verbatim}
+\end{example}
 
-\begin{table}[h!]
+\begin{widetable}[h!]
   \centering
@@
 
-  \begin{tabular}{l|c|c|c|c}
-	\hline
+  \begin{tabular}{l c c c c}
+	\toprule
 Module & Mod. size & Genes in module & p-value & Test\\
-	\hline
+	\midrule
 CellCycle & 242 & ATM ATR CHEK2   & 5.36e-07 & greater \\
@@
 
-	\hline
+	\bottomrule
 
 	\end{tabular}
-\end{table}
+\end{widetable}
 
-We can now plot the first ten rows of the results obtained for corrected and
+We can now plot the first six rows of the results obtained for corrected and
 uncorrected fisher test with heatmap format (Figure~\ref{fig:heatm}) or barplot
@@
 
-\begin{verbatim}
+\begin{example}
 # heatmap
@@
                                nrow = 1)
-\end{verbatim}
+\end{example}
 
-\begin{figure}
+\begin{figure}[h!]
 	\centering
-    \caption{Representation of the enriched modules (first ten rows), with
+    \caption{Representation of the enriched modules (first six rows), with
     either Bonferroni correction or no correction. Grey tiles means that the
@@
     represented in blue.}
-    \includegraphics{figures/comparison_corrected_unc.pdf}
+    \includegraphics[width=0.8\textwidth]{figures/comparison_corrected_unc.pdf}
     \label{fig:heatm}
@@
 
-\begin{figure}
-	\caption{Representation of the enriched modules (first ten rows), with either Bonferroni correction (left) or no correction (right).
+\begin{figure}[h!]
+	\centering
+	\caption{Representation of the enriched modules (first six rows), with either Bonferroni correction (left) or no correction (right).
 The modules are on the X axis and the p-values are on the Y axis.
 }
-	\includegraphics{figures/comparison_corrected_unc_bars.pdf}
+	\includegraphics[width=\textwidth]{figures/comparison_corrected_unc_bars.pdf}
 	\label{fig:barp}
@@
 
-\begin{verbatim}
+\begin{example}
 # a short RNaviCell script example
@@
 
-\end{verbatim}
+\end{example}
 
@@
 
-\begin{verbatim}
+\begin{example}
 # load all necessary packages
@@
 [1] 7 9
-\end{verbatim}
+\end{example}
 
@@
 
-\begin{tabular}{l|c|c|c|c}
-\hline
+\begin{tabular}{l c c c c}
+\toprule
 Module & Mod. size & Nb genes & pval & Cor. pval\\
-\hline
+\midrule
 CellCycle & 242 & 20 & 1.70e-03 & 0.021 \\
@@
 EMT\_motility:ECM & 147 & 13 & 5.09e-03 & 0.045 \\
-\hline
+\bottomrule
 \end{tabular}
@@
 
-\begin{verbatim}
+\begin{example}
 # Import MSigDB canonical pathways and calculate enrichment on this database
@@
 
-\begin{tabular}{l|c|c|c}
-\hline
+\begin{tabular}{l c c c}
+\toprule
 Pathway & Mod. size & Nb genes & Cor. pval\\
-\hline
+\midrule
 KEGG\_CELL\_CYCLE & 128 & 76 & 1.98e-07\\
@@
 REACTOME\_S\_PHASE & 109 & 58 & 4.40e-04\\
-\hline
+\bottomrule
 \end{tabular}
@@
 
-\begin{verbatim}
+\begin{example}
 # Select ER negative samples and calculate mean expression values
@@
 navicell$heatmapEditorApply()
-\end{verbatim}
+\end{example}
 
@@
 
-\begin{tabular}{l|c|c|c}
-\hline
+\begin{tabular}{l c c c}
+\toprule
 module & Mod. size & Nb genes & Cor. pval\\ 
-\hline
+\midrule
 DNA\_repair:S\_PHASE\_CHECKPOINT &   45 &      19 &     0.008\\
@@
 EMT\_motility:EMT\_REGULATORS  &   629   &  178  &   0.0002\\
-\hline
+\bottomrule
 \end{tabular}
@@
 
-\begin{verbatim}
+\begin{example}
 library(RNaviCell)
@@
 navicell$unorderedConfigApply("GBM", "size")
-\end{verbatim}
+\end{example}
 
@@
 Of course, \pkg{ACSNMineR} is not the only R package for enrichment
-calculations. For instance, \pkg{GOstats} \citep{falcon2007using} is probably
+calculations. For instance, \BIOpkg{GOstats} \citep{falcon2007using} is probably
 one of the first packages that was created to calculate enrichment for Gene
@@
 pathways. Furthermore, other authors have pointed out that the KEGG database
-used by this package has not been updated since 2012. \pkg{clusterProfiler} is a
+used by this package has not been updated since 2012. \BIOpkg{clusterProfiler} is a
 recent R package released for enrichment analysis of Gene Ontology and KEGG with
@@
 
-%\bibliographystyle{jss}
-\bibliography{biblio}
+\bibliography{ACSNMineR}
 
-\end{document}
+\address{
+  Paul Deveau\\
+  Computational Systems Biology of Cancer, Institut Curie\\
+  26, rue d'Ulm 75248 Paris\\
+  France\\
+}
+\email{paul.deveau@curie.fr}
+
+\address{
+  Emmanuel Barillot\\
+  Computational Systems Biology of Cancer, Institut Curie\\
+  26, rue d'Ulm 75248 Paris\\
+  France\\
+}
+\email{emmanuel.barillot@curie.fr}
+
+\address{
+  Valentina Boeva\\
+  Computational Systems Biology of Cancer, Institut Curie\\
+  26, rue d'Ulm 75248 Paris\\
+  France\\
+}
+\email{valentina.boeva@curie.fr}
+
+\address{
+  Andrei Zinovyev\\
+  Computational Systems Biology of Cancer, Institut Curie\\
+  26, rue d'Ulm 75248 Paris\\
+  France\\
+}
+\email{andrei.Zinovyev@curie.fr}
+
+\address{
+  Eric Bonnet\\
+  Computational Systems Biology of Cancer, Institut Curie\\
+  26, rue d'Ulm 75248 Paris\\
+  France\\
+}
+\email{eric.bonnet@curie.fr}

### 2016-03-14 jss/ACSNMineR.tex
Edited abstract
[skip ci]

@@
 \abstract{
-Biological pathways or modules represent sets of interactions or functional 
-relationships occurring at the molecular level in living cells. A large body 
-of knowledge on pathways is organized in public databases such as the KEGG,
-Reactome, or in more specialized repositories, the Atlas of Cancer Signaling 
-Network (ACSN) being an example. All these open and popular biological databases
-facilitate analyses, improving our understanding of cellular systems. Here we describe
-\CRANpkg{ACSNMineR} for calculation of enrichment or depletion
-of gene-lists of interest in biological pathways. It has integrated ACSN molecular
-pathways, but can use any molecular pathway encoded as a GMT file, for instance sets 
-of genes available in the Molecular Signatures Database (MSigDB). We also present 
-\CRANpkg{RNaviCell},that can be used in conjunction with \pkg{ACSNMineR} to
-visualize different data types on web-based, interactive ACSN maps. We
-illustrate the functionalities of the two packages with biological
-data taken from large-scale cancer datasets.
+Biological pathways or modules represent sets of interactions or functional relationships occurring at the molecular level in living cells.
+A large body of knowledge on pathways is organized in public databases such as the KEGG, Reactome, or in more specialized repositories, the Atlas of Cancer Signaling Network (ACSN) being an example.
+All these open biological databases facilitate analyses, improving our understanding of cellular systems.
+We hereby describe \CRANpkg{ACSNMineR} for calculation of enrichment or depletion of lists of genes of interest in biological pathways.
+\pkg{ACSNMineR} integrates ACSN molecular pathways, but can use any molecular pathway encoded as a GMT file, for instance sets of genes available in the Molecular Signatures Database (MSigDB).
+We also present \CRANpkg{RNaviCell}, that can be used in conjunction with \pkg{ACSNMineR} to visualize different data types on web-based, interactive ACSN maps.
+We illustrate the functionalities of the two packages with biological data taken from large-scale cancer datasets.
 

### 2016-05-16 jss/Deveau.tex
updated publication
[skip ci]

--- 2016-03-14 jss/ACSNMineR.tex
@@
 \author{by Paul Deveau, Emmanuel Barillot, Valentina Boeva, Andrei Zinovyev and Eric Bonnet} %% comma-separated
-\title{Calculating Biological Module Enrichment or Depletion and Visualizing Data on Large-scale Molecular Maps with ACSNMineR and RNaviCell packages} %% without formatting
+\title{Calculating Biological Module Enrichment or Depletion and Visualizing Data on Large-scale Molecular Maps with \pkg{ACSNMineR} and \pkg{RNaviCell} packages} %% without formatting
 %\Shorttitle{Module Enrichment and Data Visualization with ACSNMineR and RNaviCell} %% a short title (if necessary)

### 2016-05-16 R/ACSN_Enrichment.R

@@ -905,7 +905,7 @@ reverselog_trans<-function(base = 10){
   trans<-function(x) -log(x,base)
   inv<-function(x) base^(-x)
   
-  scales::trans_new(name = "reverslog",transform = trans,inverse = inv,
+  scales::trans_new(name = "reverselog",transform = trans,inverse = inv,
               breaks = scales::log_breaks(base = base),
               domain = c(10^(-60),Inf)
             )

### 2016-08-23 jss/Deveau.tex
[skip ci] update eric affiliation
[skip ci] update summary and perspectives

@@
 \section[Summary and perspectives]{Summary and perspectives}
+
 In this work, we presented the R package \pkg{ACSNMineR}, a novel package for
@@
 pathways. The package includes the six large-scale molecular maps and 55
-functional modules of the Atlas of Cancer Signaling Network (ACSN) . Enrichment can be
-calculated for those maps and modules with several options to play with, but can
-also be calculated for other databases of molecular pathways, that can be
-imported from GMT formated files.
+functional modules of the Atlas of Cancer Signaling Network (ACSN) . Enrichment
+can be calculated for those maps and modules with several options to play with,
+but can also be calculated for other databases of molecular pathways, that can
+be imported from GMT formated files.
 
-We also describe in this work the \pkg{RNaviCell} package, a R package convenient to use with \pkg{ACSNMineR}. This package is dedicated to create web-based
-and interactive data visualization on ACSN maps. Users can use this tools to
-represent genes of interest that have been shown to be related to the maps by
-calculating enrichment with the  \pkg{ACSNMineR}.  Creating maps with the
-graphical user interface of the ACSN website can be a tedious task if the user
-has multiple samples or gene lists, and wants to compare their representations
-on ACSN maps. The \pkg{RNaviCell} package can be used to automate the process of
-creating the graphical representations automatically.
+We also describe in this work the \pkg{RNaviCell} package, a R package
+convenient to use with \pkg{ACSNMineR}. This package is dedicated to create
+web-based and interactive data visualization on ACSN maps. Users can use this
+tools to represent genes of interest that have been shown to be related to the
+maps by calculating enrichment with the  \pkg{ACSNMineR}.  Creating maps with
+the graphical user interface of the ACSN website can be a tedious task if the
+user has multiple samples or gene lists, and wants to compare their
+representations on ACSN maps. The \pkg{RNaviCell} package can be used to
+automate the process of creating the graphical representations automatically.
 
@@
 Of course, \pkg{ACSNMineR} is not the only R package for enrichment
-calculations. For instance, \BIOpkg{GOstats} \citep{falcon2007using} is probably
-one of the first packages that was created to calculate enrichment for Gene
-Ontology categories.  \pkg{GOstats} can also be used to calculate enrichment for
-other biological pathways categories, such as KEGG pathways (by using an
-instance of the class \code{KEGGHyperGParams}) or PFAM protein families (using
-\code{PFAMHyperGParams}). However, its usage might not be as straightforward as
-\pkg{ACSNMineR}, and it does not seem possible to test user-defined biological
-pathways. Furthermore, other authors have pointed out that the KEGG database
-used by this package has not been updated since 2012. \BIOpkg{clusterProfiler} is a
-recent R package released for enrichment analysis of Gene Ontology and KEGG with
-either hypergeometric test or Gene Set Enrichment Analysis (GSEA)
-\citep{yu2012clusterprofiler}. Via other packages, support for analysis of
-Disease Ontology and Reactome Pathways is possible. Interestingly, this package
-also offers the possibility to import user-defined gene set, through
-tab-delimited pairwise definition files.
+calculations. For instance, \BIOpkg{GOstats} \citep{falcon2007using} is
+probably one of the first packages that was created to calculate enrichment for
+Gene Ontology categories.  \pkg{GOstats} can also be used to calculate
+enrichment for other biological pathways categories, such as KEGG pathways (by
+using an instance of the class \code{KEGGHyperGParams}) or PFAM protein
+families (using \code{PFAMHyperGParams}). However, its usage might not be as
+straightforward as \pkg{ACSNMineR}, and it does not seem possible to test
+user-defined biological pathways. Furthermore, other authors have pointed out
+that the KEGG database used by this package has not been updated since 2012.
+\BIOpkg{clusterProfiler} is a recent R package released for enrichment analysis
+of Gene Ontology and KEGG with either hypergeometric test or Gene Set
+Enrichment Analysis (GSEA) \citep{yu2012clusterprofiler}. Via other packages,
+support for analysis of Disease Ontology and Reactome Pathways is possible.
+Interestingly, this package also offers the possibility to import user-defined
+gene set, through tab-delimited pairwise definition files. Other notable
+packages for enrichment calculations are \pkg{ReactomePA} for Reactome molecular
+pathways \citep{ReactomePA2016}, \pkg{miRNApath} for microRNA pathways
+\citep{miRNApath2008} and \pkg{gage} \citep{gage2009}. We believe that the main
+advantages of ACSNMineR compared to other packages are a direct access to the
+full set of ACSN maps (updated on a regular basis) and an easy way to test
+MSigDB gene sets or any user-defined gene set formatted appropriately.
 
@@
 
+\pkg{RNaviCell} relies on standard HTTP calls to provide informations and
+calculations, and we have developped a number of bindings for popular
+programming languages such as R, Java and Python \citep{bonnet2015navicell}.
+This open architecture is designed to facilitate the development of utilities
+by other programmers and to facilitate the integration of ACSN maps in existing
+frameworks. The development of such services, sometimes called ``microservices"
+\citep{fowler2014} is in expansion. Furthermore, this kind of open architecture
+could clear the way for a more unified and general access to reaction networks
+database, including for example WikiPathways \citep{kelder2012wikipathways},
+Reactome \citep{croft2014reactome} and other databases. The PSICQUIC project is
+a successfull example of such an architecture \citep{aranda2011psicquic}. It is
+an effort of the HUPO proteomics standard initiative to standardize the access
+to molecular interaction databases programmatically, based on the specification
+of web services (using REST and SOAP calls) and a common query language (MIQL).
+
 \section[Acknowledgments]{Acknowledgements}
@@
 
-\bibliography{ACSNMineR}
+\bibliography{Deveau}
 
@@
 }
-\email{andrei.Zinovyev@curie.fr}
+\email{andrei.zinovyev@curie.fr}
 
@@
   Eric Bonnet\\
-  Computational Systems Biology of Cancer, Institut Curie\\
-  26, rue d'Ulm 75248 Paris\\
+  Centre National de Génotypage, Institut de Génomique, CEA\\
+  2, rue Gaston Crémieux, 91057 Evry\\
   France\\
 }
-\email{eric.bonnet@curie.fr}
+\email{eric.bonnet@cng.fr}

### 2016-08-25 R/ACSN_Enrichment.R
Updated documentation with reviewer suggestions

@@ -16,11 +16,20 @@
 #' @param min_module_size will remove from the analysis all modules which are (strictly) smaller than threshold
 #' @param universe Universe on which the statistical analysis should be performed. 
 #' Can be either "HUGO","ACSN","map_defined", or a character vector of genes.
-#' @param threshold : maximal p-value (corrected if correction is enabled) that will be displayed
+#' @param threshold maximal p-value (corrected if correction is enabled) that will be displayed
 #' @param alternative One of "greater", "less", "both" or "two.sided"
 #' Greater will check for enrichment, less will check for depletion, and both will look for 
 #' both and will keep track of the side,
-#' while two-sided (only for fisher test) checks if there is a difference .
+#' while two-sided (only for fisher test) checks if there is a difference.
+#' @return Output is a dataframe with the following columns:\describe{
+#'  \item{module}{The name of the map or the module preceded by the map}
+#'  \item{module_size}{The number of genes in the module after taking into account universe reduction}
+#'  \item{nb_genes_in_module}{The number of genes from input list in the module}
+#'  \item{genes_in_module}{Names of the genes from input list in the module, space separated}
+#'  \item{universe_size}{size of the input universe}
+#'  \item{nb_genes_in_universe}{number of genes from the input list that are found in the universe}
+#'  \item{test}{the kind of test that was looked for. "greater" when enrichment is tested, "less" when depletion is tested, or "two.sided"}
+#' }
 #' @examples enrichment(genes_test,min_module_size = 10, 
 #'    threshold = 0.05,
 #'    maps = list(cellcycle = ACSNMineR::ACSN_maps$CellCycle),
@@ -543,10 +552,19 @@ enrichment<-function(Genes=NULL,
 #' has p-value lower than threshold, otherwise the threshold is applied for each sample independently.
 #' @param alternative One of "greater", "less", "both", or "two.sided" (only for fisher test).
 #' Greater will check for enrichment, less will check for depletion, and both will look for both.
-#' @examples multisample_enrichment(Genes_by_sample = list(set1 = genes_test[-1],set2=genes_test[-2]),
+#' @return Output is a list of dataframes with names the names given in `Genes_by_sample` with the following columns:\describe{
+#'  \item{module}{The name of the map or the module preceded by the map}
[…]

### 2016-08-25 jss/bioarxiv.tex

--- 2016-02-15 jss/article.tex
@@
-\documentclass[article]{jss}
+\documentclass{article}
+\usepackage{hyperref}
+\usepackage{graphicx}
+\usepackage[utf8]{inputenc}  
+\usepackage[T1]{fontenc}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% declarations for jss.cls %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\author{Paul Deveau, Emmanuel Barillot, Valentina Boeva, \\Andrei Zinovyev and Eric Bonnet} %% comma-separated
+\title{Calculating Biological Module Enrichment or Depletion and Visualizing Data on Large-scale Molecular Maps with ACSNMineR and RNaviCell R packages} %% without formatting
 
-%% almost as usual
-\author{Paul Deveau\\Institut Curie\\PSL University\\Univ. Paris-Sud\\INSERM U830\\INSERM U900\\Mines-ParisTech \And 
-        Emmanuel Barillot\\Institut Curie\\PSL University\\INSERM U900\\Mines-ParisTech \And 
-        Valentina Boeva\\Institut Curie\\PSL University\\INSERM U900\\Mines-ParisTech\\Institut Cochin\\INSERM U1016\\CNRS UMR 8104\\Paris Descartes University UMR-S1016 \AND 
-        Andrei Zinovyev\\Institut Curie\\PSL University\\INSERM U900\\Mines-ParisTech \And
-        Eric Bonnet\\Institut Curie\\PSL University\\INSERM U900\\Mines-ParisTech}
-\title{Calculating Biological Module Enrichment or Depletion and Visualizing Data on Large-scale Molecular Maps with the R packages \pkg{ACSNMineR} and \pkg{RNaviCell}}
+\begin{document}
 
-%% for pretty printing and a nice hypersummary also set:
-\Plainauthor{Paul Deveau, Emmanuel Barillot, Valentina Boeva, Andrei Zinovyev, Eric Bonnet} %% comma-separated
-\Plaintitle{Calculating Biological Module Enrichment or Depletion and Visualizing Data on Large-scale Molecular Maps with the R packages ACSNMineR and RNaviCell} %% without formatting
-\Shorttitle{Module Enrichment and Data Visualization with ACSNMineR and RNaviCell} %% a short title (if necessary)
+\maketitle
 
-%% an abstract and keywords
-\Abstract{
+\begin{abstract}
 Biological pathways or modules represent sets of interactions or functional
@@
 knowledge on pathways is organized in public databases such as the KEGG,
-Reactome, MINT databases, but also in more specialized repositories such as the
-recently released Atlas of Cancer Signaling Network (ACSN). All these open and
-popular biological databases facilitate a large spectrum of analyses, thereby
-improving our understanding of cellular systems. Here we describe the
-\pkg{ACSNMineR}, a novel R package for calculation of enrichment or depletion
-of list of genes of interest in biological pathways. The package has built-in
-support for ACSN molecular pathways, but can use any molecular pathway database
-encoded as a GMT file, such as sets of gene signatures available in the
-Molecular Signatures Database (MSigDB). We also present the R package
-\pkg{RNaviCell}, that can be used in conjunction with \pkg{ACSNMineR} to
-visualize different data types on web-based, interactive ACSN maps. We
-illustrate the functionalities of the two packages with real-world biological
-data taken from large-scale cancer datasets.
-
-}
-\Keywords{biological module, biological pathway, enrichment, depletion, cancer, maps, molecular networks, molecular pathways, \proglang{R}}
-\Plainkeywords{biological module, biological pathway, enrichment, depletion, cancer, maps, molecular networks, molecular pathways, R} %% without formatting
-%% at least one keyword must be supplied
-
-%% publication information
-%% NOTE: Typically, this can be left commented and will be filled out by the technical editor
-%% \Volume{50}
-%% \Issue{9}
-%% \Month{June}
-%% \Year{2012}
-%% \Submitdate{2012-06-04}
-%% \Acceptdate{2012-06-04}
-
-%% The address of (at least) one author should be given
-%% in the following format:
-\Address{
-  Eric Bonnet\\
-  Computational Systems Biology of Cancer\\
-  Institut Curie\\
-  26, rue d'Ulm\\
-  75248 Paris, France\\
-  E-mail: \email{eric.bonnet@curie.fr}\\
-}
-%% It is also possible to add a telephone and fax number
-%% before the e-mail in the following format:
-%% Telephone: +43/512/507-7103
-%% Fax: +43/512/507-2851
-
-%% for those who use Sweave please include the following line (with % symbols):
-%% need no \usepackage{Sweave.sty}
-
-%% end of declarations %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{document}
-
-%% include your article here, just as usual
-%% Note that you should use the \pkg{}, \proglang{} and \code{} commands.
+Reactome, or in more specialized repositories, such as the Atlas of Cancer
+Signaling Network (ACSN). All these open biological databases facilitate
+analyses, improving our understanding of cellular systems. We hereby describe
+the R package \verb|ACSNMineR| for calculation of enrichment or depletion of
+lists of genes of interest in biological pathways. \verb|ACSNMineR| integrates
+ACSN molecular pathways, but can use any molecular pathway encoded as a GMT
+file, for instance sets of genes available in the Molecular Signatures Database
+(MSigDB). We also present the R package \verb|RNaviCell|, that can be used in
+conjunction with \verb|ACSNMineR| to visualize different data types on
+web-based, interactive ACSN maps. We illustrate the functionalities of the two
+packages with biological data taken from large-scale cancer datasets.
+\end{abstract}
 
@@
 relationships, occurring at the molecular level in living cells
-\citep{adriaens2008public, barillot2012computational}.  A large body of
+\cite{adriaens2008public, barillot2012computational}.  A large body of
 knowledge on cellular biochemistry is organized in publicly available
-repositories such as the KEGG database \citep{kanehisa2011kegg}, Reactome
-\citep{croft2014reactome} and MINT \citep{zanzoni2002mint}. All these
+repositories such as the KEGG database \cite{kanehisa2011kegg}, Reactome
+\cite{croft2014reactome} and MINT \cite{zanzoni2002mint}. All these
 biological databases facilitate a large spectrum of analyses, improving our
@@
 that occur in a healthy mammalian cell but that are frequently deregulated
-during cancerogenesis \citep{kuperstein2015atlas}.  The ACSN atlas aims to
+during cancerogenesis \cite{kuperstein2015atlas}.  The ACSN atlas aims to
 be a comprehensive description of cancer-related mechanisms retrieved from the
@@
 technology, a software framework dedicated to web-based visualization and
-navigation for biological pathway maps \citep{kuperstein2013navicell}. This
+navigation for biological pathway maps \cite{kuperstein2013navicell}. This
 environment is providing an easy navigation of maps through the use of the
@@
 and a comprehensive module for visualization and analysis of high-throughput
-data \citep{bonnet2015navicell}.
+data \cite{bonnet2015navicell}.
 
-In this article, we describe two software packages related to ACSN analysis and
-data visualization for the popular R statistical enrironment \citep{mainRref,
-vance2009data}. The package \pkg{ACSNMineR} is designed for the calculation of
+In this article, we describe two R packages related to ACSN analysis and
+data visualization. The package \verb|ACSNMineR| is designed for the calculation of
 gene enrichment and depletion in ACSN maps (or any user-defined gene set via
-the import function), while \pkg{RNaviCell} is dedicated to data visualization
+the import function), while \verb|RNaviCell| is dedicated to data visualization
 on ACSN maps. Both packages are available on the Comprehensive R Archive
@@
 involving 2,371 proteins (as of today), and is continuously updated and expanded.  We have
-included the three hierarchical levels in the \pkg{ACSNMineR} package, in order
+included the three hierarchical levels in the \verb|ACSNMineR| package, in order
 to be able to calculate enrichments at all three levels. The calculations are
@@
  \centering
-  \caption{ACSN maps included in the \pkg{ACSNMineR} package. Map: map name,
+  \caption{ACSN maps included in the ACSNMineR package. Map: map name,
   Total: total number of gene symbols (HUGO) used to construct the map, Nb
@@
   gene sybols per module. N.B.: one gene symbol may be present in several
-  modules of the map.} \label{tab:table1}
-  \begin{tabular}{l|c|c|c|c|c}
+  modules of the map.} 
+  \label{tab:table1}
+  \begin{tabular}{lccccc}
     \hline
@@
 The statistical significance of the counts in the modules is assessed by using
-either the Fisher exact test \citep{fisher1922interpretation,
+either the Fisher exact test \cite{fisher1922interpretation,
 fisher1934statistical} or the hypergeometric test, which are equivalent for this
-purpose \citep{rivals2007enrichment}.
+purpose \cite{rivals2007enrichment}.
 
-The current ACSN maps are included in the \pkg{ACSNMineR} package, as a list of character matrices.
+The current ACSN maps are included in the \verb|ACSNMineR| package, as a list of character matrices.
 
@@
 
-The main function of the \pkg{ACSNMineR} package is the \code{enrichment}
+The main function of the \verb|ACSNMineR| package is the \verb|enrichment|
 function, which is calculating over-representation or depletion of genes in the
 ACSN maps and modules. We have included a small list of 12 Cell Cycle related
-genes in the package, named \code{genes_test} that can be used to test the main
+genes in the package, named \verb|genes_test| that can be used to test the main
 enrichment function and to get familiar with its different options.
@@
 
-%    enrichment(Genes = NULL, maps = ACSNMineR::ACSN_maps,
-%       correction_multitest = "BH", statistical_test = "fisher",
-%       min_module_size = 5, universe = "map_defined", threshold = 0.05,
-%       alternative = "greater")
-
-The \code{enrichment} function can take up to eight arguments: the gene list (as
+The \verb|enrichment| function can take up to eight arguments: the gene list (as
 a character vector), the list of maps that will be used to calculate enrichment
@@
 
-Only the gene list is mandatory to call the \code{enrichment} function, all the
-other arguments have default values.  The \code{maps} argument can either be a
-dataframe imported from a gmt file with the \code{format_from_gmt} function or
+Only the gene list is mandatory to call the \verb|enrichment| function, all the
+other arguments have default values.  The \verb|maps| argument can either be a
+dataframe imported from a gmt file with the \verb|format_from_gmt| function or
 a list of dataframes generated by the same procedure. By default, the function
@@
 changed to any of the usual correction methods (Bonferroni, Holm, Hochberg,
-Holm, or Benjamini \& Yekutieli \citep{Benjamini2003FDR}), or even disabled .
+Holm, or Benjamini \& Yekutieli \cite{Benjamini2003FDR}), or even disabled .
 The minimal module size represents the smallest size value of a module that
@@
 the maps will be discarded. To keep all genes, the user can change the universe
-to \code{HUGO}, and in that case, the complete list of HUGO gene symbols will
+to \verb|HUGO|, and in that case, the complete list of HUGO gene symbols will
 be used as the reference ($>$ 39,000 genes). The threshold corresponds to the
@@
 
-%multisample_enrichment<-function(Genes_by_sample=NULL,
-%                                 maps = ACSNMineR::ACSN_maps, 
-%                                 correction_multitest = "BH",
-%                                 statistical_test = "fisher",
-%                                 min_module_size = 5,
-%                                 universe = "map_defined",
-%                                 threshold = 0.05,
-%                                 cohort_threshold = TRUE,
-%                                 alternative = "greater"){
-
 It may be of interest to compare enrichment of pathways in different cohorts or
@@
 differences between two cancer types or two cell lines.  To facilitate such
-comparisons, \pkg{ACSNMineR} provides a \code{multisample_enrichment} function.
-It relies on the \code{enrichment} function but takes a list of character
+comparisons, \verb|ACSNMineR| provides a \verb|multisample_enrichment| function.
+It relies on the \verb|enrichment| function but takes a list of character
 vector genes. The name of each element of the list will be assumed to be the
 name of the sample for further analysis.  Most of the arguments given to
-\code{multisample_enrichment} are the same as the ones passed to
-\code{enrichment}. However, the \code{cohort_threshold} is designed to filter
+\verb|multisample_enrichment| are the same as the ones passed to
+\verb|enrichment|. However, the \verb|cohort_threshold| is designed to filter
 out modules which would not pass the significance threshold in all samples.
 
-%represent_enrichment<-function(enrichment, plot = "heatmap" , scale = "log", 
-%                               low = "steelblue" , high ="white",
-%                              nrow = 1,sample_name = "Sample",
-%                               na.value = "grey"){
-
-Finally, to facilitate visualization of results, \pkg{ACSNMineR} integrates a
-representation function based on \pkg{ggplot2} syntax \citep{ggplot2}. It
-allows representation of results from \code{enrichment} or
-\code{multisample_enrichment} with a limited number of parameters. Two types of
+Finally, to facilitate visualization of results, \verb|ACSNMineR| integrates a
+representation function based on \verb|ggplot2| syntax \cite{ggplot2}. It
+allows representation of results from \verb|enrichment| or
+\verb|multisample_enrichment| with a limited number of parameters. Two types of
 display are available: heat-map tiles or bars. For multiple samples using a
@@
 
-We previously computed the p-value of the \code{genes_test} list with default
+We previously computed the p-value of the \verb|genes_test| list with default
 parameters. The number of modules which have a p-value below 0.05 was 36, that
@@
 
-  \begin{tabular}{l|c|c|c|c}
+  \begin{tabular}{l c c c c}
 	\hline
@@
 	\hline
-CellCycle & 242 & ATM ATR CHEK2   & 5.36e-07 & greater \\
-          &   &  CREBBP KAT2B TFDP1   &          &         \\
-          &   &  E2F1 EP300 HDAC1    &          &         \\
-          &	  & GTF2H1 GTF2H2 GTF2H2B &	& \\
-CellCycle: APOPTOSIS ENTRY & 10 & ATM ATR CHEK2 E2F1 & 3.49e-07 & greater \\
+CellCycle & 242 & ATM ATR & 5.36e-07 & greater \\
+          &   &  CHEK2 &   &     \\
+          &   &  CREBBP KAT2B &   &     \\
+          &   &  TFDP1 E2F1  &    &     \\
+          &	  &  EP300 HDAC1 &	& \\
+          &	  &  GTF2H1 GTF2H2 &	& \\
+          &	  &  GTF2H2B &	& \\
+CellCycle: APOPTOSIS ENTRY & 10 & ATM ATR & 3.49e-07 &greater  \\
+ & & CHEK2 E2F1 &  & \\
 CellCycle: CYCLINB			& 7  & ATM 				  & 3.98e-02 & greater \\
@@
 
-We can now plot the first ten rows of the results obtained for corrected and
+We can now plot the first six rows of the results obtained for corrected and
 uncorrected fisher test with heatmap format (Figure~\ref{fig:heatm}) or barplot
@@
 
-\begin{figure}
+\begin{figure}[h!]
 	\centering
-    \caption{Representation of the enriched modules (first ten rows), with
+    \caption{Representation of the enriched modules (first six rows), with
     either Bonferroni correction or no correction. Grey tiles means that the
@@
     represented in blue.}
-    \includegraphics{figures/comparison_corrected_unc.pdf}
+    \includegraphics[width=\textwidth]{figures/comparison_corrected_unc.pdf}
     \label{fig:heatm}
@@
 provided in order to facilitate the development of data visualization workflows and
-third-party applications \citep{bonnet2015navicell}. RNaviCell is the R binding
+third-party applications \cite{bonnet2015navicell}. RNaviCell is the R binding
 to the NaviCell Web Service. It is implemented as a standard R package, using
-the R object-oriented framework known as Reference Classes \citep{hwR5}. Most
+the R object-oriented framework known as Reference Classes \cite{hwR5}. Most
 of the work done by the user using graphical point-and-click operations on the
@@
 http calls to the server with appropriate parameters and data. Calls to the
-NaviCell server are performed using the library RCurl \citep{rcurl2015}, while
+NaviCell server are performed using the library RCurl \cite{rcurl2015}, while
 data encoding/decoding in JSON format is performed with the RJSONIO library
-\citep{rjsonio2014}.
+\cite{rjsonio2014}.
 
@@
 treatment (agressive), and is taken from the Cell Line Encyclopedia project
-\citep{barretina2012cancer}. We visualize the data values on the Cell Cycle map
+\cite{barretina2012cancer}. We visualize the data values on the Cell Cycle map
 (the default map), using heat maps. With this visualization mode, gene
@@
 \centering
-  \includegraphics[width=0.8\textwidth]{figures/heatmap.pdf}
+  \includegraphics[width=\textwidth]{figures/heatmap.pdf}
   \label{fig:du145}
@@
 surgery using discovery approach to reveal additional prognostic motifs
-\citep{schmidt2008humoral}. Estrogen receptor (ER) expression and proliferative
+\cite{schmidt2008humoral}. Estrogen receptor (ER) expression and proliferative
 activity of breast carcinomas are well-known and described prognostic markers.
@@
 
-\begin{tabular}{l|c|c|c|c}
+\begin{tabular}{l c c c c}
 \hline
@@
 repository of well-annotated  gene sets representing the universe of biological
-processes \citep{liberzon2011molecular}. We downloaded the canonical pathways
+processes \cite{liberzon2011molecular}. We downloaded the canonical pathways
 set, counting more than 1,300 gene sets representing canonical pathways compiled
 by domain experts. The dataset is encoded with the 'gmt' format, and can be
-imported within ACSNMineR with the \code{format\_from\_gmt} function. We
+imported within ACSNMineR with the \verb|format_from_gmt| function. We
 calculate the enrichment for the breast cancer differentially expressed gene
-list, simply specifying the MSigDB data we just imported as the \code{maps}
+list, simply specifying the MSigDB data we just imported as the \verb|maps|
 option. Table \ref{tab:table_msigdb} is displaying the pathways having a
@@
 
-\begin{tabular}{l|c|c|c}
+\begin{tabular}{l c c c}
 \hline
@@
 At last, we visualize the mean expression values for ER negative samples for all
-genes differentially expressed on the ACSN master (global) map using \code{ACSNMineR}
+genes differentially expressed on the ACSN master (global) map using \verb|ACSNMineR|
 commands to create heatmaps.
@@
   \centering
-  \includegraphics[width=0.8\textwidth]{figures/mainz_acsn.pdf}
+  \includegraphics[width=\textwidth]{figures/mainz_acsn.pdf}
   \label{fig:mainz}
@@
 genome-wide scale, including genomic, epigenomic, transcriptomic, and proteomic
-information \citep{hawkins2010next}. Integrating these heterogeneous datasets
+information \cite{hawkins2010next}. Integrating these heterogeneous datasets
 provides more biological insights than performing separate analyses. For
@@
 already led to the identification of novel cancer genes
-\citep{mclendon2008comprehensive}.
+\cite{mclendon2008comprehensive}.
 
@@
 manifestation, glioblastoma (GBM), median survival ranges from 9 to 12 months,
-despite maximum treatment efforts \citep{maher2001malignant}. In this study we
+despite maximum treatment efforts \cite{maher2001malignant}. In this study we
 have analyzed whole-genome mutation data generated by the TCGA project on
@@
 We loaded the data as a matrix in R and calculated the enrichment in ACSN maps
-with the ACSNMineR function \code{enrichment}. The results are displayed in
+with the ACSNMineR function \verb|enrichment|. The results are displayed in
 table \ref{tab:table_gbm}. There are 6 modules significantly enriched in the
@@
   \centering
-  \includegraphics[width=0.8\textwidth]{figures/gbm.pdf}
+  \includegraphics[width=\textwidth]{figures/gbm.pdf}
   \label{fig:gbm}
@@
 \section[Summary and perspectives]{Summary and perspectives}
-In this work, we presented the R package \pkg{ACSNMineR}, a novel package for
+In this work, we presented the R package \verb|ACSNMineR|, a novel package for
 the calculation of p-values for enrichment or depletion of genes in biological
@@
 
-We also describe in this work the \pkg{RNaviCell} package, a R package convenient to use with \pkg{ACSNMineR}. This package is dedicated to create web-based
+We also describe in this work the \verb|RNaviCell| package, a R package convenient to use with \verb|ACSNMineR|. This package is dedicated to create web-based
 and interactive data visualization on ACSN maps. Users can use this tools to
 represent genes of interest that have been shown to be related to the maps by
-calculating enrichment with the  \pkg{ACSNMineR}.  Creating maps with the
+calculating enrichment with the  \verb|ACSNMineR|.  Creating maps with the
 graphical user interface of the ACSN website can be a tedious task if the user
 has multiple samples or gene lists, and wants to compare their representations
-on ACSN maps. The \pkg{RNaviCell} package can be used to automate the process of
+on ACSN maps. The \verb|RNaviCell| package can be used to automate the process of
 creating the graphical representations automatically.
 
-We have shown how the packages \pkg{ACSNMineR} and \pkg{RNaviCell} can be
+We have shown how the packages \verb|ACSNMineR| and \verb|RNaviCell| can be
 combined to analyze expression data from breast cancer samples, and also to
@@
 
-Of course, \pkg{ACSNMineR} is not the only R package for enrichment
-calculations. For instance, \pkg{GOstats} \citep{falcon2007using} is probably
+Of course, \verb|ACSNMineR| is not the only R package for enrichment
+calculations. For instance, \verb|GOstats| \cite{falcon2007using} is probably
 one of the first packages that was created to calculate enrichment for Gene
-Ontology categories.  \pkg{GOstats} can also be used to calculate enrichment for
+Ontology categories.  \verb|GOstats| can also be used to calculate enrichment for
 other biological pathways categories, such as KEGG pathways (by using an
-instance of the class \code{KEGGHyperGParams}) or PFAM protein families (using
-\code{PFAMHyperGParams}). However, its usage might not be as straightforward as
-\pkg{ACSNMineR}, and it does not seem possible to test user-defined biological
+instance of the class \verb|KEGGHyperGParams|) or PFAM protein families (using
+\verb|PFAMHyperGParams|). However, its usage might not be as straightforward as
+\verb|ACSNMineR|, and it does not seem possible to test user-defined biological
 pathways. Furthermore, other authors have pointed out that the KEGG database
-used by this package has not been updated since 2012. \pkg{clusterProfiler} is a
+used by this package has not been updated since 2012. \verb|clusterProfiler| is a
 recent R package released for enrichment analysis of Gene Ontology and KEGG with
 either hypergeometric test or Gene Set Enrichment Analysis (GSEA)
-\citep{yu2012clusterprofiler}. Via other packages, support for analysis of
+\cite{yu2012clusterprofiler}. Via other packages, support for analysis of
 Disease Ontology and Reactome Pathways is possible. Interestingly, this package
@@
 
-In order to improve \pkg{ACSNMineR}, we may in the near future try to improve
+In order to improve \verb|ACSNMineR|, we may in the near future try to improve
 the speed of calculation, which might be a problem if a very large number of
 samples or experiments have to be analyzed rapidly. For instance, we could use
-the \code{foreach} and \code{\%dopar\%} operator to parallelize the most
+the \verb|foreach| and \verb|\%dopar\%| operator to parallelize the most
 computationally demanding operations. It could also be useful to implement more
@@
 authors would like to thank Pierre Gestraud for his comments on early versions
-of the \pkg{ACSNMineR} package and Eric Viara for guidance and assistance on
-the development of the \pkg{RNaviCell} package.
+of the \verb|ACSNMineR| package and Eric Viara for guidance and assistance on
+the development of the \verb|RNaviCell| package.
 
-%\bibliographystyle{jss}
-\bibliography{biblio}
+\section{Affiliations}
+
+Paul Deveau (paul.deveau@curie.fr)\\
+  Computational Systems Biology of Cancer, Institut Curie\\
+  26, rue d'Ulm 75248 Paris\\
+  France\\
+
+\noindent
+Emmanuel Barillot (emmanuel.barillot@curie.fr)\\
+  Computational Systems Biology of Cancer, Institut Curie\\
+  26, rue d'Ulm 75248 Paris\\
+  France\\
+
+\noindent
+Valentina Boeva (valentina.boeva@curie.fr)\\
+  Computational Systems Biology of Cancer, Institut Curie\\
+  26, rue d'Ulm 75248 Paris\\
+  France\\
+
+\noindent
+Andrei Zinovyev (andrei.zinovyev@curie.fr)\\
+  Computational Systems Biology of Cancer, Institut Curie\\
+  26, rue d'Ulm 75248 Paris\\
+  France\\
+
+\noindent
+Eric Bonnet (eric.bonnet@cng.fr)\\
+%  Computational Systems Biology of Cancer, Institut Curie\\
+%  26, rue d'Ulm 75248 Paris\\
+%  France\\
+Centre National de Génotypage\\ 
+Institut de Génomique\\
+CEA\\
+91057 Evry\\ 
+France\\
+
+\bibliographystyle{unsrt}
+\bibliography{Deveau}
 

### 2016-08-25 Images/Images_readme.R
Updated examples, readme and images

@@ -1,0 +1,10 @@
+# Generate images
+q<-represent_enrichment(enrichment = list(SampleA = enrichment_test, 
+                                          SampleB = enrichment_test[1:3,]), 
+                        plot = "heatmap", scale = "log")
+ggsave("Images/Heatmap.png")
+
+q<-represent_enrichment(enrichment = enrichment_test,scale = "reverselog",
+                                          sample_name = "test",plot = "bar")
+
+ggsave("Images/barplot.png")

### 2016-08-25 R/ACSN_Enrichment.R

@@ -665,7 +665,8 @@ p.val.calc<-function(x,y,z,a,stat_test,alt){
 #'                     sample_name = "test",plot = "bar")
 #'
 #'represent_enrichment(enrichment = list(SampleA = enrichment_test, 
-#'SampleB = enrichment_test[1:3]), plot = "heatmap", scale = "log")
+#'                                      SampleB = enrichment_test[1:3,]), 
+#'                      plot = "heatmap", scale = "log")
 #'@return Function returns a ggplot2 object if input is a dataframe or a gridExtra object if the output is a list. 
 #'@import ggplot2 
 #'@importFrom gridExtra grid.arrange

### 2016-08-25 Images/Images_readme.R
Added output table
[skip ci]

@@ -8,3 +8,9 @@ q<-represent_enrichment(enrichment = enrichment_test,scale = "reverselog",
                                           sample_name = "test",plot = "bar")
 
 ggsave("Images/barplot.png")
+
+# Generate table
+knitr::kable(enrichment(genes_test,
+           min_module_size = 10, 
+           threshold = 0.05,
+           maps = list(cellcycle = ACSNEnrichment::ACSN_maps$CellCycle)))

### 2016-08-26 jss/Deveau.tex
Added script to reproduce "mainz"
- Updated mainz_acsn.png
- Updated results from enrichment, added that we display the 10 most
significant out of 125
- Script run on 25-26/08/2016 on R 3.2.2, windows 10
R version 3.2.2 (2015-08-14)
Platform: x86_64-w64-mingw32/x64 (64-bit)
Running under: Windows 8 x64 (build 9200)

locale:
[1] LC_COLLATE=French_France.1252  LC_CTYPE=French_France.1252
LC_MONETARY=French_France.1252 LC_NUMERIC=C
LC_TIME=French_France.1252

attached base packages:
[1] parallel  stats     graphics  grDevices utils     datasets  methods
base

other attached packages:
[1] RNaviCel

@@
 dim(results)
-[1] 7 9
+[1] 8 9
 \end{example}
@@
 \midrule
-CellCycle & 242 & 20 & 1.70e-03 & 0.021 \\
-CellCycle:E2F1\_TARGETS & 130 & 14 & 6.26e-04 & 0.009 \\
-CellCycle:E2F2\_TARGETS & 35 & 8 & 4.70e-05 & 0.009 \\
-CellCycle:E2F3\_TARGETS & 51 & 11 & 3.82e-06 & 0.0002 \\
-CellCycle:E2F4\_TARGETS & 100 & 15 & 9.99e-06 & 0.0003 \\
-DNA\_repair:SPINDLE\_CHECKPOINT & 28 & 5 & 3.82e-03 & 0.04 \\
-EMT\_motility:ECM & 147 & 13 & 5.09e-03 & 0.045 \\
+Apoptosis:AKT\_MTOR & 79 & 47 & 0.00 & 0.01 \\ 
+CellCycle:E2F2\_TARGETS & 35 & 22 & 0.01 & 0.04 \\ 
+CellCycle:E2F3\_TARGETS & 51 & 31 & 0.00 & 0.02 \\ 
+CellCycle:E2F4\_TARGETS & 100 & 60 & 0.00 & 0.00 \\ 
+DNA\_repair & 346 & 172 & 0.00 & 0.01 \\ 
+DNA\_repair:CELL\_CYCLE & 82 & 49 & 0.00 & 0.01 \\ 
+DNA\_repair:G1\_CC\_PHASE & 25 & 18 & 0.00 & 0.02 \\ 
+DNA\_repair:S\_CC\_PHASE & 46 & 28 & 0.00 & 0.03 \\
 \bottomrule
@@
   \caption{MSigDB canonical pathway database enrichment for genes differentially expressed between ER
-positive and ER negative samples in breast cancer.  Module : name of the
+positive and ER negative samples in breast cancer. This table presents the 10 modules with lowest p-value out of 125 with corrected p-value lower than 0.05. Module : name of the
 module. Mod. size: size of the module. Nb genes: number of genes from input
@@
 \midrule
-KEGG\_CELL\_CYCLE & 128 & 76 & 1.98e-07\\
-REACTOME\_CELL\_CYCLE\_MITOTIC & 325 & 158 & 2.40e-07\\
-REACTOME\_DNA\_REPLICATION & 192 & 98 & 1.33e-05\\
-PID\_FOXM1\_PATHWAY & 40 & 29 & 5.05e-05\\
-REACTOME\_MITOTIC\_M\_M\_G1\_PHASES & 172 & 87 & 7.38e-05\\
-REACTOME\_CELL\_CYCLE & 421 & 181 & 2.97e-04\\
-PID\_AURORA\_B\_PATHWAY & 39 & 27 & 3.38e-04\\
-REACTOME\_MITOTIC\_G1\_G1\_S\_PHASES & 137 & 70 & 3.83e-04\\ 
-PID\_SYNDECAN\_1\_PATHWAY & 46 & 30 & 4.40e-04\\
-REACTOME\_S\_PHASE & 109 & 58 & 4.40e-04\\
+  KEGG\_CELL\_CYCLE & 128 & 76 & $3.9 \times 10^{-8}$ \\ 
+  REACTOME\_CELL\_CYCLE\_MITOTIC & 325 & 159 & $3.9 \times 10^{-8}$ \\ 
+  REACTOME\_DNA\_REPLICATION & 192 & 98 & $4.9 \times 10^{-6}$ \\ 
+  PID\_FOXM1PATHWAY & 40 & 29 & $3.1 \times 10^{-5}$ \\ 
+  REACTOME\_MITOTIC\_M\_M\_G1\_PHASES & 172 & 87 & $3.1 \times 10^{-5}$ \\ 
+  REACTOME\_CELL\_CYCLE & 421 & 182 & $5 \times 10^{-5}$ \\ 
+  REACTOME\_MITOTIC\_G1\_G1\_S\_PHASES & 137 & 71 & $9 \times 10^{-5}$ \\ 
+  PID\_AURORA\_B\_PATHWAY & 39 & 27 & 0.0002 \\ 
+  REACTOME\_S\_PHASE & 109 & 58 & 0.00024 \\ 
+  PID\_SYNDECAN\_1\_PATHWAY & 46 & 30 & 0.00026 \\ 
 \bottomrule
@@
   \centering
-  \includegraphics[width=0.8\textwidth]{figures/mainz_acsn.pdf}
+  \includegraphics[width=0.8\textwidth]{figures/mainz_acsn.png}
   \label{fig:mainz}

### 2016-08-26 jss/scripts/Breast_script.R

@@ -1,0 +1,59 @@
+#source("https://bioconductor.org/biocLite.R")
+#biocLite()
+library(breastCancerMAINZ)
+library(Biobase)
+library(limma)
+library(ACSNMineR)
+library(hgu133a.db)
+library(RNaviCell)
+# load data and extract expression and phenotype data
+data(mainz)
+eset <- exprs(mainz)
+pdat <- pData(mainz)
+# Create list of genes differentially expressed between ER positive and
+# ER negative samples using moderated t-test statistics
+design <- model.matrix(~factor(pdat$er == '1'))
+lmFit(eset, design) -> fit
+eBayes(fit) -> ebayes
+toptable(ebayes, coef=2,n=25000) -> tt
+which(tt$adj < 0.05) -> selection
+rownames(tt[selection,]) -> probe_list
+mget(probe_list, env=hgu133aSYMBOL) -> symbol_list
+symbol_list <- as.character(symbol_list)
+# calculate enrichment in ACSN maps
+enrichment(symbol_list) -> results
+dim(results)
+
+
+### Generate table
+if(!require("xtable")) install.packages("xtable")
+print(xtable::xtable(results[,c("module","module_size","nb_genes_in_module","p.value","p.value.corrected")],
+                     display = c("s","d","d","g","g")),
+      include.rownames = FALSE)
+
+mtsig <- format_from_gmt('jss/scripts/c2.cp.v5.0.symbols.gmt')
+mtsig_enrich<- enrichment(symbol_list, maps = mtsig)[,c("module","module_size","nb_genes_in_module","p.value.corrected")]
+print(xtable::xtable(mtsig_enrich), include.rownames=FALSE)
+
+### Show 10 with lowest p-value
+print(xtable::xtable(mtsig_enrich[order(mtsig_enrich$p.value.corrected,decreasing = FALSE)[1:10],],
+                     display = c("s","d","d","g","g")),
+      include.rownames=FALSE,math.style.exponents = TRUE)
+
+
+## Extract genes lowly expressed
+apply(eset[probe_list,pdat$er == 0],1,mean)->er_minus_mean
+names(er_minus_mean)<-symbol_list
+er_minus_mean<-as.matrix(er_minus_mean)
+colnames(er_minus_mean) <- c('exp')
+
+## NaviCell viz
+navicell<-NaviCell()
+navicell$proxy_url <- "https://acsn.curie.fr/cgi-bin/nv_proxy.php"
[…]

### 2016-08-26 jss/Deveau.tex
Updated legends and figures for figures 1 & 2

@@
 	\centering
-    \caption{Representation of the enriched modules (first six rows), with
+    \caption{Representation of the enriched modules (first six rows for each setting), with
     either Bonferroni correction or no correction. Grey tiles means that the
@@
 	\centering
-	\caption{Representation of the enriched modules (first six rows), with either Bonferroni correction (left) or no correction (right).
+	\caption{Representation of the enriched modules (first six rows for each setting), with either Bonferroni correction (left) or no correction (right).
 The modules are on the X axis and the p-values are on the Y axis.
@@
 \midrule
-Apoptosis:AKT\_MTOR & 79 & 47 & 0.00 & 0.01 \\ 
-CellCycle:E2F2\_TARGETS & 35 & 22 & 0.01 & 0.04 \\ 
-CellCycle:E2F3\_TARGETS & 51 & 31 & 0.00 & 0.02 \\ 
-CellCycle:E2F4\_TARGETS & 100 & 60 & 0.00 & 0.00 \\ 
-DNA\_repair & 346 & 172 & 0.00 & 0.01 \\ 
-DNA\_repair:CELL\_CYCLE & 82 & 49 & 0.00 & 0.01 \\ 
-DNA\_repair:G1\_CC\_PHASE & 25 & 18 & 0.00 & 0.02 \\ 
-DNA\_repair:S\_CC\_PHASE & 46 & 28 & 0.00 & 0.03 \\
+  Apoptosis:AKT\_MTOR & 79 & 47 & 0.00043 & 0.0068 \\ 
+  CellCycle:E2F2\_TARGETS & 35 & 22 & 0.0055 & 0.043 \\ 
+  CellCycle:E2F3\_TARGETS & 51 & 31 & 0.0023 & 0.025 \\ 
+  CellCycle:E2F4\_TARGETS & 100 & 60 & $5.8 \times 10^{-5}$ & 0.0037 \\ 
+  DNA\_repair & 346 & 172 & 0.00038 & 0.0068 \\ 
+  DNA\_repair:CELL\_CYCLE & 82 & 49 & 0.00029 & 0.0068 \\ 
+  DNA\_repair:G1\_CC\_PHASE & 25 & 18 & 0.0013 & 0.016 \\ 
+  DNA\_repair:S\_CC\_PHASE & 46 & 28 & 0.0036 & 0.033 \\ 
 \bottomrule

### 2016-08-26 jss/scripts/ACSNMineR.R

@@ -42,14 +42,13 @@ heatmap <- represent_enrichment(enrichment = list(Corrected = results[1:6,],
                                 plot = "heatmap", scale = "reverselog", 
                                 low = "steelblue" , high ="white", na.value = "grey")
 
-print(heatmap)
-
+ggsave("jss/figures/comparison_corrected_unc.pdf")
 ### Plot bar
-baplot <- represent_enrichment(enrichment = list(Corrected = results[1:6,], 
+barplot <- represent_enrichment(enrichment = list(Corrected = results[1:6,], 
                                                  Uncorrected = results_uncorrected[1:6,]),
-                               plot = "heatmap", scale = "reverselog", 
+                               plot = "bar", scale = "reverselog", 
                                nrow = 1)
-print(barplot)
+ggsave("jss/figures/comparison_corrected_unc_bars.pdf")
 
 
 ### End of ACSNMineR demo

### 2016-08-26 jss/Deveau.tex
Corrected master map
- Updated tables 2,3,4 and figures 1,2 for paper according to reviewer 2
comments
- Corrected table 1 with new "Create_master" function
- Split functions in three R files: data_handling, enrichment and plots

@@
     \midrule
-  ACSN global & 2243 & 63 & 2 &629& 83\\
+  ACSN global & 2239 & 67 & 2 &629& 79\\
   Survival  &1053&5  &208 &431 &328\\
   Apoptosis & 667&7 & 19& 382& 136\\
-  EMT \& Cell motility &634 &9  &18 &629 &136\\
+  EMT \& Cell motility &634 &9  &18 &629 &137\\
   DNA repair &345&21  &3 &171  &45\\
-  Cell cycle &250&25  &2 &130  &19\\
+  Cell cycle &250&25  &2 &130  &20\\
   \bottomrule
@@
 genes mentionned above and a default p-value cutoff of $0.05$, we have a set of
-36 maps or modules that are significantly enriched. The results are structured
+8 maps or modules that are significantly enriched. The results are structured
 as a data frame with nine columns displaying the module name, the module size,
@@
 We previously computed the p-value of the \code{genes\_test} list with default
-parameters. The number of modules which have a p-value below 0.05 was 36, that
-can be compared to the 49 obtained without correction with the simple command
+parameters. The number of modules which have a p-value below 0.05 was 8, that
+can be compared to the 16 obtained without correction with the simple command
 shown below (some of the results are displayed in table \ref{tab:table2}).
@@
   \centering
-  \caption{Overview of the results from enrichment analysis without correction.
+  \caption{First rows of the results from enrichment analysis without correction.
   Module : name of the module. Mod. size: size of the module. Genes in module:
@@
 	\midrule
-CellCycle & 242 & ATM ATR CHEK2   & 5.36e-07 & greater \\
-          &   &  CREBBP KAT2B TFDP1   &          &         \\
-          &   &  E2F1 EP300 HDAC1    &          &         \\
-          &	  & GTF2H1 GTF2H2 GTF2H2B &	& \\
-CellCycle: APOPTOSIS ENTRY & 10 & ATM ATR CHEK2 E2F1 & 3.49e-07 & greater \\
-CellCycle: CYCLINB			& 7  & ATM 				  & 3.98e-02 & greater \\
+CellCycle & 242 & ATM ATR CHEK2 & $5.4 \times 10^{-7}$  & greater \\
+		  &     & CREBBP TFDP1 E2F1 &  & \\
+		  &     & EP300 HDAC1 KAT2B & & \\
+		  &     & GTF2H1 GTF2H2 GTF2H2B &  &\\ 
+  CellCycle:APOPTOSIS\_ENTRY & 10 & ATM ATR CHEK2 E2F1 & $3.5 \times 10^{-7}$  & greater \\ 
+  CellCycle:CYCLINB & 7 & ATM & 0.04 & greater \\
 
@@
                              Uncorrected = results_uncorrected[1:6,]),
-                               plot = "heatmap", scale = "reverselog", 
+                               plot = "bar", scale = "reverselog", 
                                nrow = 1)
@@
 the calculation of p-values for enrichment or depletion of genes in biological
-pathways. The package includes the six large-scale molecular maps and 55
+pathways. The package includes the six large-scale molecular maps and 67
 functional modules of the Atlas of Cancer Signaling Network (ACSN) . Enrichment

### 2016-08-26 R/ACSN_Enrichment.R

@@ -648,218 +648,6 @@ p.val.calc<-function(x,y,z,a,stat_test,alt){
   }
 }
 
-#### Graphic representation of results ####
-#' Graphic representation of enrichment
-#'@param enrichment Data frame or list of dataframes with p-values or corrected p-values (whenever available) 
-#'and module names for representation.
-#'The name of the dataframe will be used as sample name.
-#'@param plot Any of "heatmap" or "bar"
-#'@param scale Any of "log", "identity" or "reverselog" (i.e. -log10(p-value))
-#'@param low Color to be used in heatmap mode corresponding to lowest value
-#'@param high Color to be used in heatmap mode corresponding to highest value
-#'@param nrow Number of rows of the grid for display in bar mode.
-#'@param sample_name  used only is enrichment is a dataframe
-#'@param na.value color for the missing values in the heatmap
-#'@examples 
-#'represent_enrichment(enrichment = enrichment_test,scale = "reverselog",
-#'                     sample_name = "test",plot = "bar")
-#'
-#'represent_enrichment(enrichment = list(SampleA = enrichment_test, 
-#'                                      SampleB = enrichment_test[1:3,]), 
-#'                      plot = "heatmap", scale = "log")
-#'@return Function returns a ggplot2 object if input is a dataframe or a gridExtra object if the output is a list. 
-#'@import ggplot2 
-#'@importFrom gridExtra grid.arrange
-#'@export
-represent_enrichment<-function(enrichment, plot = "heatmap" , scale = "log", 
-                               low = "steelblue" , high ="white",
-                               nrow = 1,sample_name = "Sample",
-                               na.value = "grey"){
-  
-  
-  if(is.data.frame(enrichment)){
-    if("p.value.corrected" %in% colnames(enrichment)){
-      enrichment$p.values<-enrichment$p.value.corrected
-    }
-    else if("p.value" %in% colnames(enrichment)){
-      enrichment$p.values<-enrichment$p.value
-    }
-    else{
-      warning("dataframe has no column 'p.value' or 'p.value.corrected'. Exiting!")
-      return(NA)
-    }
-    enrichment$sample_name<-sample_name 
-    enrichment$p.values<-cnum(enrichment$p.values)
-    if(plot == "heatmap"){
-      
-      q<-ggplot2::ggplot(enrichment,
-                         ggplot2::aes_string(x= "sample_name",
[…]

### 2016-08-26 R/data_handling.R

@@ -1,0 +1,138 @@
+#Data and data handling
+
+#' Atlas of Cancer Signalling Networks
+#' 
+#' A dataset containing the six maps of ACSN: apoptosis, cell cycle, DNA reparation, EMT motility, survival, and the master map
+#' @format A list of dataframes
+#' \describe{
+#'  \item{Apoptosis}{Map of apoptosis pathways}
+#'  \item{CellCycle}{Map of the cell cycle pathways}
+#'  \item{DNA_repair}{Map of DNA repair}
+#'  \item{EMT_motility}{Map of the Epithelial Mesenchymal Transition}
+#'  \item{Master}{Map grouping all modules from other maps, without a master module for each map}
+#'  \item{Survival}{Map of cellular survival pathways}
+#' }
+#' @source \url{https://acsn.curie.fr/downloads.html}
+"ACSN_maps"
+
+#' Set of genes to test map
+#' 
+#' Genes of high importance in oncogenesis
+#' @format A character vector
+"genes_test"
+
+#' Result from enrichment test of "genes_test" on the ACSN maps
+#' 
+#' Parameters: bonferroni correction, min module size = 5
+#' @format data.frame
+#' \describe{
+#'  \item{module}{Name of module}
+#'  \item{genes_in_module}{Genes from genes_test in module}
+#'  \item{p.value}{Uncorrected p-value}
+#'  \item{p.value.corrected}{p-value corrected for multiple testing by Bonferroni correction}
+#'  }
+"enrichment_test"
+
+
+#' Import data from gmt files
+#' Convert gmt file to dataframe that can be used for anaysis
+#'@param path Path to the gmt file to be imported
+#'@return Returns a dataframe with the module - first column -, module length - seconde column - and gene names
+#'@examples file<-system.file("extdata", "cellcycle_short.gmt", package = "ACSNMineR")
+#'format_from_gmt(file)
+#'
+#'@export
+
+format_from_gmt<-function(path = ""){
+  
+  Lines<-readLines(path,warn = FALSE)
+  
+  
+  ### Filter out non-genes
+  if(length(Lines)==1){ ### testing if gmt is single line
+    gmt<-unlist(strsplit(x = Lines,split = "\t"))
+    short_gmt<-gmt[-c(1,2)]
+    pos<-grepl(pattern = "\\*",x = short_gmt)
+    result<-c(gmt[c(1,2)],gmt[-c(1,2)][!pos])
+    result[2]<-length(result)-2
+    result<-as.data.frame(t(result))
+    
[…]

### 2016-08-26 R/plots.R

@@ -1,0 +1,231 @@
+#Plots
+
+#### Graphic representation of results ####
+#' Graphic representation of enrichment
+#'@param enrichment Data frame or list of dataframes with p-values or corrected p-values (whenever available) 
+#'and module names for representation.
+#'The name of the dataframe will be used as sample name.
+#'@param plot Any of "heatmap" or "bar"
+#'@param scale Any of "log", "identity" or "reverselog" (i.e. -log10(p-value))
+#'@param low Color to be used in heatmap mode corresponding to lowest value
+#'@param high Color to be used in heatmap mode corresponding to highest value
+#'@param nrow Number of rows of the grid for display in bar mode.
+#'@param sample_name  used only is enrichment is a dataframe
+#'@param na.value color for the missing values in the heatmap
+#'@examples 
+#'represent_enrichment(enrichment = enrichment_test,scale = "reverselog",
+#'                     sample_name = "test",plot = "bar")
+#'
+#'represent_enrichment(enrichment = list(SampleA = enrichment_test, 
+#'                                      SampleB = enrichment_test[1:3,]), 
+#'                      plot = "heatmap", scale = "log")
+#'@return Function returns a ggplot2 object if input is a dataframe or a gridExtra object if the output is a list. 
+#'@import ggplot2 
+#'@importFrom gridExtra grid.arrange
+#'@export
+represent_enrichment<-function(enrichment, plot = "heatmap" , scale = "log", 
+                               low = "steelblue" , high ="white",
+                               nrow = 1,sample_name = "Sample",
+                               na.value = "grey"){
+  
+  
+  if(is.data.frame(enrichment)){
+    if("p.value.corrected" %in% colnames(enrichment)){
+      enrichment$p.values<-enrichment$p.value.corrected
+    }
+    else if("p.value" %in% colnames(enrichment)){
+      enrichment$p.values<-enrichment$p.value
+    }
+    else{
+      warning("dataframe has no column 'p.value' or 'p.value.corrected'. Exiting!")
+      return(NA)
+    }
+    enrichment$sample_name<-sample_name 
+    enrichment$p.values<-cnum(enrichment$p.values)
+    if(plot == "heatmap"){
+      
+      q<-ggplot2::ggplot(enrichment,
+                         ggplot2::aes_string(x= "sample_name",
+                                             y = "module", 
+                                             fill = "p.values"))+ggplot2::xlab("")+ ggplot2::ylab("Modules") + ggplot2::geom_tile()
+      if(scale == "log"){
[…]

### 2016-08-26 jss/scripts/ACSNMineR.R

@@ -33,8 +33,11 @@ print(results[1,])
 ### compute uncorrected enrichment
 
 results_uncorrected<-enrichment(genes_test,correction_multitest = FALSE)
-print(head(results_uncorrected,3))
+print(nrow(results_uncorrected))
 
+print(xtable::xtable(results_uncorrected[1:3,c("module","module_size","genes_in_module","p.value","test")],
+                     display = c("s","d","s","g","g","s")),
+      include.rownames = FALSE,math.style.exponents = TRUE)
 
 ### Plot heatmap
 heatmap <- represent_enrichment(enrichment = list(Corrected = results[1:6,], 
@@ -44,11 +47,11 @@ heatmap <- represent_enrichment(enrichment = list(Corrected = results[1:6,],
 
 ggsave("jss/figures/comparison_corrected_unc.pdf")
 ### Plot bar
-barplot <- represent_enrichment(enrichment = list(Corrected = results[1:6,], 
+pdf("jss/figures/comparison_corrected_unc_bars.pdf")
+represent_enrichment(enrichment = list(Corrected = results[1:6,], 
                                                  Uncorrected = results_uncorrected[1:6,]),
                                plot = "bar", scale = "reverselog", 
                                nrow = 1)
-ggsave("jss/figures/comparison_corrected_unc_bars.pdf")
-
+dev.off()
 
 ### End of ACSNMineR demo

### 2016-08-26 jss/scripts/Breast_script.R

@@ -28,15 +28,16 @@ dim(results)
 ### Generate table
 if(!require("xtable")) install.packages("xtable")
 print(xtable::xtable(results[,c("module","module_size","nb_genes_in_module","p.value","p.value.corrected")],
-                     display = c("s","d","d","g","g")),
-      include.rownames = FALSE)
+                     display = c("s","d","d","g","g","g")),
+      include.rownames = FALSE,math.style.exponents = TRUE)
 
 mtsig <- format_from_gmt('jss/scripts/c2.cp.v5.0.symbols.gmt')
 mtsig_enrich<- enrichment(symbol_list, maps = mtsig)[,c("module","module_size","nb_genes_in_module","p.value.corrected")]
 print(xtable::xtable(mtsig_enrich), include.rownames=FALSE)
 
 ### Show 10 with lowest p-value
-print(xtable::xtable(mtsig_enrich[order(mtsig_enrich$p.value.corrected,decreasing = FALSE)[1:10],],
+print(xtable::xtable(mtsig_enrich[order(mtsig_enrich$p.value.corrected,decreasing = FALSE)[1:10],
+                                  c("module","module_size","nb_genes_in_module","p.value.corrected")],
                      display = c("s","d","d","g","g")),
       include.rownames=FALSE,math.style.exponents = TRUE)
 

### 2016-08-30 jss/Deveau.tex
[skip ci] text update, multiple paragraphs

@@
 \abstract{
-Biological pathways or modules represent sets of interactions or functional relationships occurring at the molecular level in living cells.
-A large body of knowledge on pathways is organized in public databases such as the KEGG, Reactome, or in more specialized repositories, the Atlas of Cancer Signaling Network (ACSN) being an example.
-All these open biological databases facilitate analyses, improving our understanding of cellular systems.
-We hereby describe \CRANpkg{ACSNMineR} for calculation of enrichment or depletion of lists of genes of interest in biological pathways.
-\pkg{ACSNMineR} integrates ACSN molecular pathways, but can use any molecular pathway encoded as a GMT file, for instance sets of genes available in the Molecular Signatures Database (MSigDB).
-We also present \CRANpkg{RNaviCell}, that can be used in conjunction with \pkg{ACSNMineR} to visualize different data types on web-based, interactive ACSN maps.
-We illustrate the functionalities of the two packages with biological data taken from large-scale cancer datasets.
+Biological pathways or modules represent sets of interactions or functional
+relationships occurring at the molecular level in living cells. A large body of
+knowledge on pathways is organized in public databases such as the KEGG,
+Reactome, or in more specialized repositories, the Atlas of Cancer Signaling
+Network (ACSN) being an example. All these open biological databases facilitate
+analyses, improving our understanding of cellular systems. We hereby describe
+\CRANpkg{ACSNMineR} for calculation of enrichment or depletion of lists of
+genes of interest in biological pathways. \pkg{ACSNMineR} integrates ACSN
+molecular pathways gene sets, but can use any gene set encoded as a GMT file,
+for instance sets of genes available in the Molecular Signatures Database
+(MSigDB). We also present \CRANpkg{RNaviCell}, that can be used in conjunction
+with \pkg{ACSNMineR} to visualize different data types on web-based,
+interactive ACSN maps. We illustrate the functionalities of the two packages
+with biological data taken from large-scale cancer datasets.
 
@@
 other arguments have default values.  The \code{maps} argument can either be a
-dataframe imported from a gmt file with the \code{format\_from\_gmt} function or
-a list of dataframes generated by the same procedure. By default, the function
-uses the ACSN maps previously described.  The correction for multiple testing
-is set by default to use the method of Benjamini \& Hochberg, but can be
-changed to any of the usual correction methods (Bonferroni, Holm, Hochberg,
-Holm, or Benjamini \& Yekutieli \citep{Benjamini2003FDR}), or even disabled .
-The minimal module size represents the smallest size value of a module that
-will be used to compute enrichment or depletion. This is meant to remove
-results of low significance for module of small size.  The universe in which
-the computation is made by default is defined by all the gene symbols contained
-in the maps. All the genes that were given as input and that are not present on
-the maps will be discarded. To keep all genes, the user can change the universe
-to \code{HUGO}, and in that case, the complete list of HUGO gene symbols will
-be used as the reference ($>$ 39,000 genes). The threshold corresponds to the
-maximal value of the corrected p-value (unless the user chose not to correct
-for multiple testing) that will be displayed in the result table.
+dataframe imported from a GMT file with the \code{format\_from\_gmt} function
+or a list of dataframes generated by the same procedure. The GMT format
+corresponds to the Broad Institute's Gene Matrix Transposed file format, a
+convenient and easy way to encode named sets of genes of interest in
+tab-delimited text files (it is not a graph or network format). By default, the
+function \code{enrichment} uses the ACSN maps previously described.  The
+correction for multiple testing is set by default to use the method of
+Benjamini \& Hochberg, but can be changed to any of the usual correction
+methods (Bonferroni, Holm, Hochberg, Holm, or Benjamini \& Yekutieli
+\citep{Benjamini2003FDR}), or even disabled .  The minimal module size
+represents the smallest size value of a module that will be used to compute
+enrichment or depletion. This is meant to remove results of low significance
+for module of small size.  The universe in which the computation is made by
+default is defined by all the gene symbols contained in the maps. All the genes
+that were given as input and that are not present on the maps will be
+discarded. To keep all genes, the user can change the universe to \code{HUGO},
+and in that case, the complete list of HUGO gene symbols will be used as the
+reference ($>$ 39,000 genes). The threshold corresponds to the maximal value of
+the corrected p-value (unless the user chose not to correct for multiple
+testing) that will be displayed in the result table.
 
@@
 (the default map), using heat maps. With this visualization mode, gene
-expression values are represented as a color gradient (green to red) in
-squares positioned next to the entities where the gene has been mapped (Figure~\ref{fig:du145}).
+expression values are represented as a color gradient (green to red) in squares
+positioned next to the entities where the gene has been mapped
+(Figure~\ref{fig:du145}). Note that the map is displayed in a browser and is
+\emph{interactive}, i.e users can zoom in to display more information and for
+example look in what reactions are involved the genes selected to be displayed,
+and lots of other informations (see \cite{bonnet2015navicell} and
+\cite{kuperstein2015atlas} for more details).
 
@@
 set, counting more than 1,300 gene sets representing canonical pathways compiled
-by domain experts. The dataset is encoded with the 'gmt' format, and can be
+by domain experts. The dataset is encoded with the GMT format, and can be
 imported within ACSNMineR with the \code{format\_from\_gmt} function. We
@@
 At last, we visualize the mean expression values for ER negative samples for all
-genes differentially expressed on the ACSN master (global) map using \code{ACSNMineR}
+genes differentially expressed on the ACSN master (global) map using \code{RNaviCell}
 commands to create heatmaps.
@@
 tools to represent genes of interest that have been shown to be related to the
-maps by calculating enrichment with the  \pkg{ACSNMineR}.  Creating maps with
-the graphical user interface of the ACSN website can be a tedious task if the
-user has multiple samples or gene lists, and wants to compare their
-representations on ACSN maps. The \pkg{RNaviCell} package can be used to
-automate the process of creating the graphical representations automatically.
+maps by calculating enrichment with \pkg{ACSNMineR}.  Creating maps with the
+graphical user interface of the ACSN website can be a tedious task if the user
+has multiple samples or gene lists, and wants to compare their representations
+on ACSN maps. The \pkg{RNaviCell} package can be used to automate the process
+of creating the graphical representations automatically.  The maps are
+displayed in a browser and are interactive, with the possibility for the user
+to zoom in and out, search for genes or molecular species, and see the details
+of the molecular reactions (what partners are involved, what is the state of a
+given species, etc.). For more details on how to use the interface and the
+different possibilities, see \cite{kuperstein2013navicell},
+\cite{bonnet2015navicell} and \cite{kuperstein2015atlas}.
 

### 2016-08-31 jss/Deveau.tex
[skip ci] text update

@@
 ACSN maps from this list of genes. As seen in Table \ref{tab:table_mainz},
-there is one map (Cell Cycle) and six modules (belonging to the Cell Cycle, DNA
-repair and EMT motility maps) enriched.
+there is one map (DNA repair) and seven modules (belonging to the Cell Cycle, DNA
+repair and Apoptosis maps) enriched.
 
@@
 positive and ER negative samples in breast cancer.  Module : name of the
-module. Mod. size: size of the module. Nb genes: number of genes from input
+map/module. Mod. size: size of the module. Nb genes: number of genes from input
 which are found in the module. pval: raw p-value. Cor. pval: corrected p-value.}
@@
 
-The Figure \ref{fig:mainz} is displaying the map for gene that having a
-corrected p-value $< 0.05$ and a log fold-change $>1$. we can see that the genes
-are concentrated in the regions of the map corresponding to the cell cycle, cell
-motility, apoptosis and survival.
+The Figure \ref{fig:mainz} is displaying the map for genes having a
+corrected p-value $< 0.05$.
 
@@
 different possibilities, see \cite{kuperstein2013navicell},
-\cite{bonnet2015navicell} and \cite{kuperstein2015atlas}.
-
-We have shown how the packages \pkg{ACSNMineR} and \pkg{RNaviCell} can be
-combined to analyze expression data from breast cancer samples, and also to
-analyze the frequency of mutated genes in glioblastoma cancer samples.
+\cite{bonnet2015navicell} and \cite{kuperstein2015atlas}. We have shown how the
+packages \pkg{ACSNMineR} and \pkg{RNaviCell} can be combined to analyze
+expression data from breast cancer samples, and also to analyze the frequency
+of mutated genes in glioblastoma cancer samples.
 

### 2016-08-31 R/ACSN_Enrichment.R
Added Remove_from_universe option
- Updated dpi for vignette

@@ -16,6 +16,8 @@
 #' @param min_module_size will remove from the analysis all modules which are (strictly) smaller than threshold
 #' @param universe Universe on which the statistical analysis should be performed. 
 #' Can be either "HUGO","ACSN","map_defined", or a character vector of genes.
+#' @param Remove_from_universe: Default is NULL. A list of genes that should not be considered for enrichment 
+#' (will be removed from input, maps, and universe). The size of universe and map will be updated after removal. 
 #' @param threshold maximal p-value (corrected if correction is enabled) that will be displayed
 #' @param alternative One of "greater", "less", "both" or "two.sided"
 #' Greater will check for enrichment, less will check for depletion, and both will look for 
@@ -47,6 +49,7 @@ enrichment<-function(Genes=NULL,
                      statistical_test = "fisher",
                      min_module_size = 5,
                      universe = "map_defined",
+                     Remove_from_universe = NULL,
                      threshold = 0.05,
                      alternative = "greater"){
   
@@ -93,8 +96,20 @@ enrichment<-function(Genes=NULL,
     universe_was_ACSN<-FALSE
   }
   
+
   ### Checking that gene list is unique
   Genes<-unique(Genes)
+  if(!is.null(Remove_from_universe)){
+    Remove_from_universe<-unique(Remove_from_universe)
+    test<-Genes %in% Remove_from_universe
+    if(sum(test)){
+      warning(paste("The following genes were found in input and asked to be removed from universe:\n",
+                    paste(Genes[test],collapse = " "))
+      )
+      Genes<-Genes[!test]
+    }
+    
+  }
   Genes_size<-length(Genes)    
   if(!(alternative %in% c("greater","less","both","two.sided"))){
     stop('enrichment variable should be one of:"greater","less","both","two.sided" ')
@@ -112,15 +127,37 @@ enrichment<-function(Genes=NULL,
   ### If universe is ACSN: extract genes from ACSN and define it as universe
   if(length(universe) == 1){
     if(universe == "ACSN"){
-      genesACSN<-unique(unlist(lapply(X = ACSNMineR::ACSN_maps,FUN = function(z){
-        return(as.character(unique(z[,-(1:2)])))
-      })))
-      universe<-genesACSN[genesACSN!=""]
-      size<-length(genesACSN)
+      if(is.null(Remove_from_universe)){
[…]

### 2016-08-31 R/ACSN_Enrichment.R
Fixed typo in documentation

@@ -16,7 +16,7 @@
 #' @param min_module_size will remove from the analysis all modules which are (strictly) smaller than threshold
 #' @param universe Universe on which the statistical analysis should be performed. 
 #' Can be either "HUGO","ACSN","map_defined", or a character vector of genes.
-#' @param Remove_from_universe: Default is NULL. A list of genes that should not be considered for enrichment 
+#' @param Remove_from_universe Default is NULL. A list of genes that should not be considered for enrichment 
 #' (will be removed from input, maps, and universe). The size of universe and map will be updated after removal. 
 #' @param threshold maximal p-value (corrected if correction is enabled) that will be displayed
 #' @param alternative One of "greater", "less", "both" or "two.sided"
@@ -588,7 +588,7 @@ enrichment<-function(Genes=NULL,
 #' @param min_module_size will remove from the analysis all modules which are (strictly) smaller than threshold
 #' @param universe Universe on which the statistical analysis should be performed. Can be either "HUGO","ACSN" 
 #' ,"map_defined", or a character vector of genes.
-#' @param Remove_from_universe: Default is NULL. A list of genes that should not be considered for enrichment 
+#' @param Remove_from_universe Default is NULL. A list of genes that should not be considered for enrichment 
 #' (will be removed from input, maps, and universe). The size of universe and map will be updated after removal. 
 #' @param threshold maximal p-value (corrected if correction is enabled) that will be displayed
 #' @param cohort_threshold if TRUE modules will be kept in all samples if at least one sample 

### 2016-09-01 jss/Deveau.tex
[skip ci] minor change in text

@@
 
-The \code{enrichment} function can take up to eight arguments: the gene list (as
+The \code{enrichment} function can take up to nine arguments: the gene list (as
 a character vector), the list of maps that will be used to calculate enrichment
 or depletion, the type of statistical test (either the Fisher exact test or the
-hypergeometric test), the module minimal size for which the calculations will be
-done, the universe, the p-value threshold and the alternative ("greater" for
-calculating over-representation, "less" for depletion and "both" for both
-tests).
+hypergeometric test), the module minimal size for which the calculations will
+be done, the universe, the p-value threshold, the alternative hypothesis
+("greater" for calculating over-representation, "less" for depletion and "both"
+for both tests) and a list of genes that should be removed from the universe
+(option "Remove\_from\_universe"). This option may be useful for instance if we
+know beforehand that a number of genes are not expressed in the samples
+considered.
 

### 2016-09-20 jss/Deveau.tex
very very minor changes.

@@
 \author{by Paul Deveau, Emmanuel Barillot, Valentina Boeva, Andrei Zinovyev and Eric Bonnet} %% comma-separated
-\title{Calculating Biological Module Enrichment or Depletion and Visualizing Data on Large-scale Molecular Maps with \pkg{ACSNMineR} and \pkg{RNaviCell} packages} %% without formatting
+\title{Calculating Biological Module Enrichment or Depletion and Visualizing Data on Large-scale Molecular Maps with \pkg{ACSNMineR} and \pkg{RNaviCell} Packages} %% without formatting
 %\Shorttitle{Module Enrichment and Data Visualization with ACSNMineR and RNaviCell} %% a short title (if necessary)
@@
 (Figure~\ref{fig:du145}). Note that the map is displayed in a browser and is
-\emph{interactive}, i.e users can zoom in to display more information and for
+\emph{interactive}, i.e.\ users can zoom in to display more information and for
 example look in what reactions are involved the genes selected to be displayed,
@@
 calculate gene mutation frequencies. We kept the mutations having a status
-likely to disturb the target protein's function (i.e, Frame\_Shift\_Del,
+likely to disturb the target protein's function (i.e.\, Frame\_Shift\_Del,
 Nonstop\_Mutation, In\_Frame\_Del, In\_Frame\_Ins, Missense\_Mutation,

### 2016-09-20 jss/Deveau.tex
Minor paper corrections
- space around "="
- Added U830 affiliation

@@
 rownames(tt[selection,]) -> probe_list
-mget(probe_list, env=hgu133aSYMBOL) -> symbol_list
+mget(probe_list, env = hgu133aSYMBOL) -> symbol_list
 symbol_list <- as.character(symbol_list)
@@
   Computational Systems Biology of Cancer, Institut Curie\\
+  Genetics and Biology of Cancers, Institut Curie\\
   26, rue d'Ulm 75248 Paris\\
